### 2015-09-10 paper.tex
skeleton

\documentclass{sig-alternate}
\usepackage{color}

\newcommand{\red}[1]{\textcolor{red}{#1}}

\title{Adaptation Writ Large:\\ An essay in memory of John H. Holland}

%\title{A Common Framework for Defense across Complex Systems}

%\title{Complex Defense: Towards a Theory of Security across Complex Systems}

% Listed authors alphabetically for now.

\numberofauthors{2}
\author 
{\alignauthor
 Stephanie Forrest\\
 \affaddr{University of New Mexico}\\
 \affaddr{Santa Fe Institute}\\
 \email{forrest@cs.unm.edu}
 \alignauthor
Melanie Mitchell \\
 \affaddr{Portland State University}\\
 \affaddr{Santa Fe Institute}\\
 \email{mm@cs.pdx.edu}
}

\begin{document}
\maketitle

% \begin{abstract}
% \input{abstract}
% \end{abstract}

\section{Introduction}

Prof. John H. Holland passed away recently in Ann Arbor, MI, where he had been on the Univ. of Michigan faculty for XXXX years.  John, as he was known universally to his colleagues and students, 
leaves behind a long legacy of intellectual achievements that is in some ways difficult to pin down.   He is best known for the mechanisms he proposed, especially genetic algorithms, but the motivations behind the mechanisms are less well known but probably more important in the long run.  These mechanisms did not always implement his theoretical framworks perfectly, and in many cases too much effort was spent trying to rescue the mechanisms and not enough in understanding the underlying rationale.  In this short essay, we aim to remind people of some of the key themes of his work, discuss the role they have played in computer science, and highlight especially the ideas that we think remain relevant to today's research agendas.  Many of the problems that John identified and spent his life studying, such as [FILL IN HERE], are still poorly understood.

At a time, when computer science was crytstallizing into an engineering discipline, John came out of the cybernetics era (dating back to von Neumann, Wiener, Ashby, Turing) with a view of computation as a broad interdisciplinary enterprise: "Computation as an idea is something that goes much deeper than operating systems, programming languages, and the like.  The deep ideas of computation are intimately related to the deep ideas of life and intelligence" [Holland quote: Mitchell book, p. 128].  Thus, he became an early proponent of interdisciplinary approaches to computer science and was an active evangelist of what is now called computational thinking, reaching out psychologists, economists, physicists, linguists, philosophers and pretty much anyone he came in contact with.  As a result, even though he received what was arguably the first Ph.D. in our discipline (1958, XXXX), his contributions are much better known outside of computer science than within.

Here we emphasize four recurrent themes of his work: Leveraging statistics in adaptation, learning and credit assignment, models, and the generic properties of complex adaptive systems.

\section{Leveraging statistics in adaptation}

\section{ Learning and Credit assignment}

\section{Modeling}

Modeling is central to Holland's theory of adaptive systems.  He posits that all adaptive systems create and use internal models to survive in their environments, through anticipation.  Models can be tacit and learned through evolutionary time, as in the case of bacteria swimming up a chemical gradient, or explicit and learned over a single lifespan as in the case of cognitive systems that incorporate experience into internal rule sets through learning.  In \underline{Induction}, Holland and his co-authors discuss how such internal models can be learned, without supervision, by combining environmental inputs with stored knowledge.  A key idea is that a model defines an equivalence relation over a set of environmental states, together with a set of transition rules, which are learned over time based on environmental inputs.  Models that form valid homomorphisms with the environment allow the system to make accurate predictions.  In his conception, the equivalence classes are initially very general, say 'moving object' and 'stationary object,' and over time through experience and learning they are specialized into more useful and precise subclasses, say 'insect' and 'nest', which form a default hierarchy and are associated with more specific transition rules.  Although the idea of default hierarchies is prevalent in many knowledge representation systems (LIST A FEW?), Holland made two key contributions, first by emphsizing homomorphisms as a formal way to evaluate model validity, and idea that dates back to Ross Ashby's \underline{An Introduction to Cybernetics} that Holland's student, Bernard Ziegler, developed into a formal theory of computer modeling and simulation.  Today, even though computational modeling is a major application area of computing,  essential to most natural sciences and engineering disciplines, these early homorphic theories of modeling stand as the most elegant approach we know of for characterizing when a model is consistent with the environment and how an intelligent agent, human or artificial, can update the model to better reflect reality.
Holland's second key contribution was describing a computational mechanism by which a cognitive system could iteratively build up a detailed and hierarchical model of its environment, as Melanie already discussed above.

Given that Holland believed that the ability to learn and manipulate internal models was essential for any adapting system, it is no suprise that he viewed modeling as essential for scientific inquiry.  
Today, we use computational models in several distinct ways---as a tool for analyzing data in statistical models, through simulation as a method for discovering new knowledge, and as a way to explain or understand how a system works.  This third use of models was dear to Holland's heart.  In his view, the key to science was understanding the mechanisms that cause a system to behave in a certain way, an aspiration that goes well beyond data fitting methods, which typically focus only on describing the aggregate behavior of a system.  For example, a statistical model that describes the boom and bust pattern of the stock market cannot address the underlying mechanisms that lead to these cycles, through the collective actions of many many individual buy/sell decisions.  Similarly, the genetic algorithms for which he is so famous provide a simple computational framework in which to explore the dynamics of Darwinian evolution and whether the basic mechanisms of variation, differential reproduction, and heredity are sufficient to account for the richness of our natural world.  This emphasis on exploratory models to build intuitions was an important theme of Holland's work, and he often quoted Eddington's remark on the occasion of the first experimental test of Einstein's theory of relativity: "The contemplation in natural science of a wider domain than the actual leads to a far better understanding of the actual."  Holland was interested in models that explored basic principles and mechanisms, even if they didn't make specific predcitions, models that could show generically how certain behaviors could be produced, for example how exponential growth might interact with finite resources.  Holland pioneered a style of modeling that has come to be known as 'individual based' or 'agent based,' in which every component of a system is represented explicitly and has state, e.g., every trader in a stock market system or every cell in an immune system model, and each agent had its own behaviors, which it could update (learn) over time.  These models are often defined over spatial structures, such as networks or simple grids, to capture the constraints of systems living under spatial constraints.  An agent-based model, then, encodes a theory about the mechanisms that are relevant for producting the behavior of interest.  Similar to expert systems, such models are especially useful for studying systems that don't have analytic mathematical descriptions and they can facilitate interdisicplinary collaborations because the underlying rules can be easily communicated.  It should be noted that Holland's view of modeling is by no means typical.  For example, in a textbook on computational modeling, the authors offer the following definition:  “Modeling is the application of methods to analyze complex, real-world problems in order to make predictions about what might happen with various actions.” [Shiflet and Shiflet, 2006].  This sort of perspective completely rules out the kind of modeling that Holland was most interested in, namely exploratory modeling.  Similarly, many dispute that models make any kind of scientific contribution:  “Models are metaphors that explain the world we don’t understand in terms of worlds we do.  They are merely analogies, provide partial insight, stand on someone else’s feet.  Theories stand on their own feet, and rely on no analogies.”  [Emanuel Derman, 2012].  [NOW WE NEED A STRONG FINISHING SENTENCE TO RESCUE JHH STYLE MODELING.]

\section{Complexity}

\section{Conclusion}

Introduced a few generations of students to computation in natural systems, an idea that today is better accepted.  His insights were deeper and more general than what often passes for work in biomimicry, e.g., for robots.

The ideas have had huge impact and should still be a beacon for research in intelligent and complex systems

John's personality and humanity is inextricably tangled up with his intellectual contributions.

\end{document}

### 2015-09-16 paper.tex
Trying again to change my email address
email change
Fixed a few errors in my section

@@
 
-\title{Adaptation Writ Large:\\ An essay in memory of John H. Holland}
-
-%\title{A Common Framework for Defense across Complex Systems}
-
-%\title{Complex Defense: Towards a Theory of Security across Complex Systems}
-
-% Listed authors alphabetically for now.
+\title{Adaptation Writ Large:\\ An Essay in Memory of John H. Holland}
 
@@
  \affaddr{Santa Fe Institute}\\
- \email{mm@cs.pdx.edu}
+ \email{mm@pdx.edu}
 }
@@
 
-Prof. John H. Holland passed away recently in Ann Arbor, MI, where he had been on the Univ. of Michigan faculty for XXXX years.  John, as he was known universally to his colleagues and students, 
-leaves behind a long legacy of intellectual achievements that is in some ways difficult to pin down.
-He is best known for the mechanisms he proposed, especially genetic algorithms, but the motivations behind the mechanisms are less well known but probably more important in the long run.
-These mechanisms did not always implement his theoretical framworks perfectly, and in many cases too much effort was spent trying to rescue the mechanisms and not enough in understanding the underlying rationale.
-In this short essay, we aim to remind people of some of the key themes of his work, discuss the role they have played in computer science, and highlight especially the ideas that we think remain relevant to today's research agendas.
-Many of the problems that John identified and spent his life studying, such as [FILL IN HERE], are still poorly understood.
+Professor John H. Holland passed away recently in Ann Arbor, MI, where
+he had been on the University of Michigan faculty for over 50 years.
+John, as he was known universally to his colleagues and students,
+leaves behind a long legacy of intellectual achievements that is in
+some ways difficult to pin down.  He is best known for the mechanisms
+he proposed, especially genetic algorithms, but the motivations behind
+the mechanisms are less well known but probably more important in the
+long run.  These mechanisms did not always implement his theoretical
+framworks perfectly, and in many cases too much effort was spent
+trying to rescue the mechanisms and not enough in understanding the
+underlying rationale.  In this short essay, we aim to remind people of
+some of the key themes of his work, discuss the role they have played
+in computer science, and highlight especially the ideas that we think
+remain relevant to today's research agendas.  Many of the problems
+that John identified and spent his life studying, such as [FILL IN
+  HERE], are still poorly understood.
 
-At a time, when computer science was crytstallizing into an engineering discipline, John came out of the cybernetics era (dating back to von Neumann, Wiener, Ashby, Turing) with a view of computation as a broad interdisciplinary enterprise: "Computation as an idea is something that goes much deeper than operating systems, programming languages, and the like.
-The deep ideas of computation are intimately related to the deep ideas of life and intelligence" [Holland quote: Mitchell book, p.
-128].
-Thus, he became an early proponent of interdisciplinary approaches to computer science and was an active evangelist of what is now called computational thinking, reaching out psychologists, economists, physicists, linguists, philosophers and pretty much anyone he came in contact with.
-As a result, even though he received what was arguably the first Ph.D.
-in our discipline (1958, XXXX), his contributions are much better known outside of computer science than within.
+At a time, when computer science was crytstallizing into an
+engineering discipline, John came out of the cybernetics era (dating
+back to von Neumann, Wiener, Ashby, Turing) with a view of computation
+as a broad interdisciplinary enterprise.  Thus, he became an early
+proponent of interdisciplinary approaches to computer science and was
+an active evangelist of what is now called computational thinking,
+reaching out psychologists, economists, physicists, linguists,
+philosophers and pretty much anyone he came in contact with.  As a
+result, even though he received what was arguably the first Ph.D. in
+our discipline, his contributions are much better known outside of
+computer science than within.
 
-Here we emphasize four recurrent themes of his work: Leveraging statistics in adaptation, learning and credit assignment, models, and the generic properties of complex adaptive systems.
+Here we emphasize four recurrent themes of his work: adaptation as
+exploitation and exploration, learning and credit assignment, models,
+and the generic properties of complex adaptive systems.
 
-\section{Leveraging statistics in adaptation}
+\section{Adaptation as Exploitation and Exploration}
 
-\section{ Learning and Credit assignment}
+Although John Holland is best known for inventing genetic algorithms, his
+lifelong motivation was much broader: to understand the
+phenomenon of adaptation in all its guises.  Holland's interest in
+developing a general theory of adaptation was spurred both by his
+early work on computer models of Hebbian learning
+\cite{RochesterHolland}.  This interest developed further with
+Holland's extensive reading in evolutionary biology, economics, game
+theory, and control theory.  Holland recognized that all these fields
+in essence concern agents that must use information as it is gained from
+uncertain, changing environments in order to continually improve their
+performance.  Moreover, Holland recognized that in systems with
+``adaptive'' agents of this kind, there is never a state of {\emph
+  equilibrium} or a final {\emph optimum} configuration; rather, due
+to the environment's ``perpetual novelty'' (as Holland termed it),
+adaptation continues forever in an open-ended way.
+
+Underlying Holland's theory of adaptation are three key ideas: (1) In
+a population undergoing adaptation, individuals can be decomposed into
+{\emph building blocks}---traits that are the ``atoms'' of an
+individual's fitness or performance.  (2) Successful adaptation
+requires the right tradeoff between {\emph exploitation}, in which
+tried-and-true building blocks propagate in a population, and {\emph
+  exploration}, in which new building blocks (or recombinations of
+existing building blocks) arise and are subject to the test of the
+environment. (3) Adaptive evolution (biological and otherwise) is a
+population-based stochastic process that results in a nearly optimal
+exploitation versus exploration tradeoff.
+
+Holland's early papers (e.g.,
+\cite{OutlineLogicalTheory,OptimalAllocation}) and his influential
+1975 book {\emph Adaptation in Natural and Artificial Systems}
+\cite{ANAS} developed a general, formal setting in which these ideas
+could be expressed mathematically.  Moreover, Holland stated and
+proved a set of theorems that captured the third idea: that, given certain
+assumptions, his model of adaptation, while explicitly modifying
+populations of individuals, implicitly searched the space of possible
+building blocks in a way that optimized the exploitation and
+exploration balance.  (As an earlier part of this work, Holland
+separately derived an equation for that optimal balance, via a
+so-called ``multi-armed bandit'' model.)
+
+It was this formal setting and resulting theorems that led to the
+invention of genetic algorithms, which featured stochastic population-based
+search and recombination (or {\emph crossover}) as the most important
+operation.  However, it is important to note that
+the framework developed in \cite{ANAS} was more general than the
+genetic algorithm and was created in order to develop an
+interdisciplinary theory of adaptation, one that would inform biology,
+say, as much as computer science \cite{ChristiansenFeldmanPaper}.  The
+later, successful application of genetic algorithms to real-world
+optimization and learning tasks was, for Holland, just icing on the
+cake.
+
+\section{Learning Via Credit Assignment and Rule Discovery}
 

### 2015-09-22 paper.tex
MM's edited version
Edited Introduction and Building Block section.  Added notes.

--- 2015-09-10 paper.tex
@@
 
-Prof. John H. Holland passed away recently in Ann Arbor, MI, where he had been on the Univ. of Michigan faculty for XXXX years.  John, as he was known universally to his colleagues and students, 
-leaves behind a long legacy of intellectual achievements that is in some ways difficult to pin down.
-He is best known for the mechanisms he proposed, especially genetic algorithms, but the motivations behind the mechanisms are less well known but probably more important in the long run.
-These mechanisms did not always implement his theoretical framworks perfectly, and in many cases too much effort was spent trying to rescue the mechanisms and not enough in understanding the underlying rationale.
-In this short essay, we aim to remind people of some of the key themes of his work, discuss the role they have played in computer science, and highlight especially the ideas that we think remain relevant to today's research agendas.
-Many of the problems that John identified and spent his life studying, such as [FILL IN HERE], are still poorly understood.
+Professor John H. Holland passed away recently in Ann Arbor, MI, where
+he had been on the University of Michigan faculty for over 50 years.
+John, as he was known universally to his colleagues and students,
+leaves behind a long legacy of intellectual achievements.
 
-At a time, when computer science was crytstallizing into an engineering discipline, John came out of the cybernetics era (dating back to von Neumann, Wiener, Ashby, Turing) with a view of computation as a broad interdisciplinary enterprise: "Computation as an idea is something that goes much deeper than operating systems, programming languages, and the like.
-The deep ideas of computation are intimately related to the deep ideas of life and intelligence" [Holland quote: Mitchell book, p.
-128].
-Thus, he became an early proponent of interdisciplinary approaches to computer science and was an active evangelist of what is now called computational thinking, reaching out psychologists, economists, physicists, linguists, philosophers and pretty much anyone he came in contact with.
-As a result, even though he received what was arguably the first Ph.D.
-in our discipline (1958, XXXX), his contributions are much better known outside of computer science than within.
+John was a direct intellectual descendant of the cybernetics era, and
+early on was strongly influenced by the work of von Neumann, Wiener,
+Ashby, and Turing, all of whom viewed computation as a broad,
+interdisciplinary enterprise.  Thus, John became an early proponent of
+interdisciplinary approaches to computer science and was an active
+evangelist of what is now called {\emph computational thinking}, reaching out
+enthusiastically to psychologists, economists, physicists, linguists,
+philosophers, and pretty much anyone he came in contact with.  As a
+result, even though he received what was arguably the world's first
+computer science Ph.D. in 1959, his contributions are sometimes better known
+outside of CS than within.
 
-Here we emphasize four recurrent themes of his work: Leveraging statistics in adaptation, learning and credit assignment, models, and the generic properties of complex adaptive systems.
+John Holland is best known for his invention of {\emph genetic
+  algorithms} (GAs), a family of search and learning methods inspired
+by biological evolution.  Since their invention in the 1960s, GAs,
+along with related evolutionary computation methods, have become a
+thriving subfield of computer science, with widespread scientific and
+commercial applications.  While the mechanisms of GAs are well-known
+to much of the CS community, fewer are aware that GAs were only one
+offshoot of Holland's much broader motivation---to develop a general
+theory of adaptation in complex systems.  This motivation was the
+driving force in all of Holland's research.
 
-\section{Leveraging statistics in adaptation}
+In this short essay, we sketch four key, recurrent themes of Holland's
+work on adaptive systems: (1) the evolutionary dynamics of ``building
+blocks''; (2) learning via credit assignment and rule discovery; (3)
+the emergence of internal models; and (4) the generic properties of
+complex adaptive systems.  We discuss the role these themes have
+played in computer science, and highlight especially the ideas that we
+think remain relevant to today's research agendas.
 
-\section{ Learning and Credit assignment}
+\section{Evolutionary Dynamics of ``Building Blocks''}
 
-\section{Modeling}
+Holland's goal of developing a general theory of adaptation was
+spurred both by his early work on computer models of Hebbian learning
+\cite{RochesterHolland} and his reading of Ronald Fisher's classic
+work on integrating genetics with Darwinian selection \cite{Genetical
+  Theory of Natural Selection}.  As Holland further read extensively
+in evolutionary biology, economics, game theory, and control theory,
+he came to recognize that adaptation was the most central concept in
+all these fields.  That is, these fields all concern populations of
+agents that must continually obtain information from uncertain
+changing environments and use it in order to improve their performance
+with respect to those environments.  Moreover, Holland recognized that
+in systems with {\em adaptive agents} of this kind, there is never a
+state of {\emph equilibrium} or a final {\emph optimum} configuration.
+Due to the environment's ``perpetual novelty'' (as Holland termed it),
+adaptation continues forever in an open-ended way.  Holland's focus on
+open-ended, non-equilibrium dynamics was in stark contrast with the
+mainstream approach (at the time) in all these fields---the belief
+that ``solving for'' stable equilibrium dynamics was the scientific
+goal.  Holland's contrary view was that a system in stable equilibrium
+is essentially {\emph dead}.
 
-Modeling is central to Holland's theory of adaptive systems.
-He posits that all adaptive systems create and use internal models to survive in their environments, through anticipation.
-Models can be tacit and learned through evolutionary time, as in the case of bacteria swimming up a chemical gradient, or explicit and learned over a single lifespan as in the case of cognitive systems that incorporate experience into internal rule sets through learning.
-In \underline{Induction}, Holland and his co-authors discuss how such internal models can be learned, without supervision, by combining environmental inputs with stored knowledge.
-A key idea is that a model defines an equivalence relation over a set of environmental states, together with a set of transition rules, which are learned over time based on environmental inputs.
-Models that form valid homomorphisms with the environment allow the system to make accurate predictions.
-In his conception, the equivalence classes are initially very general, say 'moving object' and 'stationary object,' and over time through experience and learning they are specialized into more useful and precise subclasses, say 'insect' and 'nest', which form a default hierarchy and are associated with more specific transition rules.
-Although the idea of default hierarchies is prevalent in many knowledge representation systems (LIST A FEW?), Holland made two key contributions, first by emphsizing homomorphisms as a formal way to evaluate model validity, and idea that dates back to Ross Ashby's \underline{An Introduction to Cybernetics} that Holland's student, Bernard Ziegler, developed into a formal theory of computer modeling and simulation.
-Today, even though computational modeling is a major application area of computing,  essential to most natural sciences and engineering disciplines, these early homorphic theories of modeling stand as the most elegant approach we know of for characterizing when a model is consistent with the environment and how an intelligent agent, human or artificial, can update the model to better reflect reality.
-Holland's second key contribution was describing a computational mechanism by which a cognitive system could iteratively build up a detailed and hierarchical model of its environment, as Melanie already discussed above.
+Underlying Holland's theory of adaptation are three core ideas:
 
-Given that Holland believed that the ability to learn and manipulate internal models was essential for any adapting system, it is no suprise that he viewed modeling as essential for scientific inquiry.
-Today, we use computational models in several distinct ways---as a tool for analyzing data in statistical models, through simulation as a method for discovering new knowledge, and as a way to explain or understand how a system works.
-This third use of models was dear to Holland's heart.
-In his view, the key to science was understanding the mechanisms that cause a system to behave in a certain way, an aspiration that goes well beyond data fitting methods, which typically focus only on describing the aggregate behavior of a system.
-For example, a statistical model that describes the boom and bust pattern of the stock market cannot address the underlying mechanisms that lead to these cycles, through the collective actions of many many individual buy/sell decisions.
-Similarly, the genetic algorithms for which he is so famous provide a simple computational framework in which to explore the dynamics of Darwinian evolution and whether the basic mechanisms of variation, differential reproduction, and heredity are sufficient to account for the richness of our natural world.
-This emphasis on exploratory models to build intuitions was an important theme of Holland's work, and he often quoted Eddington's remark on the occasion of the first experimental test of Einstein's theory of relativity: "The contemplation in natural science of a wider domain than the actual leads to a far better understanding of the actual."  Holland was interested in models that explored basic principles and mechanisms, even if they didn't make specific predcitions, models that could show generically how certain behaviors could be produced, for example how exponential growth might interact with finite resources.
-Holland pioneered a style of modeling that has come to be known as 'individual based' or 'agent based,' in which every component of a system is represented explicitly and has state, e.g., every trader in a stock market system or every cell in an immune system model, and each agent had its own behaviors, which it could update (learn) over time.
-These models are often defined over spatial structures, such as networks or simple grids, to capture the constraints of systems living under spatial constraints.
-An agent-based model, then, encodes a theory about the mechanisms that are relevant for producting the behavior of interest.
-Similar to expert systems, such models are especially useful for studying systems that don't have analytic mathematical descriptions and they can facilitate interdisicplinary collaborations because the underlying rules can be easily communicated.
-It should be noted that Holland's view of modeling is by no means typical.
-For example, in a textbook on computational modeling, the authors offer the following definition:  “Modeling is the application of methods to analyze complex, real-world problems in order to make predictions about what might happen with various actions.” [Shiflet and Shiflet, 2006].
-This sort of perspective completely rules out the kind of modeling that Holland was most interested in, namely exploratory modeling.
-Similarly, many dispute that models make any kind of scientific contribution:  “Models are metaphors that explain the world we don’t understand in terms of worlds we do.
-They are merely analogies, provide partial insight, stand on someone else’s feet.
-Theories stand on their own feet, and rely on no analogies.”  [Emanuel Derman, 2012].
-[NOW WE NEED A STRONG FINISHING SENTENCE TO RESCUE JHH STYLE MODELING.]
+\begin{itemize}
+\item {\bf Populations:} Adaptation occurs in populations of
+  individuals evolving (or learning) over time, in which statistics 
+  can be leveraged to direct population dynamics.  (More on this below.)
+
+\item {\bf Building Blocks:} In a population undergoing adaptation,
+  individuals can be decomposed into {\emph building blocks}---sets of
+  traits that are the evolutionary ``atoms'' of an individual's
+  fitness or performance. As an example from biology, Holland gives
+  the {\emph Krebs cycle}, a core cellular metabolic
+  pathway. Similarly, in neuroscience, a building block of visual
+  perception might be a group of neurons that responds to an oriented
+  edge; in computer science, a building block of a sorting program
+  might be a ``swap'' operator.
+
+\item {\bf Exploitation Versus Exploration:} Successful adaptation
+  requires the right tradeoff between {\emph exploitation}, in which
+  tried-and-true building blocks propagate in a population, and {\emph
+    exploration}, in which existing building blocks are recombined or
+  mutated in new ways.  A system that focuses too much on exploitation
+  risks never finding better individuals or not being able to adapt to
+  changing environments.  Conversely, too much focus on exploration
+  means that the system won't leverage the successful building blocks
+  that have already been discovered.  There is some optimal balance
+  between these two modes of processing.
+
+\end{itemize}
+
+Several of Holland's early papers (e.g.,
+\cite{OutlineLogicalTheory,OptimalAllocation}) and his influential
+1975 book {\emph Adaptation in Natural and Artificial Systems}
+\cite{ANAS} developed a general, formal setting in which these ideas
+could be expressed mathematically.
+
+In particular, inspired by Bellman \cite{AdaptiveControlProcesses} and
+others, Holland explored the exploitation-versus-exploration tradeoff
+via an idealized ``two-armed bandit'' problem.  Given a slot machine
+with two arms, each of which has an unknown payoff probability, how
+should you allocate $N$ trials (pulls) between the arms so as to
+maximize your total payoff?  An example of an extreme {\emph
+  exploitation} strategy would be to alternate between the arms until
+one of them gives a payoff, and then allocate all future trials to
+that arm alone.  Conversely, an example of an extreme {\emph
+  exploration} strategy would be to randomly allocate the trials,
+irrespective of the payoff rates you obtain.  Obviously, each of these
+strategies is flawed.  What is an optimal strategy?
+
+In \cite{OptimalAllocationOfTrials} and \cite{ANAS}, Holland derived
+an equation for such a strategy.  Let $A$ denote the arm currently
+observed to have the higher payoff probability, and $B$ denote the
+other arm.  Holland showed that the optimal strategy (the one that
+yields maximal payoff, or minimal loss) is for the number of trials
+allocated to $A$ to grow slightly faster than an exponential function
+of the number of trials allocated to $B$.  Holland then showed that this
+result extends naturally to the {\emph multi-armed bandit} case.
+
+In an population undergoing adaptation, the building blocks making up
+individuals can be compared to the arms of a multi-armed bandit.  Each
+building block can be said to have a probability of ``payoff'' (i.e.,
+contribution to a given individual's fitness).  Evaluating an
+individual in an environment is like ``pulling the arms'' on a
+multi-armed bandit, where the ``arms'' correspond to each of the
+building blocks making up that individual.  Thus, in the process of
+assigning a fitness to each individual in a population, adaptive
+evolution can be seen as implicitly sampling the many building blocks
+making up that collection of individuals.  The average fitness over
+many individuals containing a particular building block gives an
+estimate of that building block's payoff probability.
+
+The question of how to balance exploitation and exploration---how to
+optimally allocate trials to different arms based on their observed
+payoff---now becomes the question of how to optimally sample in the
+vast space of possible building blocks, based on their observed
+average fitness.  Of course adaptive evolution deals in populations of
+individuals, not building blocks.  There is no explicit mechanism for
+keeping statistics on building blocks' ``observed average fitness''.
+However, Holland's central idea here is that an approximation to
+optimal building-block sampling does indeed occur, as an emergent
+property of the population dynamics.
+
+To show this mathematically, Holland defined an idealization of
+adaptive evolution: an algorithm he called a {\emph reproductive
+  plan}.  A reproductive plan is a population-based stochastic process
+operating on bit strings, involving the ``genetic'' operators of
+reproduction, fitness-based selection, crossover, mutation, and
+inversion.  Building blocks are formalized as {\emph
+  schemata}---patterns of bits within a string. (Here are two examples
+of schemas: ``all strings beginning with the bits 10'' or ``all
+strings that start with the pattern 1*1'', where * can be replaced by
+either 0 or 1).
+
+Holland's most important result was the following: He proved
+mathematically that his reproductive plan results---implicitly---in a
+near-optimal allocation of trials to schemata, thus optimizing the
+exploitation versus exploration balance.  Holland noted that the
+explicit act of assigning fitnesses to the individuals in a population
+was actually implicitly sampling a much larger collection of building
+blocks.  Holland termed this {\emph implicit parallelism}; the
+parallel leveraging of statistics from implicit sampling is a main
+strength of his population-based approaches to search.
+
+It was this formal setting and resulting theorems that led to the
+invention of genetic algorithms, which featured stochastic
+population-based search and {\emph crossover} as a critical
+operation: Crossover recombined successful building blocks and
+thus allowed them to be tested in new contexts.  What made GAs unique
+among other evolution-inspired algorithms at the time were the
+mathematical foundations described above, including bit-string
+representations, the focus on recombination as a central mechanism,
+and a focus (at least in Holland's mind) on continual adaptation to
+non-stationary environments rather than optimization to static
+environments.
+
+It is important to point out again that the framework Holland
+developed in \cite{ANAS} was more general than the genetic algorithm;
+it was created in order to develop an interdisciplinary theory of
+adaptation, one that would inform biology, say, as much as computer
+science \cite{ChristiansenFeldmanPaper}.  The later, successful
+application of genetic algorithms to real-world optimization and
+learning tasks was, for Holland, just icing on the cake.
+
+\section{Learning Via Credit Assignment and Rule Discovery}
+
+\section{Internal Models}
+
+The concept of {\emph internal models} is central to Holland's theory
+of adaptive systems.  He posits that all adaptive systems create and
+use internal models to survive in their environments, through
+anticipation.  Models can be tacit and learned through evolutionary
+time, as in the case of bacteria swimming up a chemical gradient, or
+explicit and learned over a single lifespan as in the case of
+cognitive systems that incorporate experience into internal rule sets
+through learning.  In \underline{Induction}, Holland and his
+co-authors discuss how such internal models can be learned, without
+supervision, by combining environmental inputs with stored knowledge.
+A key idea is that a model defines an equivalence relation over a set
+of environmental states, together with a set of transition rules,
+which are learned over time based on environmental inputs.  Models
+that form valid homomorphisms with the environment allow the system to
+make accurate predictions.  In his conception, the equivalence classes
+are initially very general, say 'moving object' and 'stationary
+object,' and over time through experience and learning they are
+specialized into more useful and precise subclasses, say 'insect' and
+'nest', which form a default hierarchy and are associated with more
+specific transition rules.  Although the idea of default hierarchies
+is prevalent in many knowledge representation systems (LIST A FEW?),
+Holland made two key contributions, first by emphsizing homomorphisms
+as a formal way to evaluate model validity, and idea that dates back
+to Ross Ashby's \underline{An Introduction to Cybernetics} that
+Holland's student, Bernard Ziegler, developed into a formal theory of
+computer modeling and simulation.  Today, even though computational
+modeling is a major application area of computing, essential to most
+natural sciences and engineering disciplines, these early homorphic
+theories of modeling stand as the most elegant approach we know of for
+characterizing when a model is consistent with the environment and how
+an intelligent agent, human or artificial, can update the model to
+better reflect reality.
+
+Holland's second key contribution was describing a computational
+mechanism by which a cognitive system could iteratively build up a
+detailed and hierarchical model of its environment, as Melanie already
+discussed above.
+
+Given that Holland believed that the ability to learn and manipulate
+internal models was essential for any adapting system, it is no
+suprise that he viewed modeling as essential for scientific inquiry.
+Today, we use computational models in several distinct ways---as a
+tool for analyzing data in statistical models, through simulation as a
+method for discovering new knowledge, and as a way to explain or
+understand how a system works.  This third use of models was dear to
+Holland's heart.  In his view, the key to science was understanding
+the mechanisms that cause a system to behave in a certain way, an
+aspiration that goes well beyond data fitting methods, which typically
+focus only on describing the aggregate behavior of a system.  For
+example, a statistical model that describes the boom and bust pattern
+of the stock market cannot address the underlying mechanisms that lead
+to these cycles, through the collective actions of many many
+individual buy/sell decisions.  Similarly, the genetic algorithms for
+which he is so famous provide a simple computational framework in
+which to explore the dynamics of Darwinian evolution and whether the
+basic mechanisms of variation, differential reproduction, and heredity
+are sufficient to account for the richness of our natural world.  This
+emphasis on exploratory models to build intuitions was an important
+theme of Holland's work, and he often quoted Eddington's remark on the
+occasion of the first experimental test of Einstein's theory of
+relativity: "The contemplation in natural science of a wider domain
+than the actual leads to a far better understanding of the actual."
+Holland was interested in models that explored basic principles and
+mechanisms, even if they didn't make specific predcitions, models that
+could show generically how certain behaviors could be produced, for
+example how exponential growth might interact with finite resources.
+Holland pioneered a style of modeling that has come to be known as
+'individual based' or 'agent based,' in which every component of a
+system is represented explicitly and has state, e.g., every trader in
+a stock market system or every cell in an immune system model, and
+each agent had its own behaviors, which it could update (learn) over
+time.  These models are often defined over spatial structures, such as
+networks or simple grids, to capture the constraints of systems living
+under spatial constraints.  An agent-based model, then, encodes a
+theory about the mechanisms that are relevant for producting the
+behavior of interest.  Similar to expert systems, such models are
+especially useful for studying systems that don't have analytic
+mathematical descriptions and they can facilitate interdisicplinary
+collaborations because the underlying rules can be easily
+communicated.  It should be noted that Holland's view of modeling is
+by no means typical.  For example, in a textbook on computational
+modeling, the authors offer the following definition: “Modeling is the
+application of methods to analyze complex, real-world problems in
+order to make predictions about what might happen with various
+actions.” [Shiflet and Shiflet, 2006].  This sort of perspective
+completely rules out the kind of modeling that Holland was most
+interested in, namely exploratory modeling.  Similarly, many dispute
+that models make any kind of scientific contribution: “Models are
+metaphors that explain the world we don’t understand in terms of
+worlds we do.  They are merely analogies, provide partial insight,
+stand on someone else’s feet.  Theories stand on their own feet, and
+rely on no analogies.”  [Emanuel Derman, 2012].  [NOW WE NEED A STRONG
+  FINISHING SENTENCE TO RESCUE JHH STYLE MODELING.]
 
 \section{Complexity}
+
+\section{Relevance to Modern CS}
+
+\begin{itemize}
+\item Exploitation versus exploration -- relevance to reinforcement
+  learning, other parts of machine learning and optimization.
+\item Active learning.  (Two-armed bandit problem.) 
+\item On-line learning.  
+\end{itemize}
 

### 2015-10-13 paper.tex
Version with learning section plus other edits added

@@
 
+While genetic algorithms are nowadays often associated with models of
+biological evolution, in Holland's view the processes of adaptation he
+was studying manifested themselves in many different complex systems.
+One system of special interest to Holland was the human mind, and in
+particular, its capacity to learn via induction.
+
+In the 1970s and 80s Holland formulated and extensively explored a
+particular model of decision-making, action, and inductive learning,
+called the {\emph classifier system}
+(e.g., \cite{Holland&Reitman1978,EscapingBrittlenessPaper}.  In a
+classifier system, a population of ``if-then'' rules interacts with an
+environment, and over time is able to learn to improve its performance
+via both {\emph credit-assignment} and {\emph rule-discovery}
+algorithms.
+
+Like the genetic algorithm, classifier systems can be viewed both as a
+model of adaptation and as an artificial-intelligence (AI) method.  At
+the time Holland was developing classifier systems, the field of AI
+was focused on expert systems, which typically did not learn on their
+own, a fundamental deficiency in Holland's view: ``[Expert] systems
+are brittle in the sense that they respond appropriately only in
+narrow domains, requiring substantial human intervention to compensate
+for even slight shifts in domain'' \cite{EscapingBrittleness}. In
+contrast, Holland proposed that ``{\emph induction} is the basic, and
+perhaps only, way of making large advances in this direction.''
+
+Holland's inspiration for classifier systems came from several
+different disciplines.  Knowledge representation in the form of a
+population of ``if-then'' rules seemed like a good choice, not only
+because of its popularity in AI at the time but also from Holland's
+early work on modeling Hebbian cell assemblies: ``In Hebbs view, a
+cell assembly makes a simple statement: If such and such an event
+occurs, then I will fire for a while at a high
+rate.'' \cite{Quoted-in-Waldrop-Book-P.182}.
+
+The if-then rules, when activated, compete to post their results on a
+shared ``message list'', modeling the system's short-term memory
+(again inspired by Hebb's work).  The competition was based on an
+bidding process, in which a rule's current strength affected its
+ability to bid highly.  Successful rules were gradually strengthened
+(and unsuccessful rules weakened) via a credit-assignment method
+called the {\emph bucket-brigade} algorithm, in which rules gaining
+rewards from the environment or from other rules would pay back part
+of their gains to those earlier-firing ``stage-setting'' rules that
+set up the conditions for the eventual reward.  The idea for the
+bucket-brigade algorithm was inspired by the mechanisms of Hebbian
+learning and by early work on learning and optimization in economics,
+psychology, control theory, and other fields (e.g.,
+\cite{Bellman1957,Samuel1959}). \footnote{Of course similar
+  credit-assignment methods have been used extensively in
+  reinforcement learning, neural networks, and other areas of modern
+  machine learning, but Holland was among the first to propose such
+  methods.}
+
+Finally, new rules for the population are discovered by a genetic
+algorithm, using the strength of rules as a fitness function.
+
+Classifier systems are fairly complex architectures, with learning
+occurring at multiple time scales.  Nevertheless, such systems have
+been usefully applied to areas as diverse as
+poker-playing \cite{Smith1980}, control of gas pipeline
+transmission \cite{Goldberg1983}, and modeling the stock
+market \cite{PalmerEtAl}.  Modern reinforcement learning methods have
+largely taken over as more useful approaches to real-world decision
+and control problems, but classifier systems can perhaps be thought of
+as an essential ``stage-setting'' method that enabled the development
+of later approaches.
+
+John Holland's main interest in classifier systems was how the two
+learning mechanisms could work together to create subsets of rules
+that accurately model the environment, and thus are able to ``fire''
+at appropriate times.  An important part of such internal models is
+the notion of a {\emph default hierarchy}---a set of rules that
+together form a kind of hierarchical decision structure, going from
+most general to most specific.  For example, a general rule might
+state that, ``If X is a bird, then X can fly'', whereas a more specifc
+one might state, ``If X is a bird with small wings and a large body,
+then X cannot fly'' \cite{InductionBook}.  When faced with a bird in
+the environment, both rules might compete to post their conclusion,
+but the more specific one would be favored if both of its conditions are met.
+
+Holland envisioned that, with a complex environment and enough time
+for learning, a classifier system might learn an intricate default
+hierarchy of rules, and that the competition inherent in the learning
+and action mechanisms would allow the system to adapt to a continually
+changing environment without losing what it has learned in the past.
+Holland put it this way: ``Competition among rules provides the system
+with a graceful way of handling perpetual novelty.  When a system has
+strong rules that respond to a particular situation, that is the
+equivalent of saying that it has certain well-validated hypotheses.
+Offspring rules, which begin life weaker than do their parents, can
+win the competition and influence the system's behavior only when
+there are not strong rules whose conditions are satisfied---in other
+words when the system does not know what to do.  If their actions
+help, they survive; if not they are soon replaced.  Thus, the
+offspring do not interfere with the system's action in well-practiced
+situations but wait gracefully in the wings as hypotheses about what
+to do under novel circumstances." \cite{HollandSciAmArticle}.
+
+[MM: Default hierarchy emerged in Goldberg's application?  See Holland's discussion.]
+
 \section{Internal Models}
@@
 
+begin{itemize}
+\item Evolutionary computation
+\item Q Learning
+\item Backprop
+\item    From Turing nomination: "Holland's machine learning system known as the
+   Learning Classifier System (LCS), developed in the early 1980s,
+   incorporated a reinforcement learning algorithm known as the bucket
+   brigade for non-Markovian environments, anticipating by nearly a
+   decade non-Markovian learning algorithms."
+\end{itemize}
+
 \begin{itemize}

### 2015-10-21 paper.tex
MM edits
Did significant editing, plus added all references, and .bib file.

@@
 
-In this short essay, we sketch four key, recurrent themes of Holland's
+In this short essay, we sketch five key, recurrent themes of Holland's
 work on adaptive systems: (1) the evolutionary dynamics of ``building
 blocks''; (2) learning via credit assignment and rule discovery; (3)
-the emergence of internal models; and (4) the generic properties of
-complex adaptive systems.  We discuss the role these themes have
-played in computer science, and highlight especially the ideas that we
-think remain relevant to today's research agendas.
+the emergence of internal models; (4) the art of scientific modeling;
+and (5) the universal properties of complex adaptive systems.  We
+discuss the role these themes have played in computer science, and
+highlight especially the ideas that we think remain relevant to
+today's research agendas.
 
@@
 spurred both by his early work on computer models of Hebbian learning
-\cite{RochesterHolland} and his reading of Ronald Fisher's classic
-work on integrating genetics with Darwinian selection \cite{Genetical
-  Theory of Natural Selection}.  As Holland further read extensively
-in evolutionary biology, economics, game theory, and control theory,
-he came to recognize that adaptation was the most central concept in
-all these fields.  That is, these fields all concern populations of
-agents that must continually obtain information from uncertain
-changing environments and use it in order to improve their performance
-with respect to those environments.  Moreover, Holland recognized that
-in systems with {\em adaptive agents} of this kind, there is never a
-state of {\emph equilibrium} or a final {\emph optimum} configuration.
-Due to the environment's ``perpetual novelty'' (as Holland termed it),
-adaptation continues forever in an open-ended way.  Holland's focus on
-open-ended, non-equilibrium dynamics was in stark contrast with the
-mainstream approach (at the time) in all these fields---the belief
-that ``solving for'' stable equilibrium dynamics was the scientific
-goal.  Holland's contrary view was that a system in stable equilibrium
-is essentially {\emph dead}.
+\cite{Rochester1956} and his reading of Ronald Fisher's classic work
+on integrating genetics with Darwinian selection \cite{Fisher1930}.
+As Holland further read extensively in evolutionary biology,
+economics, game theory, and control theory, he came to recognize that
+adaptation was the most central concept in all these fields.  That is,
+these fields all concern populations of agents that must continually
+obtain information from uncertain changing environments and use it in
+order to improve their performance with respect to those environments.
+Moreover, Holland recognized that in systems with {\em adaptive
+  agents} of this kind, there is never a state of {\emph equilibrium}
+or a final {\emph optimum} configuration.  Due to the environment's
+``perpetual novelty'' (as Holland termed it), adaptation continues
+forever in an open-ended way.  Holland's focus on open-ended,
+non-equilibrium dynamics was in stark contrast with the mainstream
+approach (at the time) in all these fields---the belief that ``solving
+for'' stable equilibrium dynamics was the scientific goal.  Holland's
+contrary view was that a system in stable equilibrium is essentially
+{\emph dead}.
 
@@
   between these two modes of processing.
-
 \end{itemize}
@@
 Several of Holland's early papers (e.g.,
-\cite{OutlineLogicalTheory,OptimalAllocation}) and his influential
+\cite{Holland1962,Holland1973}) and his influential
 1975 book {\emph Adaptation in Natural and Artificial Systems}
-\cite{ANAS} developed a general, formal setting in which these ideas
+\cite{Holland1975} developed a general, formal setting in which these ideas
 could be expressed mathematically.
 
-In particular, inspired by Bellman \cite{AdaptiveControlProcesses} and
+In particular, inspired by Bellman \cite{Bellman1961} and
 others, Holland explored the exploitation-versus-exploration tradeoff
@@
 
-In \cite{OptimalAllocationOfTrials} and \cite{ANAS}, Holland derived
+In \cite{Holland1973} and \cite{Holland1975}, Holland derived
 an equation for such a strategy.  Let $A$ denote the arm currently
@@
 It is important to point out again that the framework Holland
-developed in \cite{ANAS} was more general than the genetic algorithm;
+developed in \cite{Holland1975} was more general than the genetic algorithm;
 it was created in order to develop an interdisciplinary theory of
 adaptation, one that would inform biology, say, as much as computer
-science \cite{ChristiansenFeldmanPaper}.  The later, successful
+science \cite{Christiansen1998}.  The later, successful
 application of genetic algorithms to real-world optimization and
@@
 
-While genetic algorithms are nowadays often associated with models of
+While genetic algorithms are nowadays most often associated with models of
 biological evolution, in Holland's view the processes of adaptation he
@@
 particular model of decision-making, action, and inductive learning,
-called the {\emph classifier system}
-(e.g., \cite{Holland&Reitman1978,EscapingBrittlenessPaper}.  In a
-classifier system, a population of ``if-then'' rules interacts with an
-environment, and over time is able to learn to improve its performance
-via both {\emph credit-assignment} and {\emph rule-discovery}
-algorithms.
+called the {\emph classifier system} (e.g.,
+\cite{Holland1977,Holland1986}.  In a classifier system, a population
+of ``if-then'' rules interacts with an environment, and over time is
+able to learn to improve its performance via both {\emph
+  credit-assignment} and {\emph rule-discovery} algorithms.
 
-Like the genetic algorithm, classifier systems can be viewed both as a
-model of adaptation and as an artificial-intelligence (AI) method.  At
+Like genetic algorithms, classifier systems can be viewed both as 
+models of adaptation and as artificial-intelligence (AI) methods.  At
 the time Holland was developing classifier systems, the field of AI
@@
 narrow domains, requiring substantial human intervention to compensate
-for even slight shifts in domain'' \cite{EscapingBrittleness}. In
+for even slight shifts in domain'' \cite{Holland1986}. In
 contrast, Holland proposed that ``{\emph induction} is the basic, and
@@
 occurs, then I will fire for a while at a high
-rate.'' \cite{Quoted-in-Waldrop-Book-P.182}.
+rate.'' \cite{Waldrop1993} (p. 182).
 
@@
 psychology, control theory, and other fields (e.g.,
-\cite{Bellman1957,Samuel1959}). \footnote{Of course similar
+\cite{Bellman1961,Samuel1959}). \footnote{Of course similar
   credit-assignment methods have been used extensively in
@@
 transmission \cite{Goldberg1983}, and modeling the stock
-market \cite{PalmerEtAl}.  Modern reinforcement learning methods have
+market \cite{Palmer1994}.  Modern reinforcement learning methods have
 largely taken over as more useful approaches to real-world decision
@@
 most general to most specific.  For example, a general rule might
-state that, ``If X is a bird, then X can fly'', whereas a more specifc
+state that, ``If X is a bird, then X can fly'', whereas a more specific
 one might state, ``If X is a bird with small wings and a large body,
-then X cannot fly'' \cite{InductionBook}.  When faced with a bird in
+then X cannot fly'' \cite{Holland1989}.  When faced with a bird in
 the environment, both rules might compete to post their conclusion,
@@
 situations but wait gracefully in the wings as hypotheses about what
-to do under novel circumstances." \cite{HollandSciAmArticle}.
+to do under novel circumstances." \cite{Holland1992}.  A detailed
+discussion of some real-world applications of classifier systems is
+given in \cite{Booker1989}.
 
-[MM: Default hierarchy emerged in Goldberg's application?  See Holland's discussion.]
-
-\section{Internal Models}
+\section{Internal Models and Default Hierarchies}
 
@@
 of adaptive systems.  He posits that all adaptive systems create and
-use internal models to survive in their environments, through
+use internal models to prosper in their environments, through
 anticipation.  Models can be tacit and learned through evolutionary
 time, as in the case of bacteria swimming up a chemical gradient, or
-explicit and learned over a single lifespan as in the case of
+explicit and learned over a single lifespan, as in the case of
 cognitive systems that incorporate experience into internal rule sets
-through learning.  In \underline{Induction}, Holland and his
-co-authors discuss how such internal models can be learned, without
-supervision, by combining environmental inputs with stored knowledge.
-A key idea is that a model defines an equivalence relation over a set
-of environmental states, together with a set of transition rules,
-which are learned over time based on environmental inputs.  Models
-that form valid homomorphisms with the environment allow the system to
-make accurate predictions.  In his conception, the equivalence classes
-are initially very general, say 'moving object' and 'stationary
-object,' and over time through experience and learning they are
-specialized into more useful and precise subclasses, say 'insect' and
-'nest', which form a default hierarchy and are associated with more
-specific transition rules.  Although the idea of default hierarchies
-is prevalent in many knowledge representation systems (LIST A FEW?),
-Holland made two key contributions, first by emphsizing homomorphisms
-as a formal way to evaluate model validity, and idea that dates back
-to Ross Ashby's \underline{An Introduction to Cybernetics} that
-Holland's student, Bernard Ziegler, developed into a formal theory of
-computer modeling and simulation.  Today, even though computational
-modeling is a major application area of computing, essential to most
-natural sciences and engineering disciplines, these early homorphic
-theories of modeling stand as the most elegant approach we know of for
-characterizing when a model is consistent with the environment and how
-an intelligent agent, human or artificial, can update the model to
-better reflect reality.
+through learning.
+
+In \cite{Holland1989}, Holland and his co-authors discuss how such
+internal models can be learned, without supervision, by combining
+environmental inputs with stored knowledge.  A key idea is that a
+model defines sets of equivalence relations over environmental states,
+together with a set of transition rules, which are learned over time
+based on environmental ``rewards'' (or ``punishments'').  Models that
+form valid homomorphisms with the environment allow the system to make
+accurate predictions.  In Holland's conception, the equivalence
+classes are initially very general, say ``moving object'' and
+``stationary object,'' and over time, through experience and learning,
+they are specialized into more useful and precise subclasses, say
+``insect'' and ``nest''.  As described above, the adaptive system
+eventually learns a default hierarchy of rules involving both general
+and specific classes.
+
+Although the idea of default hierarchies is prevalent in many
+knowledge representation systems (LIST A FEW?), Holland made two key
+contributions.  The first was his emphasis of homomorphisms as a formal way to
+evaluate model validity.  This idea that dates back to Ross Ashby's
+\emph{An Introduction to Cybernetics} \cite{Ashby1956}. Holland's student,
+Bernard Ziegler developed this idea into a formal theory of computer modeling
+and simulation \cite{Ziegler1976}.  
+%Today, even though computational modeling is a major
+%application area of computing, essential to most natural sciences and
+%engineering disciplines
+Even today, these early homomorphic theories of modeling
+stand as the most elegant approach we know of for characterizing when
+a model is consistent with the environment and how an intelligent
+agent, human or artificial, can update the model to better reflect
+reality.
 
 Holland's second key contribution was describing a computational
-mechanism by which a cognitive system could iteratively build up a
-detailed and hierarchical model of its environment, as Melanie already
-discussed above.
+mechanism---classifier systems---by which a cognitive system could
+iteratively build up a detailed and hierarchical model of its
+environment.  The key learning elements of this method, the
+bucket-brigade algorithm, combined with a genetic algorithm, as
+described above, presaged many of the ideas in modern reinforcement
+learning.
+
+\section{The Art of Scientific Modeling}
 
@@
 internal models was essential for any adapting system, it is no
-suprise that he viewed modeling as essential for scientific inquiry.
-Today, we use computational models in several distinct ways---as a
-tool for analyzing data in statistical models, through simulation as a
-method for discovering new knowledge, and as a way to explain or
-understand how a system works.  This third use of models was dear to
-Holland's heart.  In his view, the key to science was understanding
-the mechanisms that cause a system to behave in a certain way, an
-aspiration that goes well beyond data fitting methods, which typically
-focus only on describing the aggregate behavior of a system.  For
-example, a statistical model that describes the boom and bust pattern
-of the stock market cannot address the underlying mechanisms that lead
-to these cycles, through the collective actions of many many
+surprise that he viewed modeling as essential for scientific inquiry.
+
+%Today, we use computational models in several distinct ways---as
+%tools for analyzing data in statistical models; as methods for
+%discovering new knowledge; and as a means to understand how complex
+%systems works.  [Steph, I am not clear on the distinction between the
+%second and third things here, so I combined them.  Change back if you
+%want.]
+
+Today, we use computational models both for \emph{prediction}---by
+analyzing data via statistical models---and for \emph{understanding}
+how systems work---by probing the effects of hypothesized underlying
+mechanisms.  This latter use of models was dear to Holland's heart.
+In his view, the key to science was understanding the mechanisms that
+cause a system to behave in a certain way, an aspiration that goes
+well beyond data fitting methods, which typically focus only on
+describing the aggregate behavior of a system.
+
+For example, a statistical model that describes the boom and bust
+pattern of the stock market cannot address the underlying mechanisms
+that lead to these cycles, through the collective actions of myriad
 individual buy/sell decisions.  Similarly, the genetic algorithms for
-which he is so famous provide a simple computational framework in
+which Holland is so famous provide a simple computational framework in
 which to explore the dynamics of Darwinian evolution and whether the
 basic mechanisms of variation, differential reproduction, and heredity
-are sufficient to account for the richness of our natural world.  This
-emphasis on exploratory models to build intuitions was an important
-theme of Holland's work, and he often quoted Eddington's remark on the
-occasion of the first experimental test of Einstein's theory of
-relativity: "The contemplation in natural science of a wider domain
-than the actual leads to a far better understanding of the actual."
+are sufficient to account for the richness of our natural world.
+
+The emphasis on exploratory models to build intuitions was an
+important theme of Holland's work, and he often quoted Eddington's
+remark on the occasion of the first experimental test of Einstein's
+theory of relativity: "The contemplation in natural science of a wider
+domain than the actual leads to a far better understanding of the
+actual" \cite{Eddington1927}.
+
 Holland was interested in models that explored basic principles and
-mechanisms, even if they didn't make specific predcitions, models that
-could show generically how certain behaviors could be produced, for
-example how exponential growth might interact with finite resources.
+mechanisms, even if they didn't make specific or detailed predictions.
+Such models can show generically how certain behaviors could be produced.
 Holland pioneered a style of modeling that has come to be known as
@@
 system is represented explicitly and has state, e.g., every trader in
-a stock market system or every cell in an immune system model, and
-each agent had its own behaviors, which it could update (learn) over
-time.  These models are often defined over spatial structures, such as
-networks or simple grids, to capture the constraints of systems living
-under spatial constraints.  An agent-based model, then, encodes a
-theory about the mechanisms that are relevant for producting the
-behavior of interest.  Similar to expert systems, such models are
-especially useful for studying systems that don't have analytic
-mathematical descriptions and they can facilitate interdisicplinary
-collaborations because the underlying rules can be easily
-communicated.  It should be noted that Holland's view of modeling is
-by no means typical.  For example, in a textbook on computational
-modeling, the authors offer the following definition: “Modeling is the
-application of methods to analyze complex, real-world problems in
-order to make predictions about what might happen with various
-actions.” [Shiflet and Shiflet, 2006].  This sort of perspective
-completely rules out the kind of modeling that Holland was most
-interested in, namely exploratory modeling.  Similarly, many dispute
-that models make any kind of scientific contribution: “Models are
-metaphors that explain the world we don’t understand in terms of
-worlds we do.  They are merely analogies, provide partial insight,
-stand on someone else’s feet.  Theories stand on their own feet, and
-rely on no analogies.”  [Emanuel Derman, 2012].  [NOW WE NEED A STRONG
-  FINISHING SENTENCE TO RESCUE JHH STYLE MODELING.]
+a stock market system or every cell in an immune system model.  In
+such models, each agent has its own behavior rules, which it can
+update (or learn) over time.  In order to capture the constraints of
+systems living under spatial constraints, these models are often
+defined over spatial structures, such as networks or simple grids.
+
+A given agent-based model encodes a theory about the mechanisms that
+are relevant for producing the behavior of interest.  Similar to
+expert systems, such models are especially useful for studying systems
+that don't have analytic mathematical descriptions.  Agent-based
+models can facilitate interdisciplinary collaborations because the
+underlying rules can be easily communicated.  The agent-based models
+championed by Holland were typically idealized versions of complex
+systems and not intended to provide detailed, domain-specific
+predictions.  Instead they were meant to explore possible general
+mechanisms of complex systems and thus provide insights that might
+lead to more specific, detailed models.  Such idealized models are
+akin to what Dennett has called ``intuition pumps''
+\cite{Dennett1984}.
+
+It should be noted that Holland's view of modeling is by no means
+typical.  For example, in a textbook on computational modeling, the
+authors offer the following definition: “Modeling is the application
+of methods to analyze complex, real-world problems in order to make
+predictions about what might happen with various actions”
+\cite{Shiflet2014}.   This sort of perspective completely rules out the
+kind of exploratory modeling that Holland was most interested in.
+
+Some researchers dispute that models make any kind of scientific
+contribution: "Models are metaphors that explain the world we don't
+understand in terms of worlds we do.  They are merely analogies,
+provide partial insight, stand on someone else's feet.  Theories stand
+on their own feet, and rely on no analogies."  [Emanuel Derman, 2012].
+[STEPH, I COULD NOT FIND REFERENCE FOR THIS QUOTE OUTSIDE SFI VIDEO.]
+
+[NOW WE NEED A STRONG FINISHING SENTENCE TO RESCUE JHH STYLE
+  MODELING.]
 
@@
 
-begin{itemize}
+\begin{itemize}
 \item Evolutionary computation
@@
    decade non-Markovian learning algorithms."
-\end{itemize}
-
-\begin{itemize}
 \item Exploitation versus exploration -- relevance to reinforcement
@@
 
+\bibliographystyle{plain}
+
+\bibliography{paper}
+
 \end{document}

### 2015-10-23 complexity.tex
added complexity section and changed title

\section{Complex Adaptive Systems}

Holland was interested in a broad array of adaptive systems---immune systems, ecologies, financial markets, cities, and the brain---systems that are \emph{complex}.  In the 1980's, he teamed up with a small group of scientists, primarily physicists with a sprinkling of economists and biologists, to discuss what this wide swath of systems has in common.  The discussions helped define the intellectual mission of the Santa Fe Institute, the first institution dedicated to developing a science of complexity.  [add in how many complexity institutes there are worldwide and which others John played a hand in.]  Holland brought to these discussions his lifelong study of adaptation and a reminder that true theories about complexity would need to look deeper than phenomenological descriptions but also account for the  'how' and 'why' of these systems.
%describing phenomena like chaos, power laws, and ONE MORE, and acoount for the  
As the discussions matured, a consensus developed about the basic elements of complexity---systems that: (1)  are composed of many components with nontrivial (nonlinear) interactions; (2) behavior that emerges from the dynamics of the system, exhibiting higher-order patterns; (3) Scale, systems are nested and structure/behavior emerges at different scales, with some behavior being conserved across all scales and other behaviors changing at different scales, and (4) continual adaptation where systems adjust their behavioral rules through evolution and learning.  Although this is far from a formal definition of complex systems, most people working in the field today are interested in systems that have these properties.

To illustrate his ideas about the ubuiquity of adaptation, Holland developed two systems, ECHO and the SFI stock market.
Although less famous than genetic algorithms and learning classifier systems, ECHO was no less ambitious.  It's scope included the evolution of interacting populations of individual agents, which through competition and learning, discover symbiotic triads such as the famous ant-fly-caterpillar interaction,  the Wiksell triangle of trading relationships in economics, and the immune system's learned ability to distinguish 'self' from 'other.'  Without going into the details of the model [CITES HERE], ECHO formalized Holland's theories about complex adapting systems into a runnable computational system where agents evolved external markers (called tags) and internal preferences and used them to form higher level aggregate structures (trading relationships, symbiotic groups, trophic cascades, interdependent organizations etc.).  ECHO agents sometimes discovered mimicry to deceive competitors, and over time, the model developed increasingly complex structures and behaviors, often reproducing patterns observed in nature.  For example,  ECHO evolved a diversity of agent types, whose rank-frequency distribution closely parallels the well-known Preston curve in ecology, a quantiative statement of the adage that 'most species are rare'. Many of the insights behind this project are described in his book, [WHICH ONES]. The broad scope of this project was appealing to immunologists, economists, and evolutionary biologists alike, but it was the economists who took it one step further in the SFI stock market.

% Note: SF has the history backwards here.  We should correct that, but the story was so nice in this order

In the early 1990s, Holland teamed up with other SFI researchers, including two economists, to tackle the mismatch between predictions of the dominant theory in economics, the rational expectations hypothesis, and empirically observed stock market behaviors that deviated from the predicted equilibria solutions.  In brief, most economic theory of the day assumed that all participants in an economy or financial market are 100\% rational and act to maximize their individual gain.  We know, however, that real actors in real economies and markets are rarely rational, and financial markets often deviate from rationalilty, for example with speculative bubbles and crashes.  The SFI stock market project studied what happens in a model where rational traders are replaced by traders who learn to forecast stock prices over time.  The model was constructed to allow testing for the possible emergence of fundamental trading versus technical trading versus uninformed trading. The simulated market with (learning) trading agents was run many times and the dynamics of price and trading volumes were compared to observed patterns in real markets, replicating many of the features of real-life markets.  Although the model was primitive in many ways, it was wildly influential and led to many follow-on projects  It also demonstrated the important role that adaptation plays in complex systems and illustrated how Holland's theories of continual learning in response to intermittent feedbacks from the environment could be integrated into domain-specific settings.

%Blake LeBaron (economics);
% W. Brian Arthur (economics); 
% John Holland (psychology/EE/CS,
% and father of GAs);
% Richard Palmer (physics);
% Paul Taylor (computer science

%Many modeling issues not satisfactorily resolved by the SF-ASM model have been 
% taken up in later research (see Ref.[4]).

Holland's later books,  \emph{Emergence}, \emph{Signals and Boundaries}, and \emph{Complexity, A Summary} show how the theories of adaptation that he developed during the earlier part of his career fit into the larger landscape of complex systems research.   Holland's focus on how complex patterns emerge and change, rather than simply characterizing the patterns themselves (e.g., describing chaotic attractors or power laws) reflected his deterimination to 'get to the heart' of complex systems.  This determination 
represents the best of science.  Holland's willingness to tackle the hardest questions, develop his own formalisms, and use mathematics productively sets a high bar that we all should aspire to.

% Find the patterns.
% Why they occur is harder to say.  
% John focused on Mechanisms

%Ironically, Holland, didn't consider the computers of the day to be complex.

%The idea of a Nash equilibrium underlies much of economics; much of genetics, even today, considers only a single gene at a time;

% The basic elements included systems with many autonomous components acting in parallel, a focus on the dynamics of these systems and their nonlinear intereactions, and emergent behavior.  The basic tools included nonlinear dynamical systems theory, information theory, and statistical mechanics [MELANIE, ADD MORE HERE].  Holland contributed a mature and generic view of adaptive systems, pointing out that the most interesting complex systems were continually adapting.  He was instrumental in defining the intellectual vision for the Santa Fe Institute, the first of what today number about [HOW MANY] such institutes worldwide devoted to the sciences of complexity.

JOHN WAS INTERESTED IN ADAPTIVE SYSTEMS, LIKE CITIES, IMMUNE SYSTEMS, ECO-SYSTEMS, FINANCIAL MARKETS, E

AND, I REMEMBER HAVING ARGUMENTS WITH HIM ABOUT WHETHER A COMPUTER COULD PROPERLY BE VIEWED AS A COMPLEX SYSTEM.  I SAID ‘YES’, AND HE SAID ‘NO’.   HIS POINT WAS THAT COMPUTERS WERE INCAPABLE OF OPEN-ENDED EVOLUTION AND EXPLORATION LIKE THESE OTHER SYSTEMS.   IN SOME WAYS, MY ENTIRE CAREER HAS BEEN DEVOTED TO SHOWING THAT COMPUTERS
HAVE ACQUIRED AT LEAST SOME OF THE PROPERTIES OF COMPLEX SYSTEMS THAT HE WAS SO INTERESTED IN.

### 2015-10-23 paper.tex
added complexity section and changed title

@@
 
-\title{Adaptation Writ Large:\\ An Essay in Memory of John H. Holland}
+%\title{Adaptation Writ Large:\\ An Essay in Memory of JohnH. Holland}
+\title{Adaptative Computation:\\ An Essay in Memory of John H. Holland}
 
@@
 
+%\section{Complex Adaptive Systems}
+
+Holland was interested in a broad array of adaptive systems---immune systems, ecologies, financial markets, cities, and the brain---systems that are \emph{complex}.
+In the 1980's, he teamed up with a small group of scientists, primarily physicists with a sprinkling of economists and biologists, to discuss what this wide swath of systems has in common.
+The discussions helped define the intellectual mission of the Santa Fe Institute, the first institution dedicated to developing a science of complexity.
+[add in how many complexity institutes there are worldwide and which others John played a hand in.]  Holland brought to these discussions his lifelong study of adaptation and a reminder that true theories about complexity would need to look deeper than phenomenological descriptions but also account for the  'how' and 'why' of these systems.
+%describing phenomena like chaos, power laws, and ONE MORE, and acoount for the  
+As the discussions matured, a consensus developed about the basic elements of complexity---systems that: (1)  are composed of many components with nontrivial (nonlinear) interactions; (2) behavior that emerges from the dynamics of the system, exhibiting higher-order patterns; (3) Scale, systems are nested and structure/behavior emerges at different scales, with some behavior being conserved across all scales and other behaviors changing at different scales, and (4) continual adaptation where systems adjust their behavioral rules through evolution and learning.
+Although this is far from a formal definition of complex systems, most people working in the field today are interested in systems that have these properties.
+
+To illustrate his ideas about the ubuiquity of adaptation, Holland developed two systems, ECHO and the SFI stock market.
+Although less famous than genetic algorithms and learning classifier systems, ECHO was no less ambitious.
+It's scope included the evolution of interacting populations of individual agents, which through competition and learning, discover symbiotic triads such as the famous ant-fly-caterpillar interaction,  the Wiksell triangle of trading relationships in economics, and the immune system's learned ability to distinguish 'self' from 'other.'  Without going into the details of the model [CITES HERE], ECHO formalized Holland's theories about complex adapting systems into a runnable computational system where agents evolved external markers (called tags) and internal preferences and used them to form higher level aggregate structures (trading relationships, symbiotic groups, trophic cascades, interdependent organizations etc.).
+ECHO agents sometimes discovered mimicry to deceive competitors, and over time, the model developed increasingly complex structures and behaviors, often reproducing patterns observed in nature.
+For example,  ECHO evolved a diversity of agent types, whose rank-frequency distribution closely parallels the well-known Preston curve in ecology, a quantiative statement of the adage that 'most species are rare'.
+Many of the insights behind this project are described in his book, [WHICH ONES].
+The broad scope of this project was appealing to immunologists, economists, and evolutionary biologists alike, but it was the economists who took it one step further in the SFI stock market.
+
+% Note: SF has the history backwards here.  We should correct that, but the story was so nice in this order
+
+In the early 1990s, Holland teamed up with other SFI researchers, including two economists, to tackle the mismatch between predictions of the dominant theory in economics, the rational expectations hypothesis, and empirically observed stock market behaviors that deviated from the predicted equilibria solutions.
+In brief, most economic theory of the day assumed that all participants in an economy or financial market are 100\% rational and act to maximize their individual gain.
+We know, however, that real actors in real economies and markets are rarely rational, and financial markets often deviate from rationalilty, for example with speculative bubbles and crashes.
+The SFI stock market project studied what happens in a model where rational traders are replaced by traders who learn to forecast stock prices over time.
+The model was constructed to allow testing for the possible emergence of fundamental trading versus technical trading versus uninformed trading.
+The simulated market with (learning) trading agents was run many times and the dynamics of price and trading volumes were compared to observed patterns in real markets, replicating many of the features of real-life markets.
+Although the model was primitive in many ways, it was wildly influential and led to many follow-on projects  It also demonstrated the important role that adaptation plays in complex systems and illustrated how Holland's theories of continual learning in response to intermittent feedbacks from the environment could be integrated into domain-specific settings.
+
+%Blake LeBaron (economics);
+% W. Brian Arthur (economics); 
+% John Holland (psychology/EE/CS,
+% and father of GAs);
+% Richard Palmer (physics);
+% Paul Taylor (computer science
+
+%Many modeling issues not satisfactorily resolved by the SF-ASM model have been 
+% taken up in later research (see Ref.[4]).
+
+Holland's later books,  \emph{Emergence}, \emph{Signals and Boundaries}, and \emph{Complexity, A Summary} show how the theories of adaptation that he developed during the earlier part of his career fit into the larger landscape of complex systems research.
+Holland's focus on how complex patterns emerge and change, rather than simply characterizing the patterns themselves (e.g., describing chaotic attractors or power laws) reflected his deterimination to 'get to the heart' of complex systems.
+This determination
+represents the best of science.  Holland's willingness to tackle the hardest questions, develop his own formalisms, and use mathematics productively sets a high bar that we all should aspire to.
+
 \section{Relevance to Modern CS}

### 2015-10-27 paper.tex
tightened up 'evolutionary dynamics' and merged 'internal models with classifier systems

@@
   agents} of this kind, there is never a state of {\emph equilibrium}
-or a final {\emph optimum} configuration.  Due to the environment's
-``perpetual novelty'' (as Holland termed it), adaptation continues
-forever in an open-ended way.  Holland's focus on open-ended,
+or a final \emph{optimum} configuration.  Due to the environment's
+perpetual novelty,
+%SF thinks Holland did not invent this term (as Holland termed it), 
+adaptation is a continual 
+open-ended process.  Holland's focus on open-ended,
 non-equilibrium dynamics was in stark contrast with the mainstream
@@
 contrary view was that a system in stable equilibrium is essentially
-{\emph dead}.
+\emph{dead}.
+
+% Text that STEPH found in our NIPS paper on the IGA
+% First, at least theoretically the GA is fast because of implicit
+% parallelism (Holland, 1975/1992): each string in the population is an
+% instance of many different schemas, and if the population is large
+% enough and is initially chosen at random, a large number of different
+% schemas-many more than the number of strings in the population-are
+% being sampled in parallel.  This should result in a quick search for
+% short, low-order schemas that confer high fitness.  Second,
+% fitness-proportionate reproduction under the GA should conserve
+% instances of such schemas.  Third, a high crossover rate should
+% quickly combine instances oflow-order schemas on different strings to
+% create instances of longer schemas that confer even higher fitness.
 
 Underlying Holland's theory of adaptation are three core ideas:
+Population-based search, building blocks and recombination, and the
+exploration-exploitation tradeoff.
 
 \begin{itemize}
-\item {\bf Populations:} Adaptation occurs in populations of
-  individuals evolving (or learning) over time, in which statistics 
-  can be leveraged to direct population dynamics.  (More on this below.)
+\item{\bf Populations and implicit parallelism:} A population of
+  individuals can be thought of as a set of samples taken from a much larger
+  search space.  Importantly, the populations are competitive, and
+  the evolutionary dynamic biases the set of samples
+  over time towards high-fitness regions of the search space.
 
-\item {\bf Building Blocks:} In a population undergoing adaptation,
-  individuals can be decomposed into {\emph building blocks}---sets of
-  traits that are the evolutionary ``atoms'' of an individual's
-  fitness or performance. As an example from biology, Holland gives
-  the {\emph Krebs cycle}, a core cellular metabolic
-  pathway. Similarly, in neuroscience, a building block of visual
-  perception might be a group of neurons that responds to an oriented
-  edge; in computer science, a building block of a sorting program
-  might be a ``swap'' operator.
+%The population can be
+%  thought of as a set of samples in a much larger search space, where
+%  the processes of evolution
+%  population can be thought of as a set of sa in which statistics 
+%  can be leveraged to direct population dynamics.  (More on this below.)
 
-\item {\bf Exploitation Versus Exploration:} Successful adaptation
-  requires the right tradeoff between {\emph exploitation}, in which
+\item{\bf Building Blocks and recombination:} High-fitness individuals
+  are discovered in stages, first by finding high-fitness partial
+  solutions (the building blocks) through random sampling, and over
+  time assembling the partial solutions into more complete solutions
+  with even higher fitness.  For example, when searching the space of
+  all $n$-bit binary strings, almost any random sample would contain
+  examples of strings whose first bit is set to 1, and also many
+  samples of strings whose second bit is set to 1, but there would be
+  many fewer samples that have both of the first two bits set to
+  1.  More compelling examples include the \emph{Krebs cycle}, a core
+  cellular metabolic pathway, a group of neurons that responds to an
+  oriented edge in vision, or the ``swap'' operator in a sorting
+  algorithm.
+
+ To succeed, such a search requires first that
+  high-fitness subparts be preserved once found, and second, that
+  there be an efficient mechanism for combining promising building
+  blocks into higher-order building blocks.  Holland proposed using an
+  analog of Darwinian selection for the first and the idea of crossing
+  over in genetics for the second.  Holland referred to the building
+  blocks as \emph{schemas,} which correspond mathematically to hyperplanes in the
+  search space.
+
+%In a population undergoing adaptation,
+%  individuals can be decomposed into {\emph building blocks}---sets of
+% traits that are the evolutionary ``atoms'' of an individual's
+%  fitness or performance.
+
+\item{\bf Exploitation Versus Exploration:} Successful adaptation
+  requires maintaining a balance between {\emph exploitation}, in which
   tried-and-true building blocks propagate in a population, and {\emph
     exploration}, in which existing building blocks are recombined or
-  mutated in new ways.  A system that focuses too much on exploitation
-  risks never finding better individuals or not being able to adapt to
-  changing environments.  Conversely, too much focus on exploration
-  means that the system won't leverage the successful building blocks
-  that have already been discovered.  There is some optimal balance
-  between these two modes of processing.
+  mutated in new ways.  
+%A system that focuses too much on exploitation
+%  risks never finding better individuals or not being able to adapt to
+%  changing environments.  Conversely, too much focus on exploration
+%  means that the system won't leverage the successful building blocks
+%  that have already been discovered.  %There is some optimal balance
+%  between these two modes of processing.
 \end{itemize}
@@
 
-In particular, inspired by Bellman \cite{Bellman1961} and
+Inspired by Bellman \cite{Bellman1961} and
 others, Holland explored the exploitation-versus-exploration tradeoff
@@
 strategies is flawed.  What is an optimal strategy?
+In \cite{Holland1973,Holland1975} Holland derived
+an equation for such a strategy and argued that the optimal strategy
+allocated trials to the observed best arm at a slightly higher rate than
+exponential.  
+% SF used 'argue' because some claim that his derivation was wrong.
+He then extended this 
+% Let $A$ denote the arm currently
+%observed to have the higher payoff probability, and $B$ denote the
+%other arm.  Holland showed that the optimal strategy (the one that
+%yields maximal payoff, or minimal loss) is for the number of trials
+%allocated to $A$ to grow slightly faster than an exponential function
+%of the number of trials allocated to $B$.  Holland then showed that this
+result to the {\emph multi-armed bandit} case.
 
-In \cite{Holland1973} and \cite{Holland1975}, Holland derived
-an equation for such a strategy.  Let $A$ denote the arm currently
-observed to have the higher payoff probability, and $B$ denote the
-other arm.  Holland showed that the optimal strategy (the one that
-yields maximal payoff, or minimal loss) is for the number of trials
-allocated to $A$ to grow slightly faster than an exponential function
-of the number of trials allocated to $B$.  Holland then showed that this
-result extends naturally to the {\emph multi-armed bandit} case.
-
-In an population undergoing adaptation, the building blocks making up
-individuals can be compared to the arms of a multi-armed bandit.  Each
+In a population undergoing adaptation, building blocks consitute the arms of a multi-armed bandit.  Each
 building block can be said to have a probability of ``payoff'' (i.e.,
@@
 
-The question of how to balance exploitation and exploration---how to
-optimally allocate trials to different arms based on their observed
-payoff---now becomes the question of how to optimally sample in the
-vast space of possible building blocks, based on their observed
-average fitness.  Of course adaptive evolution deals in populations of
-individuals, not building blocks.  There is no explicit mechanism for
-keeping statistics on building blocks' ``observed average fitness''.
-However, Holland's central idea here is that an approximation to
-optimal building-block sampling does indeed occur, as an emergent
-property of the population dynamics.
+% The question of how to balance exploitation and exploration---how to
+% optimally allocate trials to different arms based on their observed
+% payoff---now becomes the question of how to optimally sample in the
+% vast space of possible building blocks, based on their observed
+% average fitness.  Of course adaptive evolution deals in populations of
+% individuals, not building blocks.  There is no explicit mechanism for
+% keeping statistics on building blocks' ``observed average fitness''.
+% However, Holland's central idea here is that an approximation to
+% optimal building-block sampling does indeed occur, as an emergent
+% property of the population dynamics.
 
-To show this mathematically, Holland defined an idealization of
-adaptive evolution: an algorithm he called a {\emph reproductive
-  plan}.  A reproductive plan is a population-based stochastic process
+Holland argued that an approximation to
+optimal building-block sampling actually occurs, as an emergent
+property of the population dynamics defined by a
+% To show this mathematically, he defined an idealization of
+%adaptive evolution: an algorithm he called a {\emph reproductive
+% plan}.
+a population-based stochastic process
 operating on bit strings, involving the ``genetic'' operators of
  reproduction, fitness-based selection, crossover, mutation, and
-inversion.  Building blocks are formalized as {\emph
-  schemata}---patterns of bits within a string. (Here are two examples
-of schemas: ``all strings beginning with the bits 10'' or ``all
-strings that start with the pattern 1*1'', where * can be replaced by
-either 0 or 1).
+ inversion.  
+%Building blocks are formalized as {\emph
+%   schemata}---patterns of bits within a string. (Here are two examples
+% of schemas: ``all strings beginning with the bits 10'' or ``all
+% strings that start with the pattern 1*1'', where * can be replaced by
+% either 0 or 1).
 
-Holland's most important result was the following: He proved
-mathematically that his reproductive plan results---implicitly---in a
-near-optimal allocation of trials to schemata, thus optimizing the
-exploitation versus exploration balance.  Holland noted that the
-explicit act of assigning fitnesses to the individuals in a population
-was actually implicitly sampling a much larger collection of building
-blocks.  Holland termed this {\emph implicit parallelism}; the
-parallel leveraging of statistics from implicit sampling is a main
-strength of his population-based approaches to search.
+% Holland's most important result was the following: He proved
+% mathematically that his reproductive plan results---implicitly---in a
+% near-optimal allocation of trials to schemata, thus optimizing the
+% exploitation versus exploration balance.  Holland noted that the
+% explicit act of assigning fitnesses to the individuals in a population
+% was actually implicitly sampling a much larger collection of building
+% blocks.  Holland termed this {\emph implicit parallelism}; the
+% parallel leveraging of statistics from implicit sampling is a main
+% strength of his population-based approaches to search.
 
-It was this formal setting and resulting theorems that led to the
-invention of genetic algorithms, which featured stochastic
-population-based search and {\emph crossover} as a critical
-operation: Crossover recombined successful building blocks and
-thus allowed them to be tested in new contexts.  What made GAs unique
+It was this formalization
+% and resulting theorems that led to the
+that led to the invention of genetic algorithms, which featured stochastic
+population-based search and recombination as a critical
+%{\emph crossover} as a critical
+operation that allowed successful building blocks to be recombined and
+tested in new contexts.  What made GAs unique
 among other evolution-inspired algorithms at the time were the
-mathematical foundations described above, including bit-string
-representations, the focus on recombination as a central mechanism,
+mathematical foundations described above, the emphasis on populartions
+and recombination as a central mechanism,
 and a focus (at least in Holland's mind) on continual adaptation to
@@
 
-It is important to point out again that the framework Holland
-developed in \cite{Holland1975} was more general than the genetic algorithm;
-it was created in order to develop an interdisciplinary theory of
+The framework Holland developed in \cite{Holland1975} was more general
+than the genetic algorithm; its aim was an interdisciplinary theory of
 adaptation, one that would inform biology, say, as much as computer
-science \cite{Christiansen1998}.  The later, successful
-application of genetic algorithms to real-world optimization and
-learning tasks was, for Holland, just icing on the cake.
+science \cite{Christiansen1998}.  The later, successful application of
+genetic algorithms to real-world optimization and learning tasks was,
+for Holland, just icing on the cake.
 
-\section{Learning Via Credit Assignment and Rule Discovery}
+\section{Internal Models and Perpetual Novelty}
 
-While genetic algorithms are nowadays most often associated with models of
-biological evolution, in Holland's view the processes of adaptation he
-was studying manifested themselves in many different complex systems.
-One system of special interest to Holland was the human mind, and in
-particular, its capacity to learn via induction.
+%One system of special interest to Holland was the human mind, and in
+%particular, its capacity to learn via induction.
 
-In the 1970s and 80s Holland formulated and extensively explored a
-particular model of decision-making, action, and inductive learning,
-called the {\emph classifier system} (e.g.,
-\cite{Holland1977,Holland1986}.  In a classifier system, a population
-of ``if-then'' rules interacts with an environment, and over time is
-able to learn to improve its performance via both {\emph
-  credit-assignment} and {\emph rule-discovery} algorithms.
+Internal models are central to Holland's theory of adaptive
+systems.  He posits that all adaptive systems create and use internal
+models to prosper in continually
+changing environments---\emph{perpetual novelty}. Models can be tacit and learned
+  through evolutionary time, as in the case of bacteria swimming up a
+  chemical gradient, or explicit and learned over a single lifespan,
+  as in the case of cognitive systems that incorporate experience into
+  internal representations through learning.  In
+  Holland's view, the key activity of an adaptive agent involves
+  building and refining these data-driven models
+  of the environment.
+%using them to make predictions and take actions,
+%  and occasionally receiving positive or negative feedbacks
+%  (intermittent rewards).
 
-Like genetic algorithms, classifier systems can be viewed both as 
-models of adaptation and as artificial-intelligence (AI) methods.  At
-the time Holland was developing classifier systems, the field of AI
-was focused on expert systems, which typically did not learn on their
-own, a fundamental deficiency in Holland's view: ``[Expert] systems
-are brittle in the sense that they respond appropriately only in
-narrow domains, requiring substantial human intervention to compensate
-for even slight shifts in domain'' \cite{Holland1986}. In
-contrast, Holland proposed that ``{\emph induction} is the basic, and
-perhaps only, way of making large advances in this direction.''
+In his second book, \emph{Induction} \cite{Holland1989}, Holland and
+his co-authors tackled the question of how cognitive agents can
+possibly learn and profit from
+internal models, without supervision by combining
+environmental inputs and rewards with stored knowledge.   In their
+framework, a 
+model defines a set of equivalence relations over environmental states,
+together with a set of transition rules, which are learned over time
+based on environmental rewards (or punishments).  Models that
+form valid homomorphisms with the environment allow the system to make
+accurate predictions.  In Holland's conception, the equivalence
+classes are initially very general, say ``moving object'' and
+``stationary object,'' and over time, through experience and learning,
+ are specialized into more useful and precise subclasses, say
+``insect'' and ``nest.''   Over time, the adaptive system
+builds up a default hierarchy of rules covering general
+cases and refinements for specific classes.
+
+% For example, a general rule might
+% state that, ``If X is a bird, then X can fly'', whereas a more specific
+% one might state, ``If X is a bird with small wings and a large body,
+% then X cannot fly'' \cite{Holland1989}.  When faced with a bird in
+% the environment, both rules might compete to post their conclusion,
+% but the more specific one would be favored if both of its conditions are met.
+
+% Maybe this should be moved to section on Complexity
+Although the idea of default hierarchies was prevalent in 
+knowledge representation systems of the era, Holland made two key
+contributions.  The first was his emphasis on homomorphisms as a formal way to
+evaluate model validity, and  idea that dates back to Ross Ashby's
+\emph{An Introduction to Cybernetics} \cite{Ashby1956}. Holland's student,
+Bernard Ziegler developed this idea into a formal theory of computer modeling
+and simulation \cite{Ziegler1976}.  
+% %Today, even though computational modeling is a major
+% %application area of computing, essential to most natural sciences and
+% %engineering disciplines
+Even today, these early homomorphic theories of modeling
+stand as the most elegant approach we know of to characterize when
+a model is consistent with the environment and how an intelligent
+agent, human or artificial, can update the model to better reflect
+reality.
+
+Holland's second key contribution was describing a computational
+mechanism, the \emph{learning classifier
+  system}~\cite{Holland1977,Holland1986}, to illustrate how a cognitive system could
+iteratively build up a detailed and hierarchical model of its
+environment to enhance survival.  The key learning elements of this method, the
+bucket-brigade algorithm, combined with a genetic algorithm, presaged many of the ideas in modern reinforcement
+learning, including unsupervised learning, non-Markovian learning, AND
+ONE OTHER?.
+
+% While genetic algorithms are nowadays most often associated with models of
+% biological evolution, in Holland's view the processes of adaptation he
+% was studying manifested themselves in many different complex systems.
+
+% In the 1970s and 80s Holland formulated and extensively explored a
+% particular model of decision-making, action, and inductive learning,
+% called the {\emph classifier system} (e.g.,)
+% .  In a classifier system, a population
+% of ``if-then'' rules interacts with an environment, and over time is
+% able to learn to improve its performance via both {\emph
+%   credit-assignment} and {\emph rule-discovery} algorithms.
+
+% Like genetic algorithms, classifier systems can be viewed both as 
+% models of adaptation and as artificial-intelligence (AI) methods.  At
+% the time Holland was developing classifier systems, the field of AI
+% was focused on expert systems, which typically did not learn on their
+% own, a fundamental deficiency in Holland's view: ``[Expert] systems
+% are brittle in the sense that they respond appropriately only in
+% narrow domains, requiring substantial human intervention to compensate
+% for even slight shifts in domain'' \cite{Holland1986}. In
+% contrast, Holland proposed that ``{\emph induction} is the basic, and
+% perhaps only, way of making large advances in this direction.''
 
 Holland's inspiration for classifier systems came from several
-different disciplines.  Knowledge representation in the form of a
+different disciplines, including Hebbian
+learning, artificial intelligence, evolutionary biology, economics,
+psychology, control theory, and other fields (e.g.,
+\cite{Bellman1961,Samuel1959}).  Knowledge representation in the form of a
 population of ``if-then'' rules seemed like a good choice, not only
@@
 rate.'' \cite{Waldrop1993} (p. 182).  
+The if-then rules, when activated, compete to post their results on a
+shared ``message list,'' modeling the system's short-term memory
+(again inspired by Hebb's work and AI blackboard systems of the day).
+Unlike the AI systems, however new rules were generated automatically,
+in a trial-and-error fashion using a
+genetic algorithm.
 
-The if-then rules, when activated, compete to post their results on a
-shared ``message list'', modeling the system's short-term memory
-(again inspired by Hebb's work).  The competition was based on an
-bidding process, in which a rule's current strength affected its
-ability to bid highly.  Successful rules were gradually strengthened
-(and unsuccessful rules weakened) via a credit-assignment method
+Successful rules were
+strengthened over time if their predictions led to positive rewards
+from the environment 
+(and weakened otherwise).  
+through a credit-assignment method
 called the {\emph bucket-brigade} algorithm, in which rules gaining
-rewards from the environment or from other rules would pay back part
+rewards from the environment or from other rules transferred some
 of their gains to those earlier-firing ``stage-setting'' rules that
-set up the conditions for the eventual reward.  The idea for the
-bucket-brigade algorithm was inspired by the mechanisms of Hebbian
-learning and by early work on learning and optimization in economics,
-psychology, control theory, and other fields (e.g.,
-\cite{Bellman1961,Samuel1959}). \footnote{Of course similar
-  credit-assignment methods have been used extensively in
-  reinforcement learning, neural networks, and other areas of modern
-  machine learning, but Holland was among the first to propose such
-  methods.}
+set up the conditions for the eventual reward.   [ONE SENTENCE HERE
+ABOUT SAMUEL'S CHECKER PLAYING PROGRAM.]
+%, e.g., finding a target, or avoiding a predator,
 
-Finally, new rules for the population are discovered by a genetic
-algorithm, using the strength of rules as a fitness function.
+%Finally, new rules for the population are discovered by a genetic
+%algorithm, using the strength of rules as a fitness function.
+%Classifier systems are fairly complex architectures, with learning
+%occurring at multiple time scales.  
+Although Holland proposed classifier systems as an executable theory
+of inductive processes in cognition, other researchers took the
+computational system more seriously, applying it to areas as diverse
+as poker-playing \cite{Smith1980}, control of gas pipeline
+transmission \cite{Goldberg1983}, and modeling the stock market
+\cite{Palmer1994}.  See Ref.~\cite{Booker1989} for more details about
+practical applications of classifier systems.  Today, other
+reinforcement learning methods are more popular for real-world
+decision and control problems, but classifier systems can perhaps be
+thought of as an essential ``stage-setting'' method that enabled the
+development of later approaches.
 
-Classifier systems are fairly complex architectures, with learning
-occurring at multiple time scales.  Nevertheless, such systems have
-been usefully applied to areas as diverse as
-poker-playing \cite{Smith1980}, control of gas pipeline
-transmission \cite{Goldberg1983}, and modeling the stock
-market \cite{Palmer1994}.  Modern reinforcement learning methods have
-largely taken over as more useful approaches to real-world decision
-and control problems, but classifier systems can perhaps be thought of
-as an essential ``stage-setting'' method that enabled the development
-of later approaches.
+% \footnote{Of course similar
+%   credit-assignment methods have been used extensively in
+%   reinforcement learning, neural networks, and other areas of modern
+%   machine learning, but Holland was among the first to propose such
+%   methods.}
 
-John Holland's main interest in classifier systems was how the two
-learning mechanisms could work together to create subsets of rules
-that accurately model the environment, and thus are able to ``fire''
-at appropriate times.  An important part of such internal models is
-the notion of a {\emph default hierarchy}---a set of rules that
-together form a kind of hierarchical decision structure, going from
-most general to most specific.  For example, a general rule might
-state that, ``If X is a bird, then X can fly'', whereas a more specific
-one might state, ``If X is a bird with small wings and a large body,
-then X cannot fly'' \cite{Holland1989}.  When faced with a bird in
-the environment, both rules might compete to post their conclusion,
-but the more specific one would be favored if both of its conditions are met.
-
-Holland envisioned that, with a complex environment and enough time
-for learning, a classifier system might learn an intricate default
-hierarchy of rules, and that the competition inherent in the learning
+Holland was primarily interested in how the two
+learning mechanisms (discovery of new rules and apportioning credit to
+existing rules) could work together to create 
+%subsets of rules
+%that accurately model the environment, and thus are able to ``fire''
+%at appropriate times.  
+%Holland envisioned that, with a complex environment and enough time
+%for learning, a classifier system might learn an intricate default
+useful default hierarchies of rules.  He emphasized that the competition inherent in the learning
 and action mechanisms would allow the system to adapt to a continually
-changing environment without losing what it has learned in the past.
+changing environment without losing what it had learned in the past.
 Holland put it this way: ``Competition among rules provides the system
@@
 equivalent of saying that it has certain well-validated hypotheses.
-Offspring rules, which begin life weaker than do their parents, can
-win the competition and influence the system's behavior only when
-there are not strong rules whose conditions are satisfied---in other
-words when the system does not know what to do.  If their actions
-help, they survive; if not they are soon replaced.  Thus, the
-offspring do not interfere with the system's action in well-practiced
+%Offspring rules, which begin life weaker than do their parents, can
+%win the competition and influence the system's behavior only when
+%there are not strong rules whose conditions are satisfied---in other
+%words when the system does not know what to do.  If their actions
+%help, they survive; if not they are soon replaced.  
+%Thus,
+... new rules do not interfere with the system's action in well-practiced
 situations but wait gracefully in the wings as hypotheses about what
-to do under novel circumstances." \cite{Holland1992}.  A detailed
-discussion of some real-world applications of classifier systems is
-given in \cite{Booker1989}.
-
-\section{Internal Models and Default Hierarchies}
-
-The concept of {\emph internal models} is central to Holland's theory
-of adaptive systems.  He posits that all adaptive systems create and
-use internal models to prosper in their environments, through
-anticipation.  Models can be tacit and learned through evolutionary
-time, as in the case of bacteria swimming up a chemical gradient, or
-explicit and learned over a single lifespan, as in the case of
-cognitive systems that incorporate experience into internal rule sets
-through learning.
-
-In \cite{Holland1989}, Holland and his co-authors discuss how such
-internal models can be learned, without supervision, by combining
-environmental inputs with stored knowledge.  A key idea is that a
-model defines sets of equivalence relations over environmental states,
-together with a set of transition rules, which are learned over time
-based on environmental ``rewards'' (or ``punishments'').  Models that
-form valid homomorphisms with the environment allow the system to make
-accurate predictions.  In Holland's conception, the equivalence
-classes are initially very general, say ``moving object'' and
-``stationary object,'' and over time, through experience and learning,
-they are specialized into more useful and precise subclasses, say
-``insect'' and ``nest''.  As described above, the adaptive system
-eventually learns a default hierarchy of rules involving both general
-and specific classes.
-
-Although the idea of default hierarchies is prevalent in many
-knowledge representation systems (LIST A FEW?), Holland made two key
-contributions.  The first was his emphasis of homomorphisms as a formal way to
-evaluate model validity.  This idea that dates back to Ross Ashby's
-\emph{An Introduction to Cybernetics} \cite{Ashby1956}. Holland's student,
-Bernard Ziegler developed this idea into a formal theory of computer modeling
-and simulation \cite{Ziegler1976}.  
-%Today, even though computational modeling is a major
-%application area of computing, essential to most natural sciences and
-%engineering disciplines
-Even today, these early homomorphic theories of modeling
-stand as the most elegant approach we know of for characterizing when
-a model is consistent with the environment and how an intelligent
-agent, human or artificial, can update the model to better reflect
-reality.
-
-Holland's second key contribution was describing a computational
-mechanism---classifier systems---by which a cognitive system could
-iteratively build up a detailed and hierarchical model of its
-environment.  The key learning elements of this method, the
-bucket-brigade algorithm, combined with a genetic algorithm, as
-described above, presaged many of the ideas in modern reinforcement
-learning.
+to do under novel circumstances." \cite{Holland1992}.
 

### 2015-11-03 paper.tex
second try on commit

@@
 
-%\title{Adaptation Writ Large:\\ An Essay in Memory of JohnH. Holland}
+%\title{Adaptation Writ Large:\\ An Essay in Memory of John H. Holland}
+
 \title{Adaptative Computation:\\ An Essay in Memory of John H. Holland}
@@
 Professor John H. Holland passed away recently in Ann Arbor, MI, where
-he had been on the University of Michigan faculty for over 50 years.
-John, as he was known universally to his colleagues and students,
-leaves behind a long legacy of intellectual achievements.
+he had been on the University of Michigan faculty for over 50 years.  John, as he was known
+universally to his colleagues and students, leaves behind a long
+legacy of intellectual achievements.
 
-John was a direct intellectual descendant of the cybernetics era, and
-early on was strongly influenced by the work of von Neumann, Wiener,
-Ashby, and Turing, all of whom viewed computation as a broad,
-interdisciplinary enterprise.  Thus, John became an early proponent of
-interdisciplinary approaches to computer science and was an active
-evangelist of what is now called {\emph computational thinking}, reaching out
-enthusiastically to psychologists, economists, physicists, linguists,
-philosophers, and pretty much anyone he came in contact with.  As a
-result, even though he received what was arguably the world's first
-computer science Ph.D. in 1959, his contributions are sometimes better known
-outside of CS than within.
+As a descendant of the cybernetics era, Holland was strongly influenced
+by the work of von Neumann, Wiener, Ashby, and Turing, all of whom
+viewed computation as a broad, interdisciplinary enterprise.  Thus,
+Holland became an early proponent of interdisciplinary approaches to
+computer science and an active evangelist of what is now called
+\emph{computational thinking}, reaching out enthusiastically to
+psychologists, economists, physicists, linguists, philosophers, and
+pretty much anyone he came in contact with.  As a result, even though
+he received what was arguably the world's first computer science
+Ph.D. in 1959, his contributions are sometimes better known outside of
+computer science than within.
 
-John Holland is best known for his invention of {\emph genetic
-  algorithms} (GAs), a family of search and learning methods inspired
-by biological evolution.  Since their invention in the 1960s, GAs,
-along with related evolutionary computation methods, have become a
-thriving subfield of computer science, with widespread scientific and
-commercial applications.  While the mechanisms of GAs are well-known
-to much of the CS community, fewer are aware that GAs were only one
-offshoot of Holland's much broader motivation---to develop a general
-theory of adaptation in complex systems.  This motivation was the
-driving force in all of Holland's research.
+Holland is best known for his invention of \emph{genetic
+  algorithms} (GAs), a family of search and optimization methods inspired
+by biological evolution.  Since their invention in the 1960s, GAs have
+inspired many related methods and led to the thriving subfield of
+\emph{evolutionary computation}, with widespread scientific and
+commercial applications.  While the mechanisms of GAs are well-known,
+%to much of the computer science community, 
+the ideas that motivated Holland are less widely appreciated---to develop a general
+%Holland's motivation 
+%offshoot of Holland's much broader motivation--
+theory of adaptation in complex systems.  
+%This motivation was the
+%driving force in all of Holland's research.
 
-In this short essay, we sketch five key, recurrent themes of Holland's
-work on adaptive systems: (1) the evolutionary dynamics of ``building
-blocks''; (2) learning via credit assignment and rule discovery; (3)
-the emergence of internal models; (4) the art of scientific modeling;
-and (5) the universal properties of complex adaptive systems.  We
-discuss the role these themes have played in computer science, and
-highlight especially the ideas that we think remain relevant to
-today's research agendas.
+In this short essay, we consider this larger framework, sketching the
+recurring themes that were central to Holland's theory of
+adaptive systems: (1) discovery and 
+%evolutionary 
+dynamics of building blocks; (2) internal models and [prediction?] perpetual
+novelty; (3) the art [SF doesn't like the term ART] of scientific modeling; and (4) common
+properties of complex adaptive systems.  
+%We discuss the role these
+%themes have played in computer science, and highlight especially the
+%ideas that we think remain relevant to today's research agendas.  [MM
+%says: Are we going to do this?]
 
-\section{Evolutionary Dynamics of ``Building Blocks''}
+%\section{Evolutionary Dynamics of ``Building Blocks''}  
+\section{Discovery and Dynamics of Building Blocks}
 
-Holland's goal of developing a general theory of adaptation was
-spurred both by his early work on computer models of Hebbian learning
-\cite{Rochester1956} and his reading of Ronald Fisher's classic work
-on integrating genetics with Darwinian selection \cite{Fisher1930}.
-As Holland further read extensively in evolutionary biology,
-economics, game theory, and control theory, he came to recognize that
-adaptation was the most central concept in all these fields.  That is,
-these fields all concern populations of agents that must continually
-obtain information from uncertain changing environments and use it in
-order to improve their performance with respect to those environments.
-Moreover, Holland recognized that in systems with {\em adaptive
-  agents} of this kind, there is never a state of {\emph equilibrium}
-or a final \emph{optimum} configuration.  Due to the environment's
-perpetual novelty,
+Holland's interest in a general theory of adaptation grew out of
+his early work developing computer models of Hebbian learning
+\cite{Rochester1956} and 
+his reading of 
+Ronald Fisher's classic work that
+integrated genetics with Darwinian selection \cite{Fisher1930}.
+%As Holland further read extensively in 
+Over time, he recognized that adaptation was central to 
+evolutionary biology,
+economics, psychology, game theory, control theory, and many other fields.
+%he came to recognize that
+%adaptation was the most central concept in all these fields.  That is,
+Each field concerns populations of agents that must continually
+obtain information from uncertain changing environments and use it to
+improve their performance and enhance survival.
+%with respect to those environments.
+Moreover, Holland recognized that adaptation must be a continual
+open-ended process due to the
+environment's \emph{perpetual novelty}.
 %SF thinks Holland did not invent this term (as Holland termed it), 
-adaptation is a continual 
-open-ended process.  Holland's focus on open-ended,
-non-equilibrium dynamics was in stark contrast with the mainstream
-approach (at the time) in all these fields---the belief that ``solving
-for'' stable equilibrium dynamics was the scientific goal.  Holland's
-contrary view was that a system in stable equilibrium is essentially
-\emph{dead}.
+Thus, adaptive systems never achieve a state of equilibrium or a final
+optimum configuration.  This emphasis on open-ended, non-equilibrium
+dynamics was in stark contrast with the mainstream approach (at the
+time) in all these fields---the belief that solving for stable
+equilibrium dynamics was the scientific goal.  Holland's contrary view
+was that a system in stable equilibrium is essentially \emph{dead}.
 
@@
 
-Underlying Holland's theory of adaptation are three core ideas:
-Population-based search, building blocks and recombination, and the
-exploration-exploitation tradeoff.
+Underlying Holland's theory of adaptation are the following core ideas:
 
 \begin{itemize}
-\item{\bf Populations and implicit parallelism:} A population of
-  individuals can be thought of as a set of samples taken from a much larger
-  search space.  Importantly, the populations are competitive, and
-  the evolutionary dynamic biases the set of samples
-  over time towards high-fitness regions of the search space.
+\item{\bf Populations and implicit parallelism:} Adaptation occurs in
+  populations of individuals evolving (or learning) over time, in
+  which statistics can be leveraged to direct population dynamics.
+  Evolution can be thought of as a search for populations that are highly fit with respect to their environment.  
+  A population can be thought of as a parallel sampling of many
+  individuals (from the space of possible individuals).  Moreover,
+  that same population can be thought of as implicitly sampling a much
+  space of \emph{traits} that comprise those individuals.
+  Holland termed this implicit large-scale sampling of traits ``implicit
+  parallelism.''    Evolutionary dynamics biases these samples over time towards
+  high-fitness regions of the search space.
 
@@
 
-\item{\bf Building Blocks and recombination:} High-fitness individuals
-  are discovered in stages, first by finding high-fitness partial
-  solutions (the building blocks) through random sampling, and over
-  time assembling the partial solutions into more complete solutions
-  with even higher fitness.  For example, when searching the space of
-  all $n$-bit binary strings, almost any random sample would contain
-  examples of strings whose first bit is set to 1, and also many
-  samples of strings whose second bit is set to 1, but there would be
-  many fewer samples that have both of the first two bits set to
-  1.  More compelling examples include the \emph{Krebs cycle}, a core
-  cellular metabolic pathway, a group of neurons that responds to an
-  oriented edge in vision, or the ``swap'' operator in a sorting
-  algorithm.
+\item{\bf Building blocks and recombination:} 
+% Melanie rewrote this some to take out references to ``solutions''
+  In a population undergoing adaptation,
+  individuals can be decomposed into \emph{building blocks}---sets of
+  traits that are the evolutionary ``atoms'' of an individual's
+  fitness or performance. (As an example from biology, Holland gives
+  the \emph{Krebs cycle}, a core cellular metabolic
+  pathway.)   Successful individuals
+  are discovered in stages, first by finding useful
+  building blocks through random sampling, and over
+  time recombining such building blocks to create higher-fitness individuals.
+
+%  For example, when searching the space of
+%  all $n$-bit binary strings, almost any random sample would contain
+%  examples of strings whose first bit is set to 1, and also many
+%  samples of strings whose second bit is set to 1, but there would be
+%  many fewer samples that have both of the first two bits set to
+%  1.  More compelling examples include the \emph{Krebs cycle}, a core
+%  cellular metabolic pathway,
 
   To succeed, such a search requires first that
-  high-fitness subparts be preserved once found, and second, that
+  useful building blocks be preserved once discovered, and second that
   there be an efficient mechanism for combining promising building
@@
   over in genetics for the second.  Holland referred to the building
-  blocks as \emph{schemas,} which correspond mathematically to hyperplanes in the
-  search space.
+  blocks as \emph{schemas.} 
+%which correspond mathematically to hyperplanes in the search space. MM: this makes sense only in bit-string example.
 
 %In a population undergoing adaptation,
-%  individuals can be decomposed into {\emph building blocks}---sets of
+%  individuals can be decomposed into \emph{building blocks}---sets of
 % traits that are the evolutionary ``atoms'' of an individual's
@@
 \item{\bf Exploitation Versus Exploration:} Successful adaptation
-  requires maintaining a balance between {\emph exploitation}, in which
-  tried-and-true building blocks propagate in a population, and {\emph
+  requires maintaining a balance between \emph{exploitation}, in which
+ successful building blocks propagate in a population, and \emph{
     exploration}, in which existing building blocks are recombined or
-  mutated in new ways.  
+  mutated in new ways.  Inspired by Bellman \cite{Bellman1961} and
+others, Holland formalized this exploitation-versus-exploration tradeoff
+via an idealized ``two-armed bandit'' problem.  Given a slot machine
+with two arms, each of which has an unknown payoff probability, how
+should you allocate $N$ trials (pulls) between the arms so as to
+maximize your total payoff?   In \cite{Holland1973,Holland1975}
+Holland 
+%derived
+%an equation for such a strategy and 
+argued that the optimal strategy
+allocates trials to the observed best arm at a slightly higher rate than
+exponential.
+% and extended the result to multi-armed bandits.
 %A system that focuses too much on exploitation
@@
 %  between these two modes of processing.
+Next, he showed that 
+populations undergoing adaptation resemble multi-armed bandits, where
+each building block is an 
+%undergoing adaptation constitute the 
+arms in a multi-armed bandit problem.
+%Each building block can be said to have a probability of ``payoff''
+%(i.e., contribution to a given individual's fitness).
+Thus, in the process of assigning a fitness to each individual in a
+population, adaptive evolution can be seen as implicitly sampling the
+many building blocks making up that collection of individuals.  The
+average fitness over many individuals containing a particular building
+block gives an estimate of that building block's payoff probability.
+
 \end{itemize}
@@
 \cite{Holland1962,Holland1973}) and his influential
-1975 book {\emph Adaptation in Natural and Artificial Systems}
+1975 book \emph{Adaptation in Natural and Artificial Systems}
 \cite{Holland1975} developed a general, formal setting in which these ideas
 could be expressed mathematically.
-
-Inspired by Bellman \cite{Bellman1961} and
-others, Holland explored the exploitation-versus-exploration tradeoff
-via an idealized ``two-armed bandit'' problem.  Given a slot machine
-with two arms, each of which has an unknown payoff probability, how
-should you allocate $N$ trials (pulls) between the arms so as to
-maximize your total payoff?  An example of an extreme {\emph
-  exploitation} strategy would be to alternate between the arms until
-one of them gives a payoff, and then allocate all future trials to
-that arm alone.  Conversely, an example of an extreme {\emph
-  exploration} strategy would be to randomly allocate the trials,
-irrespective of the payoff rates you obtain.  Obviously, each of these
-strategies is flawed.  What is an optimal strategy?
-In \cite{Holland1973,Holland1975} Holland derived
-an equation for such a strategy and argued that the optimal strategy
-allocated trials to the observed best arm at a slightly higher rate than
-exponential.  
-% SF used 'argue' because some claim that his derivation was wrong.
-He then extended this 
-% Let $A$ denote the arm currently
-%observed to have the higher payoff probability, and $B$ denote the
-%other arm.  Holland showed that the optimal strategy (the one that
-%yields maximal payoff, or minimal loss) is for the number of trials
-%allocated to $A$ to grow slightly faster than an exponential function
-%of the number of trials allocated to $B$.  Holland then showed that this
-result to the {\emph multi-armed bandit} case.
-
-In a population undergoing adaptation, building blocks consitute the arms of a multi-armed bandit.  Each
-building block can be said to have a probability of ``payoff'' (i.e.,
-contribution to a given individual's fitness).  Evaluating an
-individual in an environment is like ``pulling the arms'' on a
-multi-armed bandit, where the ``arms'' correspond to each of the
-building blocks making up that individual.  Thus, in the process of
-assigning a fitness to each individual in a population, adaptive
-evolution can be seen as implicitly sampling the many building blocks
-making up that collection of individuals.  The average fitness over
-many individuals containing a particular building block gives an
-estimate of that building block's payoff probability.
-
-% The question of how to balance exploitation and exploration---how to
-% optimally allocate trials to different arms based on their observed
-% payoff---now becomes the question of how to optimally sample in the
-% vast space of possible building blocks, based on their observed
-% average fitness.  Of course adaptive evolution deals in populations of
-% individuals, not building blocks.  There is no explicit mechanism for
-% keeping statistics on building blocks' ``observed average fitness''.
-% However, Holland's central idea here is that an approximation to
-% optimal building-block sampling does indeed occur, as an emergent
-% property of the population dynamics.
-
-Holland argued that an approximation to
-optimal building-block sampling actually occurs, as an emergent
-property of the population dynamics defined by a
-% To show this mathematically, he defined an idealization of
-%adaptive evolution: an algorithm he called a {\emph reproductive
-% plan}.
-a population-based stochastic process
-operating on bit strings, involving the ``genetic'' operators of
- reproduction, fitness-based selection, crossover, mutation, and
- inversion.  
-%Building blocks are formalized as {\emph
-%   schemata}---patterns of bits within a string. (Here are two examples
-% of schemas: ``all strings beginning with the bits 10'' or ``all
-% strings that start with the pattern 1*1'', where * can be replaced by
-% either 0 or 1).
-
-% Holland's most important result was the following: He proved
-% mathematically that his reproductive plan results---implicitly---in a
-% near-optimal allocation of trials to schemata, thus optimizing the
-% exploitation versus exploration balance.  Holland noted that the
-% explicit act of assigning fitnesses to the individuals in a population
-% was actually implicitly sampling a much larger collection of building
-% blocks.  Holland termed this {\emph implicit parallelism}; the
-% parallel leveraging of statistics from implicit sampling is a main
-% strength of his population-based approaches to search.
 
@@
 that led to the invention of genetic algorithms, which featured stochastic
-population-based search and recombination as a critical
-%{\emph crossover} as a critical
+population-based search, as well as
+%recombination as a critical
+crossover between individuals as a critical
 operation that allowed successful building blocks to be recombined and
@@
 among other evolution-inspired algorithms at the time were the
-mathematical foundations described above, the emphasis on populartions
-and recombination as a central mechanism,
-and a focus (at least in Holland's mind) on continual adaptation to
+mathematical foundations described above, the emphasis on populations
+and recombination as central mechanisms,
+and a focus 
+%(at least in Holland's mind) [MM SAYS: THIS SOUNDS TOO NEGATIVE]
+on continual adaptation to
 non-stationary environments rather than optimization to static
@@
 
-\section{Internal Models and Perpetual Novelty}
+%\section{Internal Models and Perpetual Novelty}
+\section{Internal Models and Prediction}
 
@@
 systems.  He posits that all adaptive systems create and use internal
-models to prosper in continually
-changing environments---\emph{perpetual novelty}. Models can be tacit and learned
+models to prosper in the face of ``perpetual novelty'' in continually changing environments.   Models can be tacit and learned
   through evolutionary time, as in the case of bacteria swimming up a
@@
 his co-authors tackled the question of how cognitive agents can
-possibly learn and profit from
-internal models, without supervision by combining
+possibly learn and profit from internal models by combining
 environmental inputs and rewards with stored knowledge.  In their
-framework, a 
-model defines a set of equivalence relations over environmental states,
-together with a set of transition rules, which are learned over time
-based on environmental rewards (or punishments).  Models that
-form valid homomorphisms with the environment allow the system to make
-accurate predictions.  In Holland's conception, the equivalence
-classes are initially very general, say ``moving object'' and
-``stationary object,'' and over time, through experience and learning,
- are specialized into more useful and precise subclasses, say
-``insect'' and ``nest.''   Over time, the adaptive system
-builds up a default hierarchy of rules covering general
-cases and refinements for specific classes.
+framework, a model defines a set of equivalence relations over
+environmental states, together with a set of transition rules, which
+are learned over time based on environmental rewards (or punishments).
+Models that form valid homomorphisms with the environment allow the
+system to make accurate predictions.  In Holland's conception, the
+equivalence classes are initially very general, say ``moving object''
+and ``stationary object.'' Through experience and
+learning, these can be specialized into more useful and precise subclasses, say
+``insect'' and ``nest.''  Over time, the adaptive system builds up a
+default hierarchy of rules covering general cases and refinements for
+specific classes.
 
@@
 contributions.  The first was his emphasis on homomorphisms as a formal way to
-evaluate model validity, and  idea that dates back to Ross Ashby's
-\emph{An Introduction to Cybernetics} \cite{Ashby1956}. Holland's student,
+evaluate model validity, an idea that dates back to Ross Ashby's
+\emph{An Introduction to Cybernetics} \cite{Ashby1956}. Holland's student
 Bernard Ziegler developed this idea into a formal theory of computer modeling
@@
 environment to enhance survival.  The key learning elements of this method, the
-bucket-brigade algorithm, combined with a genetic algorithm, presaged many of the ideas in modern reinforcement
-learning, including unsupervised learning, non-Markovian learning, AND
-ONE OTHER?.
+bucket-brigade algorithm combined with a genetic algorithm, presaged many of the ideas in modern reinforcement
+learning.
+
+%including unsupervised learning, non-Markovian learning, AND
+%ONE OTHER?.
 
@@
 % particular model of decision-making, action, and inductive learning,
-% called the {\emph classifier system} (e.g.,)
+% called the \emph{classifier system} (e.g.,)
 % .  In a classifier system, a population
 % of ``if-then'' rules interacts with an environment, and over time is
-% able to learn to improve its performance via both {\emph
-%   credit-assignment} and {\emph rule-discovery} algorithms.
+% able to learn to improve its performance via both \emph{
+%   credit-assignment} and \emph{rule-discovery} algorithms.
 
@@
 % for even slight shifts in domain'' \cite{Holland1986}. In
-% contrast, Holland proposed that ``{\emph induction} is the basic, and
+% contrast, Holland proposed that ``\emph{induction} is the basic, and
 % perhaps only, way of making large advances in this direction.''
@@
 because of its popularity in AI at the time but also from Holland's
-early work on modeling Hebbian cell assemblies: ``In Hebbs view, a
+early work on modeling Hebbian cell assemblies: ``In Hebb's view, a
 cell assembly makes a simple statement: If such and such an event
@@
 (again inspired by Hebb's work and AI blackboard systems of the day).
-Unlike the AI systems, however new rules were generated automatically,
-in a trial-and-error fashion using a
-genetic algorithm.
+Unlike the AI systems, however, new rules were generated automatically
+%in a trial-and-error fashion
+using a genetic algorithm.
 
-Successful rules were
-strengthened over time if their predictions led to positive rewards
-from the environment 
-(and weakened otherwise).  
-through a credit-assignment method
-called the {\emph bucket-brigade} algorithm, in which rules gaining
-rewards from the environment or from other rules transferred some
-of their gains to those earlier-firing ``stage-setting'' rules that
-set up the conditions for the eventual reward.   [ONE SENTENCE HERE
-ABOUT SAMUEL'S CHECKER PLAYING PROGRAM.]
+Successful rules were strengthened over time if their predictions led
+to positive rewards from the environment (and weakened otherwise)
+through a credit-assignment method called the \emph{bucket-brigade}
+algorithm, in which rules gaining rewards from the environment or from
+other rules transferred some of their gains to those earlier-firing
+``stage-setting'' rules that set up the conditions for the eventual
+reward. Holland credited Arthur Samuel's pioneering work on machine
+learning applied to checkers \cite{Samuel1959} as a key inspiration for these ideas.
+
 %, e.g., finding a target, or avoiding a predator,
@@
 %occurring at multiple time scales.  
-Although Holland proposed classifier systems as an executable theory
-of inductive processes in cognition, other researchers took the
-computational system more seriously, applying it to areas as diverse
+Although Holland proposed classifier systems as 
+an executable theory of inductive processes in cognition, other researchers took it further, 
+%other researchers took the
+%computational system more seriously
+applying it to areas as diverse
 as poker-playing \cite{Smith1980}, control of gas pipeline
 transmission \cite{Goldberg1983}, and modeling the stock market
-\cite{Palmer1994}.  See Ref.~\cite{Booker1989} for more details about
-practical applications of classifier systems.  Today, other
+\cite{Palmer1994}.  (See Ref.~\cite{Booker1989} for more details about
+practical applications of classifier systems.)  Today, other
 reinforcement learning methods are more popular for real-world
 decision and control problems, but classifier systems can perhaps be
-thought of as an essential ``stage-setting'' method that enabled the
+thought of as an essential stage-setting method that enabled the
 development of later approaches.
@@
 %for learning, a classifier system might learn an intricate default
-useful default hierarchies of rules.  He emphasized that the competition inherent in the learning
-and action mechanisms would allow the system to adapt to a continually
-changing environment without losing what it had learned in the past.
-Holland put it this way: ``Competition among rules provides the system
-with a graceful way of handling perpetual novelty.  When a system has
-strong rules that respond to a particular situation, that is the
-equivalent of saying that it has certain well-validated hypotheses.
+useful default hierarchies of rules.  He emphasized that the
+competition inherent in the learning and action mechanisms would allow
+the system to adapt to a continually changing environment without
+losing what it had learned in the past.  Holland put it this way:
+``Competition among rules provides the system with a graceful way of
+handling perpetual novelty.  When a system has strong rules that
+respond to a particular situation, that is the equivalent of saying
+that it has certain well-validated hypotheses.... New rules do not
+interfere with the system's action in well-practiced situations but
+wait gracefully in the wings as hypotheses about what to do under
+novel circumstances.'' \cite{Holland1992}.
+
 %Offspring rules, which begin life weaker than do their parents, can
@@
 %Thus,
-... new rules do not interfere with the system's action in well-practiced
-situations but wait gracefully in the wings as hypotheses about what
-to do under novel circumstances." \cite{Holland1992}.
 
@@
 Given that Holland believed that the ability to learn and manipulate
-internal models was essential for any adapting system, it is no
+internal models was essential for any adaptive system, it is no
 surprise that he viewed modeling as essential for scientific inquiry.
@@
 well beyond data fitting methods, which typically focus only on
-describing the aggregate behavior of a system.
+the aggregate behavior of a system.
 
-For example, a statistical model that describes the boom and bust
-pattern of the stock market cannot address the underlying mechanisms
+For example, a purely statistical model that describes the boom and bust
+pattern of the stock market
+%cannot
+does not address the underlying mechanisms
 that lead to these cycles, through the collective actions of myriad
-individual buy/sell decisions.  Similarly, the genetic algorithms for
-which Holland is so famous provide a simple computational framework in
-which to explore the dynamics of Darwinian evolution and whether the
-basic mechanisms of variation, differential reproduction, and heredity
-are sufficient to account for the richness of our natural world.
-
-The emphasis on exploratory models to build intuitions was an
-important theme of Holland's work, and he often quoted Eddington's
-remark on the occasion of the first experimental test of Einstein's
-theory of relativity: "The contemplation in natural science of a wider
-domain than the actual leads to a far better understanding of the
-actual" \cite{Eddington1927}.
+individual buy/sell decisions.  
+%Similarly
+In contrast, the genetic algorithms for which Holland is so famous
+are exploratory models of \emph{mechanism}: they provide a simple
+computational framework in which to explore the dynamics of Darwinian
+evolution and whether the basic mechanisms of variation, differential
+reproduction, and heredity are sufficient to account for the richness
+of our natural world.
 
 Holland was interested in models that explored basic principles and
-mechanisms, even if they didn't make specific or detailed predictions.
-Such models can show generically how certain behaviors could be produced.
-Holland pioneered a style of modeling that has come to be known as
-'individual based' or 'agent based,' in which every component of a
-system is represented explicitly and has state, e.g., every trader in
-a stock market system or every cell in an immune system model.  In
-such models, each agent has its own behavior rules, which it can
-update (or learn) over time.  In order to capture the constraints of
-systems living under spatial constraints, these models are often
-defined over spatial structures, such as networks or simple grids.
+mechanisms, even if they did not make specific or detailed
+predictions.  Such models can show generically how certain behaviors
+could be produced.  Holland pioneered a style of modeling that has
+come to be known as ``individual-based'' or ``agent-based,'' in which
+every component of a system is represented explicitly---e.g., every
+trader in a stock market system or every cell in an immune system
+model---and has a dynamic internal state. In such models, each agent has
+its own behavior rules, which it can modify over time through learning.  In
+order to capture the behavior of systems under spatial
+constraints, these models are often defined over spatial structures,
+such as networks or simple grids.
 
 A given agent-based model encodes a theory about the mechanisms that
-are relevant for producing the behavior of interest.  Similar to
-expert systems, such models are especially useful for studying systems
-that don't have analytic mathematical descriptions.  Agent-based
-models can facilitate interdisciplinary collaborations because the
-underlying rules can be easily communicated.  The agent-based models
+are relevant for producing the behavior of interest.  
+%Similar to expert systems [MM says: ?]
+%Such models are especially useful for studying systems
+%that do not have analytic mathematical descriptions.  [MM took out to shorten this paragraph]
+%Agent-based models can facilitate interdisciplinary collaborations because the
+%underlying rules can be easily communicated.  
+[SF: THINKS this next text doesn't make sense until after complex
+systems seciton.]
+The agent-based models
 championed by Holland were typically idealized versions of complex
@@
 
+The emphasis on exploratory models to build intuitions was an
+important theme of Holland's work, and he often quoted Eddington's
+remark on the occasion of the first experimental test of Einstein's
+theory of relativity: ``The contemplation in natural science of a wider
+domain than the actual leads to a far better understanding of the
+actual'' \cite{Eddington1927}.
+
 It should be noted that Holland's view of modeling is by no means
-typical.  For example, in a textbook on computational modeling, the
-authors offer the following definition: “Modeling is the application
+typical.  For example, a textbook on computational modeling offers
+ the following definition: ``Modeling is the application
 of methods to analyze complex, real-world problems in order to make
-predictions about what might happen with various actions”
-\cite{Shiflet2014}.   This sort of perspective completely rules out the
+predictions about what might happen with various actions''
+\cite{Shiflet2014}.   This perspective rules out the
 kind of exploratory modeling that Holland was most interested in.
 
-Some researchers dispute that models make any kind of scientific
-contribution: "Models are metaphors that explain the world we don't
-understand in terms of worlds we do.  They are merely analogies,
-provide partial insight, stand on someone else's feet.  Theories stand
-on their own feet, and rely on no analogies."  [Emanuel Derman, 2012].
-[STEPH, I COULD NOT FIND REFERENCE FOR THIS QUOTE OUTSIDE SFI VIDEO.]
+% Some researchers dispute that models make any kind of scientific
+% contribution: ``Models are metaphors that explain the world we don't
+% understand in terms of worlds we do.  They are merely analogies,
+% provide partial insight, stand on someone else's feet.  Theories stand
+% on their own feet, and rely on no analogies.''  [Emanuel Derman, 2012].
+% [MM SAYS:  I COULD NOT FIND REFERENCE FOR THIS QUOTE OUTSIDE SFI VIDEO.]
+% [MM SAYS:  Possibly delete this paragraph.]
 
-[NOW WE NEED A STRONG FINISHING SENTENCE TO RESCUE JHH STYLE
-  MODELING.]
+% [NOW WE NEED A STRONG FINISHING SENTENCE TO RESCUE JHH STYLE
+%   MODELING.]
+
+%\section{Common Properties of Complex Adaptive Systems}
 
@@
 
-Holland was interested in a broad array of adaptive systems---immune systems, ecologies, financial markets, cities, and the brain---systems that are \emph{complex}.
-In the 1980's, he teamed up with a small group of scientists, primarily physicists with a sprinkling of economists and biologists, to discuss what this wide swath of systems has in common.
-The discussions helped define the intellectual mission of the Santa Fe Institute, the first institution dedicated to developing a science of complexity.
-[add in how many complexity institutes there are worldwide and which others John played a hand in.]  Holland brought to these discussions his lifelong study of adaptation and a reminder that true theories about complexity would need to look deeper than phenomenological descriptions but also account for the  'how' and 'why' of these systems.
+Holland was interested in a broad array of adaptive systems---immune
+systems, ecologies, financial markets, cities, and the brain---systems
+that are \emph{complex}.  In the 1980s, he teamed up with a small
+group of scientists, primarily physicists with a sprinkling of
+economists and biologists, to discuss what properties this wide swath of systems
+have in common.  The discussions helped define the intellectual mission
+of the Santa Fe Institute (SFI), the first institution dedicated to
+developing a science of complexity.  [add in how many complexity
+  institutes there are worldwide and which others John played a hand
+  in. MM says: Do we really need that?]  Holland brought to these discussions his lifelong study of
+adaptation and a reminder that true theories about complexity would
+need to look deeper than phenomenological descriptions but also
+account for the `how' and `why' of these systems.
 %describing phenomena like chaos, power laws, and ONE MORE, and acoount for the
-As the discussions matured, a consensus developed about the basic elements of complexity---systems that: (1)  are composed of many components with nontrivial (nonlinear) interactions; (2) behavior that emerges from the dynamics of the system, exhibiting higher-order patterns; (3) Scale, systems are nested and structure/behavior emerges at different scales, with some behavior being conserved across all scales and other behaviors changing at different scales, and (4) continual adaptation where systems adjust their behavioral rules through evolution and learning.
-Although this is far from a formal definition of complex systems, most people working in the field today are interested in systems that have these properties.
 
-To illustrate his ideas about the ubuiquity of adaptation, Holland developed two systems, ECHO and the SFI stock market.
-Although less famous than genetic algorithms and learning classifier systems, ECHO was no less ambitious.
-It's scope included the evolution of interacting populations of individual agents, which through competition and learning, discover symbiotic triads such as the famous ant-fly-caterpillar interaction,  the Wiksell triangle of trading relationships in economics, and the immune system's learned ability to distinguish 'self' from 'other.'  Without going into the details of the model [CITES HERE], ECHO formalized Holland's theories about complex adapting systems into a runnable computational system where agents evolved external markers (called tags) and internal preferences and used them to form higher level aggregate structures (trading relationships, symbiotic groups, trophic cascades, interdependent organizations etc.).
-ECHO agents sometimes discovered mimicry to deceive competitors, and over time, the model developed increasingly complex structures and behaviors, often reproducing patterns observed in nature.
-For example,  ECHO evolved a diversity of agent types, whose rank-frequency distribution closely parallels the well-known Preston curve in ecology, a quantiative statement of the adage that 'most species are rare'.
-Many of the insights behind this project are described in his book, [WHICH ONES].
-The broad scope of this project was appealing to immunologists, economists, and evolutionary biologists alike, but it was the economists who took it one step further in the SFI stock market.
+As the discussions matured, a consensus developed about the basic
+elements of complexity: (1) complex systems are composed of many
+components with nonlinear interactions; (2) such systems
+are characterized by complex \emph{emergent} behavior, exhibiting
+higher-order patterns; (3) such systems operate at multiple (and often
+nested) spatial and temporal scales, with some behavior being
+conserved across all scales and other behaviors changing at different
+scales, and (4) these systems exhibit continual adaptation, adjusting
+their behavioral rules through evolution and learning.  Although this
+is far from a formal definition of complex systems, most people
+working in the field today are interested in systems that have these
+properties.
 
-% Note: SF has the history backwards here.  We should correct that, but the story was so nice in this order
+To illustrate his ideas about the ubiquity of adaptation, Holland and
+co-workers developed two exploratory models, the SFI Artificial Stock
+Market and ECHO.
 
-In the early 1990s, Holland teamed up with other SFI researchers, including two economists, to tackle the mismatch between predictions of the dominant theory in economics, the rational expectations hypothesis, and empirically observed stock market behaviors that deviated from the predicted equilibria solutions.
-In brief, most economic theory of the day assumed that all participants in an economy or financial market are 100\% rational and act to maximize their individual gain.
-We know, however, that real actors in real economies and markets are rarely rational, and financial markets often deviate from rationalilty, for example with speculative bubbles and crashes.
-The SFI stock market project studied what happens in a model where rational traders are replaced by traders who learn to forecast stock prices over time.
-The model was constructed to allow testing for the possible emergence of fundamental trading versus technical trading versus uninformed trading.
-The simulated market with (learning) trading agents was run many times and the dynamics of price and trading volumes were compared to observed patterns in real markets, replicating many of the features of real-life markets.
-Although the model was primitive in many ways, it was wildly influential and led to many follow-on projects  It also demonstrated the important role that adaptation plays in complex systems and illustrated how Holland's theories of continual learning in response to intermittent feedbacks from the environment could be integrated into domain-specific settings.
+In the early 1990s, Holland teamed up with other Santa Fe Institute researchers,
+including several economists, to tackle the mismatch between predictions
+of rational expectations theory (the then dominant theory in economics) 
+and empirically observed stock market behaviors 
+%that deviated from the predicted equilibrium solutions.  
+In brief, most
+economic theory of the day assumed that all participants in an economy
+or financial market are 100\% rational and act to maximize their
+individual gain.  In real life, however, actors in economies
+and markets are rarely wholly rational, and financial markets often deviate
+from rationality, for example with speculative bubbles and crashes.
+
+The SFI Artificial Stock Market project \cite{Arthur1997,Palmer1994} explored a model in which rational
+traders are replaced by \emph{adaptive} traders---those who learn to
+forecast stock prices over time.  The model allowed testing for the
+possible emergence of different trading strategies, including
+\emph{fundamental}, \emph{technical}, or \emph{uninformed}
+strategies. The simulated market with adaptive trading agents was run
+many times, and the dynamics of price and trading volumes were
+compared to observed patterns in real markets.  Holland and his
+collaborators found that the model's dynamics replicated several
+features of real-life markets.
+
+Although the SFI Stock Market model was highly simplified, it was very
+influential and led to many follow-on projects.  It was a clear
+demonstration of the essential role that adaptation plays in complex
+systems, and illustrated how Holland's theories of continual learning
+in response to intermittent feedbacks from the environment could be
+integrated into domain-specific settings.
+
+ECHO \cite{Holland1999,Hraber1997} was an even more ambitious model
+that Holland and collaborators developed during the 1990s.  ECHO's
+scope included the evolution of interacting populations of individual
+agents, which through competition and learning discover symbiotic
+triads such as the famous ant-fly-caterpillar interaction, the Wiksell
+triangle of trading relationships in economics, and the immune
+system's learned ability to distinguish `self' from `other.'
+
+ECHO formalized Holland's theories about complex adaptive systems into
+a runnable computational system where agents evolved external markers
+(called tags) and internal preferences, and used them to form higher
+level aggregate structures (trading relationships, symbiotic groups,
+trophic cascades, interdependent organizations, etc.).  ECHO agents
+sometimes discovered mimicry to deceive competitors, and over time,
+the model developed increasingly complex structures and behaviors,
+often reproducing patterns observed in nature.  For example, ECHO
+evolved a diversity of agent types whose rank-frequency distribution
+closely parallels the well-known Preston curve in ecology, a
+quantitative statement of the adage that ``most species are rare''. Many
+of the insights behind this project are described in Holland's book \emph{Hidden Order} \cite{Holland1995}. 
+The broad scope of this project was appealing to immunologists, economists, and evolutionary biologists alike.
+
+% but it
+%was the economists who took it one step further in the SFI stock
+%market.
 
@@
 
-Holland's later books,  \emph{Emergence}, \emph{Signals and Boundaries}, and \emph{Complexity, A Summary} show how the theories of adaptation that he developed during the earlier part of his career fit into the larger landscape of complex systems research.
-Holland's focus on how complex patterns emerge and change, rather than simply characterizing the patterns themselves (e.g., describing chaotic attractors or power laws) reflected his deterimination to 'get to the heart' of complex systems.
-This determination
-represents the best of science.  Holland's willingness to tackle the hardest questions, develop his own formalisms, and use mathematics productively sets a high bar that we all should aspire to.
+Holland's later books, \emph{Emergence} \cite{Holland2000},
+\emph{Signals and Boundaries} \cite{Holland2012}, and
+\emph{Complexity: A Very Short Introduction} \cite{Holland2014} show how the
+theories of adaptation that Holland developed during the earlier part
+of his career fit into the larger landscape of complex systems
+research.  Holland's focus on understanding the mechanisms by which complex patterns emerge and change,
+rather than simply characterizing the patterns themselves (e.g.,
+describing chaotic attractors or power laws), reflected his
+determination to get to the heart of complex adaptive systems.
+This determination represents the best of science.  Holland's
+willingness to tackle the hardest questions, develop his own
+formalisms, and use mathematics productively sets a high bar to whch we
+all should aspire.
 
 \section{Relevance to Modern CS}
+[MM says:  Do we need this section?  Or can we incorporate these points within the previous sections?]
 

### 2015-11-04 paper.tex
a wimpy conclusion

@@
 %evolutionary 
-dynamics of building blocks; (2) internal models and [prediction?] perpetual
-novelty; (3) the art [SF doesn't like the term ART] of scientific modeling; and (4) common
-properties of complex adaptive systems.  
+dynamics of building blocks; (2) internal models and prediction; (3) 
+exploratory modeling; and (4) complex adaptive systems.  
 %We discuss the role these
@@
   over in genetics for the second.  Holland referred to the building
-  blocks as \emph{schemas.} 
-%which correspond mathematically to hyperplanes in the search space. MM: this makes sense only in bit-string example.
+  blocks as \emph{schemas} and defined them as hyperplanes in the search space.
 
@@
 
-\section{The Art of Scientific Modeling}
+\section{Exploratory Modeling}
 
@@
 %underlying rules can be easily communicated.  
-[SF: THINKS this next text doesn't make sense until after complex
-systems seciton.]
-The agent-based models
-championed by Holland were typically idealized versions of complex
-systems and not intended to provide detailed, domain-specific
+%[SF: THINKS this next text doesn't make sense until after complex
+%systems seciton.]
+The exploratory models
+championed by Holland were idealized and not intended to provide detailed, domain-specific
 predictions.  Instead they were meant to explore possible general
@@
 
-\section{Complexity}
+\section{Complex Adaptive Systems}
 
@@
 of the Santa Fe Institute (SFI), the first institution dedicated to
-developing a science of complexity.  [add in how many complexity
-  institutes there are worldwide and which others John played a hand
-  in. MM says: Do we really need that?]  Holland brought to these discussions his lifelong study of
-adaptation and a reminder that true theories about complexity would
+developing a science of complexity, and the many others that followed.
+% [add in how many complexity
+%  institutes there are worldwide and which others John played a hand
+%  in. MM says: Do we really need that?]  
+Holland brought to these discussions his lifelong study of
+adaptation and a reminder that serious theories about complexity would
 need to look deeper than phenomenological descriptions but also
@@
 As the discussions matured, a consensus developed about the basic
-elements of complexity: (1) complex systems are composed of many
-components with nonlinear interactions; (2) such systems
-are characterized by complex \emph{emergent} behavior, exhibiting
-higher-order patterns; (3) such systems operate at multiple (and often
+elements of complex systems, which are: (1) 
+%complex systems are 
+composed of many
+components with nonlinear interactions; (2) 
+%such systems
+%are 
+characterized by complex \emph{emergent} behavior, exhibiting
+higher-order patterns; (3) 
+%such systems 
+operating at multiple (and often
+% interacting? 
 nested) spatial and temporal scales, with some behavior being
 conserved across all scales and other behaviors changing at different
-scales, and (4) these systems exhibit continual adaptation, adjusting
-their behavioral rules through evolution and learning.  Although this
+scales, and (4) 
+%these systems exhibit 
+adaptative, with 
+behavioral rules continually adjusting through evolution and learning.  Although this
 is far from a formal definition of complex systems, most people
@@
 
-To illustrate his ideas about the ubiquity of adaptation, Holland and
-co-workers developed two exploratory models, the SFI Artificial Stock
-Market and ECHO.
+%To illustrate his ideas about the ubiquity of adaptation, Holland and
+%co-workers developed two exploratory models, the SFI Artificial Stock
+%Market and ECHO.
 
@@
 of rational expectations theory (the then dominant theory in economics) 
-and empirically observed stock market behaviors 
+and empirically observed stock market behaviors. 
 %that deviated from the predicted equilibrium solutions.  
@@
 
-The SFI Artificial Stock Market project \cite{Arthur1997,Palmer1994} explored a model in which rational
-traders are replaced by \emph{adaptive} traders---those who learn to
-forecast stock prices over time.  The model allowed testing for the
-possible emergence of different trading strategies, including
+The SFI Artificial Stock Market project \cite{Arthur1997,Palmer1994}
+developed an exploratory model in which rational
+traders were replaced by \emph{adaptive} traders---those who learn to
+forecast stock prices over time.  The model tested for the
+emergence of different trading strategies, including
 \emph{fundamental}, \emph{technical}, or \emph{uninformed}
@@
 Although the SFI Stock Market model was highly simplified, it was very
-influential and led to many follow-on projects.  It was a clear
-demonstration of the essential role that adaptation plays in complex
-systems, and illustrated how Holland's theories of continual learning
+influential and led to many follow-on projects.  It 
+demonstrated clearly the essential role that adaptation plays in complex
+systems, and it illustrated how Holland's theories of continual learning
 in response to intermittent feedbacks from the environment could be
@@
 ECHO \cite{Holland1999,Hraber1997} was an even more ambitious model
-that Holland and collaborators developed during the 1990s.  ECHO's
-scope included the evolution of interacting populations of individual
-agents, which through competition and learning discover symbiotic
-triads such as the famous ant-fly-caterpillar interaction, the Wiksell
-triangle of trading relationships in economics, and the immune
-system's learned ability to distinguish `self' from `other.'
-
-ECHO formalized Holland's theories about complex adaptive systems into
+that Holland and collaborators developed during the 1990s. 
+% ECHO's
+% scope included the evolution of interacting populations of individual
+% agents, which through competition and learning discover symbiotic
+% triads such as the famous ant-fly-caterpillar interaction, the Wiksell
+% triangle of trading relationships in economics, and the immune
+% system's learned ability to distinguish `self' from `other.'
+ECHO formalized Holland's idealization of complex adaptive systems into
 a runnable computational system where agents evolved external markers
-(called tags) and internal preferences, and used them to form higher
-level aggregate structures (trading relationships, symbiotic groups,
+(called tags) and internal preferences, and then used them to acquire
+resources, form higher
+level aggregate structures (such trading relationships, symbiotic groups,
 trophic cascades, interdependent organizations, etc.).  ECHO agents
 sometimes discovered mimicry to deceive competitors, and over time,
-the model developed increasingly complex structures and behaviors,
-often reproducing patterns observed in nature.  For example, ECHO
-evolved a diversity of agent types whose rank-frequency distribution
-closely parallels the well-known Preston curve in ecology, a
-quantitative statement of the adage that ``most species are rare''. Many
-of the insights behind this project are described in Holland's book \emph{Hidden Order} \cite{Holland1995}. 
-The broad scope of this project was appealing to immunologists, economists, and evolutionary biologists alike.
+the model developed increasingly complex structures and behaviors, at
+least up to a point.  ECHO reproduced several well-known patterns 
+observed in nature, for example, over time the population of ECHO
+evolved agents exhibited diversity of agent types, diversity with a 
+rank-frequency distribution that
+closely paralleled the well-known Preston curve in ecology.  In
+layman's terms, ECHO evolved populations where 
+%quantitative statement of the adage that
+``most species are rare.''  
+The broad scope of the model, together with its ability to produce
+easily identifiable patterns, was appealing to immunologists, economists, and evolutionary biologists alike.   Many
+of the insights behind this project are described in Holland's third book \emph{Hidden Order} \cite{Holland1995}.
 
@@
 
-\section{Relevance to Modern CS}
-[MM says:  Do we need this section?  Or can we incorporate these points within the previous sections?]
-
-\begin{itemize}
-\item Evolutionary computation
-\item Q Learning
-\item Backprop
-\item    From Turing nomination: "Holland's machine learning system known as the
-   Learning Classifier System (LCS), developed in the early 1980s,
-   incorporated a reinforcement learning algorithm known as the bucket
-   brigade for non-Markovian environments, anticipating by nearly a
-   decade non-Markovian learning algorithms."
-\item Exploitation versus exploration -- relevance to reinforcement
-  learning, other parts of machine learning and optimization.
-\item Active learning.  (Two-armed bandit problem.) 
-\item On-line learning.  
-\end{itemize}
+%\section{Relevance to Modern CS}
+%[MM says:  Do we need this section?  Or can we incorporate these points within the previous sections?]
 
@@
 
-Introduced a few generations of students to computation in natural systems, an idea that today is better accepted.
-His insights were deeper and more general than what often passes for work in biomimicry, e.g., for robots.
+Holland was unusual in his ability to absorb the essence of other
+disciplines, articulate grand overarching
+theories, and them back them up with computational mechanisms and
+mathematics.  Unlike most researchers he moved seamlessly among these
+three modes of thinking, developing models that were years ahead of their time.  A
+close reading of his work reveals the antecedents of many ideas prevalent in
+machine learning today, e.g., reinforcement, learning in non-Markovian
+environments, active learning, etc.   His seminal genetic algorithm
+spawned the field of evolutionary computation, and his insights
+and wisdom helped define what are now referred to as the
+\emph{sciences of complexity}.   With such a broad scope, it is easy
+to pick at the details.
 
-The ideas have had huge impact and should still be a beacon for research in intelligent and complex systems
+His many books and articles have influenced 
+scientists around the work and across many disciplines.  Closer to home, he introduced several
+generations of students at the University of Michigan to 
+computation in natural systems, an idea that even today
+remains somewhat controversial, despite successes in biomimicry
+for robotics, abstractions of phermone trails in ant colonies for
+optimization methods, the use of mechanisms from immunology to
+improve computer security, and recent work on technological
+evolution.
 
-John's personality and humanity is inextricably tangled up with his intellectual contributions.
+Behind the ideas, of course, is the man himself.   Everyone who knew
+John personally will remember the gleam in his eye when he
+encountered a new idea; his willingness to talk to anyone, no matter
+how famous or how unfamous; and his incredible generosity and patience.
+%His insights aboutwere deeper and more general than what often passes for work in biomimicry, e.g., for robots.
+Holland's personality and humanity were somehow inextricably tangled
+up with his intellectual contributions, and although his ideas will serve as a beacon
+for research in intelligent and complex systems
+for many years to come, the world of science is poorer without John
+Holland on the scene.
+
+% \begin{itemize}
+% \item Evolutionary computation
+% \item Q Learning
+% \item Backprop
+% \item    From Turing nomination: "Holland's machine learning system known as the
+%    Learning Classifier System (LCS), developed in the early 1980s,
+%    incorporated a reinforcement learning algorithm known as the bucket
+%    brigade for non-Markovian environments, anticipating by nearly a
+%    decade non-Markovian learning algorithms."
+% \item Exploitation versus exploration -- relevance to reinforcement
+%   learning, other parts of machine learning and optimization.
+% \item Active learning.  (Two-armed bandit problem.) 
+% \item On-line learning.  
+% \end{itemize}
 

### 2015-11-18 paper.tex
version sent out to friends + 2 Jana typos corrected

@@
 Professor John H. Holland passed away recently in Ann Arbor, MI, where
-he had been on the University of Michigan faculty for over 50 years.  John, as he was known
-universally to his colleagues and students, leaves behind a long
-legacy of intellectual achievements.
+he had served on the University of Michigan faculty for over 50 years.
+John, as he was known universally to his colleagues and students,
+leaves behind a long legacy of intellectual achievements.
 
-As a descendant of the cybernetics era, Holland was strongly influenced
-by the work of von Neumann, Wiener, Ashby, and Turing, all of whom
-viewed computation as a broad, interdisciplinary enterprise.  Thus,
-Holland became an early proponent of interdisciplinary approaches to
-computer science and an active evangelist of what is now called
-\emph{computational thinking}, reaching out enthusiastically to
+As a descendant of the cybernetics era, Holland was strongly
+influenced by the work of von Neumann, Wiener, Ashby, and Turing, all
+of whom viewed computation as a broad, interdisciplinary enterprise.
+Thus, Holland became an early proponent of interdisciplinary
+approaches to computer science and an active evangelist of what is now
+called \emph{computational thinking}, reaching out enthusiastically to
 psychologists, economists, physicists, linguists, philosophers, and
@@
 commercial applications.  While the mechanisms of GAs are well-known,
+%Holland's broader motivations are less widely appreciated---to develop a 
+%general theory of adaptation in complex systems.  
+they were only one
+offshoot of Holland's broader motivation---to develop a general
+theory of adaptation in complex systems.  
+%This motivation was the
+%driving force in all of Holland's research.   
 %to much of the computer science community, 
-the ideas that motivated Holland are less widely appreciated---to develop a general
+%the ideas that motivated Holland are less widely appreciated---to develop a general
 %Holland's motivation 
 %offshoot of Holland's much broader motivation--
-theory of adaptation in complex systems.  
 %This motivation was the
@@
 adaptive systems: (1) discovery and 
-%evolutionary 
-dynamics of building blocks; (2) internal models and prediction; (3) 
-exploratory modeling; and (4) complex adaptive systems.  
+dynamics in adaptive search; (2) internal models and prediction; (3) 
+exploratory modeling; and (4) universal properties of complex adaptive systems.  
 %We discuss the role these
@@
 
-%\section{Evolutionary Dynamics of ``Building Blocks''}  
-\section{Discovery and Dynamics of Building Blocks}
+\section{Discovery and Dynamics in Adaptive Search}
 
-Holland's interest in a general theory of adaptation grew out of
-his early work developing computer models of Hebbian learning
-\cite{Rochester1956} and 
-his reading of 
-Ronald Fisher's classic work that
-integrated genetics with Darwinian selection \cite{Fisher1930}.
-%As Holland further read extensively in 
-Over time, he recognized that adaptation was central to 
-evolutionary biology,
-economics, psychology, game theory, control theory, and many other fields.
-%he came to recognize that
-%adaptation was the most central concept in all these fields.  That is,
-Each field concerns populations of agents that must continually
-obtain information from uncertain changing environments and use it to
-improve their performance and enhance survival.
-%with respect to those environments.
+Holland's goal of developing a general theory of adaptation 
+%spurred
+began with his early work on computer models of Hebbian learning
+\cite{Rochester1956} and his reading of Ronald Fisher's classic book
+that integrated genetics with Darwinian selection \cite{Fisher1930}.
+As Holland read extensively in evolutionary biology,
+economics, game theory, and control theory, he came to recognize that
+adaptation was the most central concept in all these fields.  That is,
+these fields all concern populations of agents that must continually
+obtain information from uncertain changing environments and use it in
+order to improve performance and enhance the chance of survival.
+
 Moreover, Holland recognized that adaptation must be a continual
-open-ended process due to the
-environment's \emph{perpetual novelty}.
-%SF thinks Holland did not invent this term (as Holland termed it), 
-Thus, adaptive systems never achieve a state of equilibrium or a final
-optimum configuration.  This emphasis on open-ended, non-equilibrium
-dynamics was in stark contrast with the mainstream approach (at the
-time) in all these fields---the belief that solving for stable
-equilibrium dynamics was the scientific goal.  Holland's contrary view
-was that a system in stable equilibrium is essentially \emph{dead}.
+open-ended process due to perpetually uncertain and changing
+environments.  Thus, adaptive systems never achieve a state of
+equilibrium or a final optimum configuration.  This emphasis on
+open-ended, non-equilibrium dynamics was in stark contrast with the
+mainstream approach (at the time) in all these fields---the belief
+that solving for stable equilibrium dynamics was the scientific goal.
+Holland's contrary view was that a system in stable equilibrium is
+essentially \emph{dead}.
 
-% Text that STEPH found in our NIPS paper on the IGA
-% First, at least theoretically the GA is fast because of implicit
-% parallelism (Holland, 1975/1992): each string in the population is an
-% instance of many different schemas, and if the population is large
-% enough and is initially chosen at random, a large number of different
-% schemas-many more than the number of strings in the population-are
-% being sampled in parallel.  This should result in a quick search for
-% short, low-order schemas that confer high fitness.  Second,
-% fitness-proportionate reproduction under the GA should conserve
-% instances of such schemas.  Third, a high crossover rate should
-% quickly combine instances oflow-order schemas on different strings to
-% create instances of longer schemas that confer even higher fitness.
-
-Underlying Holland's theory of adaptation are the following core ideas:
+Underlying Holland's theory of adaptation are the following core ideas.
 
 \begin{itemize}
-\item{\bf Populations and implicit parallelism:} Adaptation occurs in
-  populations of individuals evolving (or learning) over time, in
-  which statistics can be leveraged to direct population dynamics.
-  Evolution can be thought of as a search for populations that are highly fit with respect to their environment.  
-  A population can be thought of as a parallel sampling of many
-  individuals (from the space of possible individuals).  Moreover,
-  that same population can be thought of as implicitly sampling a much
-  space of \emph{traits} that comprise those individuals.
-  Holland termed this implicit large-scale sampling of traits ``implicit
-  parallelism.''    Evolutionary dynamics biases these samples over time towards
+\item{\bf Populations, sampling, and implicit parallelism:}
+
+% Adaptation occurs in
+% populations of individuals evolving (or learning) over time, in
+% which statistics can be leveraged to direct population dynamics.
+%  The process of adaptive 
+  Evolution can be framed as a search that leverages statistics to
+  direct population dynamics.  A population can be thought of as a
+  parallel sampling of many individuals (from the space of possible
+  individuals).  Moreover, that same population can be thought of as
+  implicitly sampling a much larger space of \emph{traits} that
+  comprise those individuals.  Holland termed this implicit
+  large-scale sampling of traits ``implicit parallelism.''
+  Evolutionary dynamics biases these samples over time towards
   high-fitness regions of the search space.
@@
 \item{\bf Building blocks and recombination:} 
-% Melanie rewrote this some to take out references to ``solutions''
   In a population undergoing adaptation,
@@
 
-  To succeed, such a search requires first that
-  useful building blocks be preserved once discovered, and second that
-  there be an efficient mechanism for combining promising building
-  blocks into higher-order building blocks.  Holland proposed using an
-  analog of Darwinian selection for the first and the idea of crossing
-  over in genetics for the second.  Holland referred to the building
-  blocks as \emph{schemas} and defined them as hyperplanes in the search space.
+%  To succeed, such a search requires first that
+%  useful building blocks be preserved once discovered, and second that
+%  there be an efficient mechanism for combining promising building
+%  blocks into higher-order building blocks.  Holland proposed using an
+%  analog of Darwinian selection for the first and the idea of crossing
+%  over in genetics for the second.  Holland referred to the building
+%  blocks as \emph{schemas} and defined them as hyperplanes in the search space.
 
@@
   requires maintaining a balance between \emph{exploitation}, in which
- successful building blocks propagate in a population, and \emph{
-    exploration}, in which existing building blocks are recombined or
-  mutated in new ways.  Inspired by Bellman \cite{Bellman1961} and
-others, Holland formalized this exploitation-versus-exploration tradeoff
-via an idealized ``two-armed bandit'' problem.  Given a slot machine
-with two arms, each of which has an unknown payoff probability, how
-should you allocate $N$ trials (pulls) between the arms so as to
-maximize your total payoff?   In \cite{Holland1973,Holland1975}
-Holland 
-%derived
-%an equation for such a strategy and 
-argued that the optimal strategy
-allocates trials to the observed best arm at a slightly higher rate than
-exponential.
-% and extended the result to multi-armed bandits.
-%A system that focuses too much on exploitation
-%  risks never finding better individuals or not being able to adapt to
-%  changing environments.  Conversely, too much focus on exploration
-%  means that the system won't leverage the successful building blocks
-%  that have already been discovered.  %There is some optimal balance
-%  between these two modes of processing.
-Next, he showed that 
-populations undergoing adaptation resemble multi-armed bandits, where
-each building block is an 
-%undergoing adaptation constitute the 
-arms in a multi-armed bandit problem.
-%Each building block can be said to have a probability of ``payoff''
-%(i.e., contribution to a given individual's fitness).
-Thus, in the process of assigning a fitness to each individual in a
-population, adaptive evolution can be seen as implicitly sampling the
-many building blocks making up that collection of individuals.  The
-average fitness over many individuals containing a particular building
-block gives an estimate of that building block's payoff probability.
+  successful building blocks propagate in a population, and
+  \emph{exploration}, in which existing building blocks are recombined
+  or mutated in new ways.
 
@@
 
+Inspired by Bellman \cite{Bellman1961} and
+others, Holland formalized the exploitation-versus-exploration tradeoff
+via an idealized ``two-armed bandit'' problem. Given a slot machine
+with two arms, each of which has an unknown payoff probability, how
+should you allocate trials (pulls) between the arms so as to
+maximize your total payoff?   In \cite{Holland1973,Holland1975}
+Holland argued that the optimal strategy
+allocates trials to the observed best arm at a slightly higher rate than
+an exponential function of the trials allocated to the observed worse arm.
+
+Next, Holland showed that in populations undergoing adaptation,
+building blocks are analogous to arms on a multi-armed bandit.  
+%Thus,
+%in the process of assigning a fitness to each individual in a
+%population, adaptive evolution can be seen as implicitly sampling the
+%many building blocks making up that collection of individuals.
+Evaluating an individual in an environment is analogous to pulling selected
+arms on a multi-armed bandit, where the arms correspond to each of the
+building blocks making up that individual.
+
+The question of how to balance exploitation and exploration---how to
+optimally allocate trials to different arms based on their observed 
+payoff---now becomes the question of how to optimally sample in the
+vast space of possible building blocks, based on their estimated
+contribution to fitness.  Of course, evolution deals in populations of
+individuals, not building blocks.  There is no {\emph explicit} mechanism for
+keeping statistics on how building blocks contribute to fitness. 
+Holland's central idea here is that an approximation to
+optimal building-block sampling does implicitly occur, as an emergent
+property of evolutionary population dynamics.
+
 Several of Holland's early papers (e.g.,
-\cite{Holland1962,Holland1973}) and his influential
-1975 book \emph{Adaptation in Natural and Artificial Systems}
-\cite{Holland1975} developed a general, formal setting in which these ideas
-could be expressed mathematically.
-
-It was this formalization
-% and resulting theorems that led to the
-that led to the invention of genetic algorithms, which featured stochastic
-population-based search, as well as
-%recombination as a critical
-crossover between individuals as a critical
-operation that allowed successful building blocks to be recombined and
-tested in new contexts.  What made GAs unique
-among other evolution-inspired algorithms at the time were the
-mathematical foundations described above, the emphasis on populations
-and recombination as central mechanisms,
-and a focus 
-%(at least in Holland's mind) [MM SAYS: THIS SOUNDS TOO NEGATIVE]
-on continual adaptation to
-non-stationary environments rather than optimization to static
-environments.
-
-The framework Holland developed in \cite{Holland1975} was more general
-than the genetic algorithm; its aim was an interdisciplinary theory of
-adaptation, one that would inform biology, say, as much as computer
-science \cite{Christiansen1998}.  The later, successful application of
+\cite{Holland1962,Holland1973}) and his influential 1975 book
+\emph{Adaptation in Natural and Artificial Systems} \cite{Holland1975}
+developed a general, formal setting in which these ideas could be
+expressed mathematically.  It was this formalization that led to the
+invention of genetic algorithms, which featured stochastic
+population-based search, as well as crossover between individuals as a
+critical operation that allowed successful building blocks to be
+recombined and tested in new contexts.  Holland's aim, however, was
+more general than a new class of algorithms---his goal was an
+interdisciplinary theory of adaptation, one that would inform biology, 
+say, as much as computer science \cite{Christiansen1998}.  The later, successful application of
 genetic algorithms to real-world optimization and learning tasks was,
 for Holland, just icing on the cake.
+
+% SF liked 'icing on the cake.'
+
+%What made GAs unique
+%among other evolution-inspired algorithms at the time were the
+%mathematical foundations described above, the emphasis on populations
+%and recombination as central mechanisms,
+%and a focus 
+%(at least in Holland's mind) [MM SAYS: THIS SOUNDS TOO NEGATIVE]
+%on continual adaptation in
+%non-stationary environments rather than optimization in static
+%environments.
 
@@
 
-Internal models are central to Holland's theory of adaptive
-systems.  He posits that all adaptive systems create and use internal
-models to prosper in the face of ``perpetual novelty'' in continually changing environments.   Models can be tacit and learned
-  through evolutionary time, as in the case of bacteria swimming up a
-  chemical gradient, or explicit and learned over a single lifespan,
-  as in the case of cognitive systems that incorporate experience into
-  internal representations through learning.  In
-  Holland's view, the key activity of an adaptive agent involves
-  building and refining these data-driven models
-  of the environment.
+Internal models are central to Holland's theory of adaptive systems.
+He posits that all adaptive systems create and use internal models to
+prosper in the face of ``perpetual novelty.'' 
+%in continually changing environments.  
+Models can be tacit and learned through evolutionary
+time, as in the case of bacteria swimming up a chemical gradient, or
+explicit and learned over a single lifespan, as in the case of
+cognitive systems that incorporate experience into internal
+representations through learning.  In Holland's view, the key activity
+of an adaptive agent involves building and refining these data-driven
+models of the environment.
 %using them to make predictions and take actions,
@@
 In his second book, \emph{Induction} \cite{Holland1989}, Holland and
-his co-authors tackled the question of how cognitive agents can
-possibly learn and profit from internal models by combining
+his co-authors proposed inductive methods by which 
+%tackled the question of how 
+cognitive agents can
+%learn 
+construct---and profit from---internal models by combining
 environmental inputs and rewards with stored knowledge.  In their
 framework, a model defines a set of equivalence relations over
-environmental states, together with a set of transition rules, which
+environmental states, together with a set of transition rules which
 are learned over time based on environmental rewards (or punishments).
@@
 system to make accurate predictions.  In Holland's conception, the
-equivalence classes are initially very general, say ``moving object''
+equivalence classes are initially very general, for example, ``moving object''
 and ``stationary object.'' Through experience and
@@
 a model is consistent with the environment and how an intelligent
-agent, human or artificial, can update the model to better reflect
+agent, human or artificial, can update a model to better reflect
 reality.
@@
 mechanism, the \emph{learning classifier
-  system}~\cite{Holland1977,Holland1986}, to illustrate how a cognitive system could
+  system}~\cite{Holland1986,Holland1977}, to illustrate how a cognitive system could
 iteratively build up a detailed and hierarchical model of its
@@
 learning.
+
+%Offspring rules, which begin life weaker than do their parents, can
+%win the competition and influence the system's behavior only when
+%there are not strong rules whose conditions are satisfied---in other
+%words when the system does not know what to do.  If their actions
+%help, they survive; if not they are soon replaced.  
+%Thus,
 
@@
 learning, artificial intelligence, evolutionary biology, economics,
-psychology, control theory, and other fields (e.g.,
-\cite{Bellman1961,Samuel1959}).  Knowledge representation in the form of a
+psychology, and control theory.
+% MM says: We cite these elsewhere in the paper
+%(e.g.,\cite{Bellman1961,Samuel1959}).  
+Knowledge representation in the form of a
 population of ``if-then'' rules seemed like a good choice, not only
@@
 occurs, then I will fire for a while at a high
-rate.'' \cite{Waldrop1993} (p. 182).  
+rate'' \cite{Waldrop1993}. 
 The if-then rules, when activated, compete to post their results on a
-shared ``message list,'' modeling the system's short-term memory
+shared ``message list,'' serving as the system's short-term memory
 (again inspired by Hebb's work and AI blackboard systems of the day).
-Unlike the AI systems, however, new rules were generated automatically
-%in a trial-and-error fashion
-using a genetic algorithm.
+Unlike blackboard systems, however, new rules are generated automatically
+in a trial-and-error fashion.
+% SF reverted this because lcs had several non-GA learning mechanisms,
+% like 'covering'
+%using a genetic algorithm.
 
-Successful rules were strengthened over time if their predictions led
+Successful rules are strengthened over time if their predictions led
 to positive rewards from the environment (and weakened otherwise)
@@
 algorithm, in which rules gaining rewards from the environment or from
-other rules transferred some of their gains to those earlier-firing
+other rules transfer some of their gains to those earlier-firing
 ``stage-setting'' rules that set up the conditions for the eventual
@@
 %, e.g., finding a target, or avoiding a predator,
+
+Holland was primarily interested in how the two
+learning mechanisms (discovery of new rules and apportioning credit to
+existing rules) could work together to create 
+%subsets of rules
+%that accurately model the environment, and thus are able to ``fire''
+%at appropriate times.  
+%Holland envisioned that, with a complex environment and enough time
+%for learning, a classifier system might learn an intricate default
+useful default hierarchies of rules.  He emphasized that the
+competition inherent in the learning and action mechanisms would allow
+the system to adapt to a continually changing environment without
+losing what it had learned in the past.  Holland put it this way:
+``Competition among rules provides the system with a graceful way of
+handling perpetual novelty.  When a system has strong rules that
+respond to a particular situation, that is the equivalent of saying
+that it has certain well-validated hypotheses.... New rules do not
+interfere with the system's action in well-practiced situations but
+wait gracefully in the wings as hypotheses about what to do under
+novel circumstances.'' \cite{Holland1992}.
 
@@
 %   methods.}
-
-Holland was primarily interested in how the two
-learning mechanisms (discovery of new rules and apportioning credit to
-existing rules) could work together to create 
-%subsets of rules
-%that accurately model the environment, and thus are able to ``fire''
-%at appropriate times.  
-%Holland envisioned that, with a complex environment and enough time
-%for learning, a classifier system might learn an intricate default
-useful default hierarchies of rules.  He emphasized that the
-competition inherent in the learning and action mechanisms would allow
-the system to adapt to a continually changing environment without
-losing what it had learned in the past.  Holland put it this way:
-``Competition among rules provides the system with a graceful way of
-handling perpetual novelty.  When a system has strong rules that
-respond to a particular situation, that is the equivalent of saying
-that it has certain well-validated hypotheses.... New rules do not
-interfere with the system's action in well-practiced situations but
-wait gracefully in the wings as hypotheses about what to do under
-novel circumstances.'' \cite{Holland1992}.
-
-%Offspring rules, which begin life weaker than do their parents, can
-%win the competition and influence the system's behavior only when
-%there are not strong rules whose conditions are satisfied---in other
-%words when the system does not know what to do.  If their actions
-%help, they survive; if not they are soon replaced.  
-%Thus,
 
@@
 
-Holland was interested in models that explored basic principles and
-mechanisms, even if they did not make specific or detailed
+Holland's work focused on {\emph exploratory} models---those that explore basic principles and
+mechanisms, even if they do not make specific or detailed
 predictions.  Such models can show generically how certain behaviors
@@
 
-A given agent-based model encodes a theory about the mechanisms that
-are relevant for producing the behavior of interest.  
+%MM says:  This next sentence not needed. 
+%A given agent-based model encodes a theory about the mechanisms that
+%are relevant for producing the behavior of interest.
+
 %Similar to expert systems [MM says: ?]
@@
 
-\section{Complex Adaptive Systems}
-
-%\section{Complex Adaptive Systems}
+\section{Universal Properties of Complex Adaptive Systems}
 
@@
 systems, ecologies, financial markets, cities, and the brain---systems
-that are \emph{complex}.  In the 1980s, he teamed up with a small
+that are \emph{complex}.  In the early 1980s, he teamed up with a small
 group of scientists, primarily physicists with a sprinkling of
@@
 %these systems exhibit 
-adaptative, with 
+adaptive, with 
 behavioral rules continually adjusting through evolution and learning.  Although this
@@
 
-ECHO \cite{Holland1999,Hraber1997} was an even more ambitious model
-that Holland and collaborators developed during the 1990s. 
+ECHO \cite{Holland1999,Hraber1997} was an even more ambitious exploratory model
+that Holland and collaborators developed next. 
 % ECHO's
@@
 resources, form higher
-level aggregate structures (such trading relationships, symbiotic groups,
+level aggregate structures (such as trading relationships, symbiotic groups,
 trophic cascades, interdependent organizations, etc.).  ECHO agents
-sometimes discovered mimicry to deceive competitors, and over time,
-the model developed increasingly complex structures and behaviors, at
-least up to a point.  ECHO reproduced several well-known patterns 
-observed in nature, for example, over time the population of ECHO
-evolved agents exhibited diversity of agent types, diversity with a 
-rank-frequency distribution that
-closely paralleled the well-known Preston curve in ecology.  In
-layman's terms, ECHO evolved populations where 
+sometimes discovered mimicry and used it to deceive competitors.  
+In many runs,  
+the model developed increasingly complex structures and behaviors up
+to a point.  When these runs stabilized, the resulting population of agents
+% at least up to a point.   
+was found to reproduce several well-known patterns 
+observed in nature, most famously the rank-frequency distribution of
+species diversity known as the Preston curve in ecology.
+In layman's terms, ECHO evolved populations where 
 %quantitative statement of the adage that
@@
 The broad scope of the model, together with its ability to produce
-easily identifiable patterns, was appealing to immunologists, economists, and evolutionary biologists alike.   Many
+easily identifiable and well-known patterns from nature, was appealing to immunologists, economists, and evolutionary biologists alike.   Many
 of the insights behind this project are described in Holland's third book \emph{Hidden Order} \cite{Holland1995}.
@@
 willingness to tackle the hardest questions, develop his own
-formalisms, and use mathematics productively sets a high bar to whch we
-all should aspire.
+formalisms, and use mathematics to provide insight sets a high bar 
+scientists in all disciplines.
+%to whch we
+%all should aspire.
 
@@
 
-Holland was unusual in his ability to absorb the essence of other
+John Holland was unusual in his ability to absorb the essence of other
 disciplines, articulate grand overarching
-theories, and them back them up with computational mechanisms and
+principles, and them back them up with computational mechanisms and
 mathematics.  Unlike most researchers he moved seamlessly among these
@@
 machine learning today, e.g., reinforcement, learning in non-Markovian
-environments, active learning, etc.   His seminal genetic algorithm
+environments and active learning.   His seminal genetic algorithm
 spawned the field of evolutionary computation, and his insights
 and wisdom helped define what are now referred to as the
-\emph{sciences of complexity}.   With such a broad scope, it is easy
-to pick at the details.
+\emph{sciences of complexity}.
 
-His many books and articles have influenced 
-scientists around the work and across many disciplines.  Closer to home, he introduced several
-generations of students at the University of Michigan to 
-computation in natural systems, an idea that even today
-remains somewhat controversial, despite successes in biomimicry
-for robotics, abstractions of phermone trails in ant colonies for
-optimization methods, the use of mechanisms from immunology to
-improve computer security, and recent work on technological
-evolution.
+Holland's many books and articles have influenced scientists around
+the work and across many disciplines.  Closer to home, he introduced
+several generations of students at the University of Michigan to
+computation in natural systems, an idea that even today remains
+somewhat controversial, despite successes in genetic algorithms for
+engineering design, biomimicry for robotics, abstractions of phermone
+trails in ant colonies for optimization methods, and the use of
+mechanisms from immunology to improve computer security.
 
@@
 Holland's personality and humanity were somehow inextricably tangled
-up with his intellectual contributions, and although his ideas will serve as a beacon
-for research in intelligent and complex systems
-for many years to come, the world of science is poorer without John
-Holland on the scene.
+up with his intellectual contributions.  Since his death, many of
+Holland's former students and colleagues have movingly described their
+desire to emulate his personal qualities as much as his scientific
+excellence.  John Holland's ideas, intellectual passion, and personal approach will serve as
+beacons for research in intelligent and complex systems for many
+years to come. 
+%the world of science is poorer without John
+%Holland on the scene.
 

### 2015-12-15 paper.tex
Submitted Draft
Submitted draft of paper.

@@
 
-% \begin{abstract}
-% \input{abstract}
-% \end{abstract}
+\begin{abstract}
+In this essay we review the work and influence of the late John H.
+Holland, pioneer in the fields of evolutionary computation, machine learning, and complex systems.
+While best known for his invention of \emph{genetic algorithms}, Holland's scientific motivation was much broader:  to develop a general theory of adaptation in complex systems.
+We survey Holland's motivation and work with respect to four central themes, and describe how this work has had both broad and deep influence in modern approaches to computer science and complex systems science.
+\end{abstract}
 
@@
 commercial applications.  While the mechanisms of GAs are well-known,
-%Holland's broader motivations are less widely appreciated---to develop a 
-%general theory of adaptation in complex systems.  
 they were only one
@@
 theory of adaptation in complex systems.
-%This motivation was the
-%driving force in all of Holland's research.   
-%to much of the computer science community, 
-%the ideas that motivated Holland are less widely appreciated---to develop a general
-%Holland's motivation 
-%offshoot of Holland's much broader motivation--
-%This motivation was the
-%driving force in all of Holland's research.
 
@@
 exploratory modeling; and (4) universal properties of complex adaptive systems.
-%We discuss the role these
-%themes have played in computer science, and highlight especially the
-%ideas that we think remain relevant to today's research agendas.  [MM
-%says: Are we going to do this?]
 
@@
 
-Holland's goal of developing a general theory of adaptation 
-%spurred
-began with his early work on computer models of Hebbian learning
+Holland's goal of developing a general theory of adaptation was
+spurred both by his early work on computer models of Hebbian learning
 \cite{Rochester1956} and his reading of Ronald Fisher's classic book
-that integrated genetics with Darwinian selection \cite{Fisher1930}.
-As Holland read extensively in evolutionary biology,
+on integrating genetics with Darwinian selection \cite{Fisher1930}.
+As Holland further read extensively in evolutionary biology,
 economics, game theory, and control theory, he came to recognize that
@@
 these fields all concern populations of agents that must continually
-obtain information from uncertain changing environments and use it in
+obtain information from uncertain, changing environments and use it in
 order to improve performance and enhance the chance of survival.
@@
 \item{\bf Populations, sampling, and implicit parallelism:} 
-
-% Adaptation occurs in
-% populations of individuals evolving (or learning) over time, in
-% which statistics can be leveraged to direct population dynamics.
-%  The process of adaptive 
   Evolution can be framed as a search that leverages statistics to
@@
 
-%The population can be
-%  thought of as a set of samples in a much larger search space, where
-%  the processes of evolution
-%  population can be thought of as a set of sa in which statistics 
-%  can be leveraged to direct population dynamics.  (More on this below.)
-
 \item{\bf Building blocks and recombination:} 
@@
   are discovered in stages, first by finding useful
-  building blocks through random sampling, and over
+  building blocks through stochastic sampling, and over
   time recombining such building blocks to create higher-fitness individuals.
@@
 
-The question of how to balance exploitation and exploration---how to
+The question of balancing exploitation and exploration---how to
 optimally allocate trials to different arms based on their observed 
-payoff---now becomes the question of how to optimally sample in the
+payoffs---now becomes the question of how to optimally sample in the
 vast space of possible building blocks, based on their estimated
@@
 
+%What made GAs unique
+%among other evolution-inspired algorithms at the time were the
+%mathematical foundations described above, the emphasis on populations
+%and recombination as central mechanisms,
+%and a focus 
+%(at least in Holland's mind) [MM SAYS: THIS SOUNDS TOO NEGATIVE]
+%on continual adaptation in
+%non-stationary environments rather than optimization in static
+%environments.
+
+%The later, successful application of
+%genetic algorithms to real-world optimization and learning tasks was,
+%for Holland, just icing on the cake.
 % SF liked 'icing on the cake.'
@@
 prosper in the face of ``perpetual novelty.''
-%in continually changing environments.  
-Models can be tacit and learned through evolutionary
-time, as in the case of bacteria swimming up a chemical gradient, or
-explicit and learned over a single lifespan, as in the case of
-cognitive systems that incorporate experience into internal
-representations through learning.  In Holland's view, the key activity
-of an adaptive agent involves building and refining these data-driven
-models of the environment.
-%using them to make predictions and take actions,
-%  and occasionally receiving positive or negative feedbacks
-%  (intermittent rewards).
+
+Models can be tacit and learned through evolutionary time, as in the
+case of bacteria swimming up a chemical gradient, or explicit and
+learned over a single lifespan, as in the case of cognitive systems
+that incorporate experience into internal representations through
+learning.  In Holland's view, the key activity of an adaptive agent
+involves building and refining these data-driven models of the
+environment.
 
 In his second book, \emph{Induction} \cite{Holland1989}, Holland and
-his co-authors proposed inductive methods by which 
-%tackled the question of how 
-cognitive agents can
-%learn 
-construct---and profit from---internal models by combining
+his co-authors proposed inductive methods by which cognitive agents
+can construct---and profit from---internal models by combining
 environmental inputs and rewards with stored knowledge.  In their
-framework, a model defines a set of equivalence relations over
-environmental states, together with a set of transition rules which
-are learned over time based on environmental rewards (or punishments).
-Models that form valid homomorphisms with the environment allow the
-system to make accurate predictions.  In Holland's conception, the
-equivalence classes are initially very general, for example, ``moving object''
-and ``stationary object.'' Through experience and
-learning, these can be specialized into more useful and precise subclasses, say
-``insect'' and ``nest.''  Over time, the adaptive system builds up a
-default hierarchy of rules covering general cases and refinements for
-specific classes.
+framework, a model defines a set of equivalence relations---a
+homomorphism---over environmental states, together with a set of
+transition rules which are learned over time based on environmental
+rewards (or punishments).  Models that form valid homomorphisms with
+the environment allow the system to make accurate predictions.  In
+Holland's conception, the equivalence classes are initially very
+general, for example, ``moving object'' and ``stationary object.''
+Through experience and learning, these can be specialized into more
+useful and precise subclasses, say ``insect'' and ``nest.''  Over
+time, the adaptive system builds up a default hierarchy of rules
+covering general cases and refinements for specific classes.
 
-% For example, a general rule might
-% state that, ``If X is a bird, then X can fly'', whereas a more specific
-% one might state, ``If X is a bird with small wings and a large body,
-% then X cannot fly'' \cite{Holland1989}.  When faced with a bird in
-% the environment, both rules might compete to post their conclusion,
-% but the more specific one would be favored if both of its conditions are met.
-
-% Maybe this should be moved to section on Complexity
-Although the idea of default hierarchies was prevalent in 
-knowledge representation systems of the era, Holland made two key
-contributions.  The first was his emphasis on homomorphisms as a formal way to
+Although the idea of default hierarchies was prevalent in knowledge
+representation systems of the era, Holland made two key contributions.
+The first was his emphasis on homomorphisms as a formal way to
 evaluate model validity, an idea that dates back to Ross Ashby's
-\emph{An Introduction to Cybernetics} \cite{Ashby1956}. Holland's student
-Bernard Ziegler developed this idea into a formal theory of computer modeling
-and simulation \cite{Ziegler1976}.  
-% %Today, even though computational modeling is a major
-% %application area of computing, essential to most natural sciences and
-% %engineering disciplines
-Even today, these early homomorphic theories of modeling
-stand as the most elegant approach we know of to characterize when
-a model is consistent with the environment and how an intelligent
-agent, human or artificial, can update a model to better reflect
-reality.
+\emph{An Introduction to Cybernetics} \cite{Ashby1956}. Holland's
+student Bernard Ziegler developed this idea into a formal theory of
+computer modeling and simulation \cite{Ziegler1976}.  Even today,
+these early homomorphic theories of modeling stand as the most elegant
+approach we know of to characterize when a model is consistent with
+the environment and how an intelligent agent, human or artificial, can
+update a model to better reflect reality.
 
@@
 environment to enhance survival.  The key learning elements of this method, the
-bucket-brigade algorithm combined with a genetic algorithm, presaged many of the ideas in modern reinforcement
+\emph{bucket-brigade} algorithm combined with a genetic algorithm, presaged many of the ideas in modern reinforcement
 learning.
@@
 %(e.g.,\cite{Bellman1961,Samuel1959}).  
-Knowledge representation in the form of a
-population of ``if-then'' rules seemed like a good choice, not only
-because of its popularity in AI at the time but also from Holland's
-early work on modeling Hebbian cell assemblies: ``In Hebb's view, a
-cell assembly makes a simple statement: If such and such an event
-occurs, then I will fire for a while at a high
-rate'' \cite{Waldrop1993}. 
-The if-then rules, when activated, compete to post their results on a
-shared ``message list,'' serving as the system's short-term memory
-(again inspired by Hebb's work and AI blackboard systems of the day).
-Unlike blackboard systems, however, new rules are generated automatically
-in a trial-and-error fashion.
-% SF reverted this because lcs had several non-GA learning mechanisms,
-% like 'covering'
-%using a genetic algorithm.
+Knowledge representation in the form of a population of ``if-then''
+rules seemed like a good choice, not only because of its popularity in
+AI at the time but also from Holland's early work on modeling Hebbian
+cell assemblies: ``In Hebb's view, a cell assembly makes a simple
+statement: If such and such an event occurs, then I will fire for a
+while at a high rate'' \cite{Waldrop1993}.  The if-then rules, when
+activated, compete to post their results on a shared ``message list,''
+serving as the system's short-term memory (again inspired by Hebb's
+work and AI blackboard systems of the day).  Unlike blackboard
+systems, however, new rules are generated automatically in a
+trial-and-error fashion, and can be selected and recombined via a
+genetic algorithm.
 
@@
 through a credit-assignment method called the \emph{bucket-brigade}
-algorithm, in which rules gaining rewards from the environment or from
-other rules transfer some of their gains to those earlier-firing
+algorithm, in which rules gaining rewards from the environment, or from
+other rules, transfer some of their gains to earlier-firing
 ``stage-setting'' rules that set up the conditions for the eventual
@@
 learning applied to checkers \cite{Samuel1959} as a key inspiration for these ideas.
-
-%, e.g., finding a target, or avoiding a predator,
 
@@
 %computational system more seriously
-applying it to areas as diverse
-as poker-playing \cite{Smith1980}, control of gas pipeline
-transmission \cite{Goldberg1983}, and modeling the stock market
-\cite{Palmer1994}.  (See Ref.~\cite{Booker1989} for more details about
-practical applications of classifier systems.)  Today, other
-reinforcement learning methods are more popular for real-world
-decision and control problems, but classifier systems can perhaps be
-thought of as an essential stage-setting method that enabled the
-development of later approaches.
+applying it to areas as diverse as poker-playing \cite{Smith1980},
+control of gas pipeline transmission \cite{Goldberg1983}, and modeling
+the stock market \cite{Palmer1994}.  (See Ref.~\cite{Booker1989} for
+more details about practical applications of classifier systems.)
+Today, other reinforcement learning methods are more popular for
+real-world decision and control problems, but classifier systems can
+perhaps be thought of as an early stage-setting method that helped
+build the foundations for later approaches.
 
@@
 order to capture the behavior of systems under spatial
-constraints, these models are often defined over spatial structures,
+constraints, these models are often defined over spatial structures
 such as networks or simple grids.
@@
 %systems seciton.]
-The exploratory models
-championed by Holland were idealized and not intended to provide detailed, domain-specific
-predictions.  Instead they were meant to explore possible general
-mechanisms of complex systems and thus provide insights that might
-lead to more specific, detailed models.  Such idealized models are
-akin to what Dennett has called ``intuition pumps''
-\cite{Dennett1984}.
+The exploratory models championed by Holland were not intended to
+provide detailed, domain-specific predictions.  Instead they were
+meant to explore possible general mechanisms of complex systems and
+thus provide insights that might lead to more specific, detailed
+models.  Such idealized models are akin to what philosopher Daniel Dennett has called
+``intuition pumps'' \cite{Dennett1984}.
 
@@
 
-It should be noted that Holland's view of modeling is by no means
-typical.  For example, a textbook on computational modeling offers
- the following definition: ``Modeling is the application
-of methods to analyze complex, real-world problems in order to make
-predictions about what might happen with various actions''
-\cite{Shiflet2014}.   This perspective rules out the
-kind of exploratory modeling that Holland was most interested in.
+%It should be noted that Holland's view of modeling is by no means
+%typical.  For example, a textbook on computational modeling offers
+% the following definition: ``Modeling is the application
+%of methods to analyze complex, real-world problems in order to make
+%predictions about what might happen with various actions''
+%\cite{Shiflet2014}.   This perspective rules out the
+%kind of exploratory modeling that Holland was most interested in.
 
@@
 including several economists, to tackle the mismatch between predictions
-of rational expectations theory (the then dominant theory in economics) 
+of rational expectations theory (the dominant theory in economics at the time) 
 and empirically observed stock market behaviors. 
@@
 economic theory of the day assumed that all participants in an economy
-or financial market are 100\% rational and act to maximize their
+or financial market are fully rational, and act to maximize their
 individual gain.  In real life, however, actors in economies
@@
 ECHO \cite{Holland1999,Hraber1997} was an even more ambitious exploratory model
-that Holland and collaborators developed next. 
+that Holland and collaborators developed during the 1990s. 
 % ECHO's
@@
 % system's learned ability to distinguish `self' from `other.'
-ECHO formalized Holland's idealization of complex adaptive systems into
-a runnable computational system where agents evolved external markers
-(called tags) and internal preferences, and then used them to acquire
-resources, form higher
-level aggregate structures (such as trading relationships, symbiotic groups,
-trophic cascades, interdependent organizations, etc.).  ECHO agents
-sometimes discovered mimicry and used it to deceive competitors.  
-In many runs,  
-the model developed increasingly complex structures and behaviors up
-to a point.  When these runs stabilized, the resulting population of agents
-% at least up to a point.   
-was found to reproduce several well-known patterns 
-observed in nature, most famously the rank-frequency distribution of
-species diversity known as the Preston curve in ecology.
-In layman's terms, ECHO evolved populations where 
-%quantitative statement of the adage that
-``most species are rare.''  
+ECHO formalized Holland's idealization of complex adaptive systems
+into a runnable computational system where agents evolved external
+markers (called tags) and internal preferences, and then used them to
+acquire resources, form higher-level aggregate structures, such as
+trading relationships, symbiotic groups, trophic cascades, and
+interdependent organizations.  ECHO agents sometimes discovered
+mimicry and used it to deceive competitors.  In many runs, the model
+developed increasingly complex structures and behaviors up to a point.
+When these runs stabilized, the resulting population of agents was
+found to reproduce several well-known patterns observed in nature,
+most famously the rank-frequency distribution of species diversity
+known as the Preston curve in ecology.  In layman's terms, ECHO
+evolved populations where ``most species are rare.''
+
 The broad scope of the model, together with its ability to produce
-easily identifiable and well-known patterns from nature, was appealing to immunologists, economists, and evolutionary biologists alike.   Many
-of the insights behind this project are described in Holland's third book \emph{Hidden Order} \cite{Holland1995}.
+easily identifiable and well-known patterns from nature, was appealing
+to immunologists, economists, and evolutionary biologists alike.  Many
+of the insights behind this project are described in Holland's third
+book, \emph{Hidden Order} \cite{Holland1995}.
 
@@
 willingness to tackle the hardest questions, develop his own
-formalisms, and use mathematics to provide insight sets a high bar 
+formalisms, and use mathematics to provide insight sets a high bar for
 scientists in all disciplines.
@@
 principles, and them back them up with computational mechanisms and
-mathematics.  Unlike most researchers he moved seamlessly among these
+mathematics.  Unlike most researchers, Holland moved seamlessly among these
 three modes of thinking, developing models that were years ahead of their time.  A
@@
 Holland's many books and articles have influenced scientists around
-the work and across many disciplines.  Closer to home, he introduced
+the world and across many disciplines.  Closer to home, he introduced
 several generations of students at the University of Michigan to
@@
 somewhat controversial, despite successes in genetic algorithms for
-engineering design, biomimicry for robotics, abstractions of phermone
+engineering design, biomimicry for robotics, abstractions of pheromone
 trails in ant colonies for optimization methods, and the use of
@@
 encountered a new idea; his willingness to talk to anyone, no matter
-how famous or how unfamous; and his incredible generosity and patience.
+how famous or obscure; and his incredible generosity and patience.
 %His insights aboutwere deeper and more general than what often passes for work in biomimicry, e.g., for robots.

### 2016-02-08 misc.tex
incorporated reviewer suggestions as discussed in phone call.  Added acknowledgments and steph's grant support.  still need a footntoe with date and stating where it will be published.

%[MM SAYS:  IF WE WANT TO MAKE THIS SECTION SHORTER, WE COULD CUT EVERYTHING FROM HERE UP TO ``[UP TO HERE]'' (SEE BELOW)]

Inspired by Bellman \cite{Bellman1961} and
others, Holland explored the exploitation-versus-exploration tradeoff
via an idealized ``two-armed bandit'' problem.  Given a slot machine
with two arms, each of which has an unknown payoff probability, how
should you allocate $N$ trials (pulls) between the arms so as to
maximize your total payoff?  An example of an extreme \emph{exploitation} strategy would be to alternate between the arms until
one of them gives a payoff, and then allocate all future trials to
that arm alone.  Conversely, an example of an extreme \emph{exploration} strategy would be to randomly allocate the trials,
irrespective of the payoff rates you obtain.  Obviously, each of these
strategies is flawed.  What is an optimal strategy?
In \cite{Holland1973,Holland1975} Holland derived
an equation for such a strategy and argued that the optimal strategy
allocated trials to the observed best arm at a slightly higher rate than
exponential.  
% SF used 'argue' because some claim that his derivation was wrong.
He then extended this 
% Let $A$ denote the arm currently
%observed to have the higher payoff probability, and $B$ denote the
%other arm.  Holland showed that the optimal strategy (the one that
%yields maximal payoff, or minimal loss) is for the number of trials
%allocated to $A$ to grow slightly faster than an exponential function
%of the number of trials allocated to $B$.  Holland then showed that this
result to the \emph{multi-armed bandit} case.

Holland then made an analogy between multi-armed bandits and
populations undergoing adaptation: building blocks in a population
undergoing adaptation constitute the ``arms'' of a multi-armed bandit.
Each building block can be said to have a probability of ``payoff''
(i.e., contribution to a given individual's fitness).
%Evaluating an individual in an environment is
%like ``pulling the arms'' on a multi-armed bandit, where the ``arms''
%correspond to each of the building blocks making up that individual.
Thus, in the process of assigning a fitness to each individual in a
population, adaptive evolution can be seen as implicitly sampling the
many building blocks making up that collection of individuals.  The
average fitness over many individuals containing a particular building
block gives an estimate of that building block's payoff probability.

% The question of how to balance exploitation and exploration---how to
% optimally allocate trials to different arms based on their observed
% payoff---now becomes the question of how to optimally sample in the
% vast space of possible building blocks, based on their observed
% average fitness.  Of course adaptive evolution deals in populations of
% individuals, not building blocks.  There is no explicit mechanism for
% keeping statistics on building blocks' ``observed average fitness''.
% However, Holland's central idea here is that an approximation to
% optimal building-block sampling does indeed occur, as an emergent
% property of the population dynamics.

%Holland argued that an approximation to
%optimal building-block sampling actually occurs, as an emergent
%property of the population dynamics defined by a
% To show this mathematically, he defined an idealization of
%adaptive evolution: an algorithm he called a \emph{reproductive
% plan}.
To show this mathematically, Holland defined an idealization of adaptive evolution: 
a population-based stochastic process
operating on bit strings, involving the ``genetic'' operators of
 reproduction, fitness-based selection, crossover, mutation, and
 inversion.  
%Building blocks are formalized as \emph{
%   schemata}---patterns of bits within a string. (Here are two examples
% of schemas: ``all strings beginning with the bits 10'' or ``all
% strings that start with the pattern 1*1'', where * can be replaced by
% either 0 or 1).

% Holland's most important result was the following: He proved
% mathematically that his reproductive plan results---implicitly---in a
% near-optimal allocation of trials to schemata, thus optimizing the

% exploitation versus exploration balance.  Holland noted that the
% explicit act of assigning fitnesses to the individuals in a population
% was actually implicitly sampling a much larger collection of building
% blocks.  Holland termed this \emph{implicit parallelism}; the
% parallel leveraging of statistics from implicit sampling is a main
% strength of his population-based approaches to search.

[UP TO HERE]

### 2016-02-08 paper.tex
incorporated reviewer suggestions as discussed in phone call.  Added acknowledgments and steph's grant support.  still need a footntoe with date and stating where it will be published.

@@
 
-\title{Adaptative Computation:\\ An Essay in Memory of John H. Holland}
+\title{Adaptative Computation:\\ An Essay in Memory of John
+  H. Holland}
 
@@
 \begin{abstract}
-In this essay we review the work and influence of the late John H.
-Holland, pioneer in the fields of evolutionary computation, machine learning, and complex systems.
-While best known for his invention of \emph{genetic algorithms}, Holland's scientific motivation was much broader:  to develop a general theory of adaptation in complex systems.
-We survey Holland's motivation and work with respect to four central themes, and describe how this work has had both broad and deep influence in modern approaches to computer science and complex systems science.
+This essay reviews the work and influence of the late John H. Holland,
+pioneer in the fields of evolutionary computation, machine learning,
+and complex systems.  While best known for his invention of
+\emph{genetic algorithms}, Holland's scientific motivation was much
+broader:  to develop a general theory of adaptation in complex
+systems.  We survey Holland's motivation and research with respect to
+four central themes, describing his broad and deep influence on modern approaches to computer science and complex systems science.  
 \end{abstract}
@@
 
-Professor John H. Holland passed away recently in Ann Arbor, MI, where
+In Aug. 2015, Professor John H. Holland passed in Ann Arbor, MI, where
 he had served on the University of Michigan faculty for over 50 years.
@@
 
-As a descendant of the cybernetics era, Holland was strongly
+As a descendant of the cybernetics era, Holland was 
 influenced by the work of von Neumann, Wiener, Ashby, and Turing, all
@@
 pretty much anyone he came in contact with.  As a result, even though
-he received what was arguably the world's first computer science
-Ph.D. in 1959, his contributions are sometimes better known outside of
+he received what was arguably one of the world's first computer
+science
+% SF softened the claim here after reading the comments section on the blog.
+Ph.D. degrees in 1959\cite{London2003}, his contributions are sometimes better known outside of
 computer science than within.
@@
 \emph{evolutionary computation}, with widespread scientific and
-commercial applications.  While the mechanisms of GAs are well-known,
+commercial applications.  Although the mechanisms of GAs are well-known,
 they were only one
@@
 on integrating genetics with Darwinian selection \cite{Fisher1930}.
-As Holland further read extensively in evolutionary biology,
-economics, game theory, and control theory, he came to recognize that
-adaptation was the most central concept in all these fields.  That is,
+As Holland read further in evolutionary biology,
+economics, game theory, and control theory, he recognized that
+adaptation is central to all these fields.  That is,
 these fields all concern populations of agents that must continually
-obtain information from uncertain, changing environments and use it in
-order to improve performance and enhance the chance of survival.
+obtain information from uncertain, changing environments and use it 
+to improve performance and enhance the chance of survival.
 
-Moreover, Holland recognized that adaptation must be a continual
+Holland also recognized that adaptation must be a continual
 open-ended process due to perpetually uncertain and changing
-environments.  Thus, adaptive systems never achieve a state of
+environments.  In his view, adaptive systems never achieve a state of
 equilibrium or a final optimum configuration.  This emphasis on
@@
 
-Underlying Holland's theory of adaptation are the following core ideas.
+Underlying Holland's theory of adaptation are the following core ideas:
 
@@
 \item{\bf Populations, sampling, and implicit parallelism:} 
-  Evolution can be framed as a search that leverages statistics to
-  direct population dynamics.  A population can be thought of as a
-  parallel sampling of many individuals (from the space of possible
-  individuals).  Moreover, that same population can be thought of as
-  implicitly sampling a much larger space of \emph{traits} that
+  Evolution is a form of search that leverages statistics to
+  direct population dynamics.  Initially, the population is an 
+ independent sample of many individuals (from the space of all possible
+  individuals), and over time the sample is increasingly biased
+  towards the high-fitness regions of the search space.  In addition,
+the population can be viewed as an
+  implicit sample of the much larger space of \emph{traits} that
   comprise those individuals.  Holland termed this implicit
@@
   traits that are the evolutionary ``atoms'' of an individual's
-  fitness or performance. (As an example from biology, Holland gives
+  fitness or performance. (As an example from biology, Holland often mentioned
   the \emph{Krebs cycle}, a core cellular metabolic
-  pathway.)   Successful individuals
+  pathway that is highly conserved across living systems.)   Successful individuals
   are discovered in stages, first by finding useful
   building blocks through stochastic sampling, and over
-  time recombining such building blocks to create higher-fitness individuals.
+  time recombining them to create higher-fitness individuals out of
+  more complex building blocks.
 
@@
 others, Holland formalized the exploitation-versus-exploration tradeoff
-via an idealized ``two-armed bandit'' problem. Given a slot machine
+as an idealized ``two-armed bandit'' problem. Given a slot machine
 with two arms, each of which has an unknown payoff probability, how
@@
 contribution to fitness.  Of course, evolution deals in populations of
-individuals, not building blocks.  There is no {\emph explicit} mechanism for
-keeping statistics on how building blocks contribute to fitness. 
+individuals, not building blocks.  There is no \emph{explicit} mechanism that
+keeps statistics on how building blocks contribute to fitness. 
 Holland's central idea here is that an approximation to
-optimal building-block sampling does implicitly occur, as an emergent
+optimal building-block sampling occurs implicitly, as an emergent
 property of evolutionary population dynamics.
@@
 population-based search, as well as crossover between individuals as a
-critical operation that allowed successful building blocks to be
-recombined and tested in new contexts.  Holland's aim, however, was
+critical operation that allows successful building blocks to be
+recombined and tested in new contexts.
+
+% Reviewer 3: I'd be careful with anything relating to the building block hypothesis (that recombination is successful by combining building blocks).
+How recombination really works and what it serves to seems to be little understood both in genetic algorithms and in biology.
+In the genetic algorithms literature, there is at least as many negative evidence for the building block hypothesis as support, including a well-known paper by the authors.
+
+Holland's aim, however, was
 more general than a new class of algorithms---his goal was an
@@
 for Holland, just icing on the cake.  
+This broader view of adaptation
+has inspired many, and engendered criticism
+from others, leading to spirited intellectual debates and
+controversies.  Most controversial is the extent to which it can be
+demonstrated, either empirically or mathematically, that the behavior
+of adaptive systems is actually governed by these principles.
+Regardless of one's position on this question, the ideas are compelling
+and provide a set of unifying concepts for thinking about adaptation.
+%as follows:
+%Underlying Holland's theory of adaptation are the following core ideas:
 
@@
 \emph{bucket-brigade} algorithm combined with a genetic algorithm, presaged many of the ideas in modern reinforcement
-learning.
+learning, notably the temporal-difference methods introduced by
+Sutton and Barto~\cite{SuttonAndBarto1990}.
 
@@
 real-world decision and control problems, but classifier systems can
-perhaps be thought of as an early stage-setting method that helped
-build the foundations for later approaches.
+perhaps be thought of as an early stage-setting method that
+foreshadowed these later approaches.
 
@@
 
-Holland's work focused on {\emph exploratory} models---those that explore basic principles and
+Holland's work focused on \emph{exploratory} models---those that explore basic principles and
 mechanisms, even if they do not make specific or detailed
-predictions.  Such models can show generically how certain behaviors
+predictions\cite{Holland1995}.  Such models can show generically how certain behaviors
 could be produced.  Holland pioneered a style of modeling that has
@@
 actual'' \cite{Eddington1927}.
-
-%It should be noted that Holland's view of modeling is by no means
-%typical.  For example, a textbook on computational modeling offers
-% the following definition: ``Modeling is the application
-%of methods to analyze complex, real-world problems in order to make
-%predictions about what might happen with various actions''
-%\cite{Shiflet2014}.   This perspective rules out the
-%kind of exploratory modeling that Holland was most interested in.
+Holland's view of modeling was by no means
+typical.  As just one example of the prevailing view that models are
+for predicting: 
+%For example, a textbook on computational modeling offers
+% the following definition: 
+``Modeling is the application
+of methods to analyze complex, real-world problems in order to make
+predictions about what might happen with various actions''
+\cite{Shiflet2014}.   This perspective rules out the
+kind of exploratory modeling that interested Holland the most.
 
@@
 of the Santa Fe Institute (SFI), the first institution dedicated to
-developing a science of complexity, and the many others that followed.
+developing a science of complexity, as well as the other complexity institutes that followed.
 % [add in how many complexity
@@
 % interacting? 
-nested) spatial and temporal scales, with some behavior being
+nested) spatial and temporal scales, with some behavior 
 conserved across all scales and other behaviors changing at different
@@
 adaptive, with 
-behavioral rules continually adjusting through evolution and learning.  Although this
+behavioral rules continually adjusted through evolution and learning.  Although this
 is far from a formal definition of complex systems, most people
@@
 
-The SFI Artificial Stock Market project \cite{Arthur1997,Palmer1994}
+The SFI Artificial Stock Market project \cite{Arthur1997b,Palmer1994}
 developed an exploratory model in which rational
@@
 
+%Reviewer 3It seems to me that the essay could stress even more that Holland was interested in "open ended systems" and that,  ironically,   as genetic algorithms were turned into static function optimizers these "next generation evolutionary algorithms" lost the adaptive properties that Holland valued and emphasized.
+We still have very little idea about how to build truly open ended systems.
+
 %Blake LeBaron (economics);
@@
 of his career fit into the larger landscape of complex systems
-research.  Holland's focus on understanding the mechanisms by which complex patterns emerge and change,
+research.  In these works he especially emphasized the open-ended
+nature of complex systems, where change and adaptation are
+continual, systems co-evolve with each other, and ecological niches arise
+and decay spontaneously depending on resource availability and competition..
+
+%i thnk you left out a major  theme in john: the role of the concept of niche.
+ % for example, i think that he  was so interested in flows.  because a niche
+% can be thought of as a pattterns of flows.  and ibelfeve -one of his
+% major goal  ofor ECHO was to study   niche formation and structure 
+% did yolu knolw that af\ef the bach gtoujp  \faded we stArted  A group 
+% dedicted to zstudyung nichrd ijn all  thire guiidee, i/with john, nmercdes, 
+% scott, mi\\chael cohen ,evan ecnomo (a rackam fcfelw) , and me as the core,
+%  wiyth visits by;\mike gordon, bob savitn ]marjk newman maria riolo,ed askerville, etc, 
+% as far as i can tell, we ]met on avg about 1/mon from  oct 200 - feb
+% 2012.
+
+Holland's focus on understanding the mechanisms by which complex patterns emerge and change,
 rather than simply characterizing the patterns themselves (e.g.,
@@
 disciplines, articulate grand overarching
-principles, and them back them up with computational mechanisms and
+principles, and then back them up with computational mechanisms and
 mathematics.  Unlike most researchers, Holland moved seamlessly among these
@@
 
+\section{Acknowledgments}
+
+The authors thank L. Booker, R. Riolo, K. Winter, and the anonymous
+reviewers for their careful reading of the manuscript and many helpful
+suggestions.  SF gratefully acknowledges the partial support of NSF
+(NeTS 1518878, CNS 1444500), DARPA (FA8750-15-C-0118), 
+Air Force Office of Scientific Research (FA8750-15-2-0075), and the Santa Fe Institute.
+
 \bibliographystyle{plain}

### 2016-02-09 paper.tex
made axelrod edits, except for adding more refs to markets section

@@
 collaborators found that the model's dynamics replicated several
-features of real-life markets.
+otherwise puzzling features of real-life markets.
 
@@
 willingness to tackle the hardest questions, develop his own
-formalisms, and use mathematics to provide insight sets a high bar for
+formalisms, and use mathematics and simulation to provide insight sets a high bar for
 scientists in all disciplines.
@@
 %His insights aboutwere deeper and more general than what often passes for work in biomimicry, e.g., for robots.
-Holland's personality and humanity were somehow inextricably tangled
-up with his intellectual contributions.  Since his death, many of
+Holland's personality and humanity were somehow inextricably entangled
+with his intellectual contributions.  Since his death, many of
 Holland's former students and colleagues have movingly described their
@@
 
-The authors thank L. Booker, R. Riolo, K. Winter, and the anonymous
+The authors thank R. Axelrod, L. Booker, R. Riolo, K. Winter, and the anonymous
 reviewers for their careful reading of the manuscript and many helpful

### 2016-02-24 paper.tex
melanie's edits through steph

@@
 
-In Aug. 2015, Professor John H. Holland passed in Ann Arbor, MI, where
+In August 2015, Professor John H. Holland passed away in Ann Arbor, MI, where
 he had served on the University of Michigan faculty for over 50 years.
@@
 % SF softened the claim here after reading the comments section on the blog.
-Ph.D. degrees in 1959\cite{London2003}, his contributions are sometimes better known outside of
+Ph.D. degrees in 1959 \cite{London2003}, his contributions are sometimes better known outside of
 computer science than within.
@@
 
-Holland's aim, however, was
-more general than a new class of algorithms---his goal was an
+Holland's aim, however, was more general than a new class of algorithms.  His goal was an
 interdisciplinary theory of adaptation, one that would inform biology, 
@@
 for Holland, just icing on the cake.  
-This broader view of adaptation
+Holland's broader view of adaptation
 has inspired many, and engendered criticism
@@
 demonstrated, either empirically or mathematically, that the behavior
-of adaptive systems is actually governed by these principles.
+of adaptive systems is actually governed by Holland's proposed principles.
 Regardless of one's position on this question, the ideas are compelling
@@
 
-Models can be tacit and learned through evolutionary time, as in the
+Models can be tacit and learned over evolutionary time, as in the
 case of bacteria swimming up a chemical gradient, or explicit and
@@
 environment to enhance survival.  The key learning elements of this method, the
-\emph{bucket-brigade} algorithm combined with a genetic algorithm, presaged many of the ideas in modern reinforcement
+bucket-brigade algorithm, combined with a genetic algorithm, presaged many of the ideas in modern reinforcement
 learning, notably the temporal-difference methods introduced by
@@
 activated, compete to post their results on a shared ``message list,''
-serving as the system's short-term memory (again inspired by Hebb's
-work and AI blackboard systems of the day).  Unlike blackboard
+serving as the system's short-term memory, again inspired by Hebb's
+work as well as by AI blackboard systems of the day.  Unlike blackboard
 systems, however, new rules are generated automatically in a
@@
 cause a system to behave in a certain way, an aspiration that goes
-well beyond data fitting methods, which typically focus only on
+well beyond data-fitting methods, which typically focus only on
 the aggregate behavior of a system.
@@
 reproduction, and heredity are sufficient to account for the richness
-of our natural world.
+of the natural world.
 
@@
 mechanisms, even if they do not make specific or detailed
-predictions\cite{Holland1995}.  Such models can show generically how certain behaviors
+predictions \cite{Holland1995}.  Such models can show generically how certain behaviors
 could be produced.  Holland pioneered a style of modeling that has
@@
 actual'' \cite{Eddington1927}.
-Holland's view of modeling was by no means
-typical.  As just one example of the prevailing view that models are
-for predicting: 
+%Holland's view of modeling was by no means
+%typical.  As just one example of the prevailing view that models are
+%for predicting: 
 %For example, a textbook on computational modeling offers
 % the following definition: 
-``Modeling is the application
-of methods to analyze complex, real-world problems in order to make
-predictions about what might happen with various actions''
-\cite{Shiflet2014}.   This perspective rules out the
-kind of exploratory modeling that interested Holland the most.
+%``Modeling is the application
+%of methods to analyze complex, real-world problems in order to make
+%predictions about what might happen with various actions''
+%\cite{Shiflet2014}.   This perspective rules out the
+%kind of exploratory modeling that interested Holland the most.
 
@@
 
-Holland was interested in a broad array of adaptive systems---immune
+Holland was interested in a broad array of adaptive systems, such as immune
 systems, ecologies, financial markets, cities, and the brain---systems
 that are \emph{complex}.  In the early 1980s, he teamed up with a small
-group of scientists, primarily physicists with a sprinkling of
+group of scientists, primarily physicists, with a sprinkling of
 economists and biologists, to discuss what properties this wide swath of systems
@@
 markers (called tags) and internal preferences, and then used them to
-acquire resources, form higher-level aggregate structures, such as
+acquire resources and form higher-level aggregate structures, such as
 trading relationships, symbiotic groups, trophic cascades, and
@@
 mimicry and used it to deceive competitors.  In many runs, the model
-developed increasingly complex structures and behaviors up to a point.
+developed increasingly complex structures and behaviors. % up to a point.
 When these runs stabilized, the resulting population of agents was
@@
 continual, systems co-evolve with each other, and ecological niches arise
-and decay spontaneously depending on resource availability and competition..
+and decay spontaneously depending on resource availability and competition.
 
@@
 machine learning today, e.g., reinforcement, learning in non-Markovian
-environments and active learning.   His seminal genetic algorithm
+environments, and active learning.   His seminal genetic algorithm
 spawned the field of evolutionary computation, and his insights
@@
 
-The authors thank R. Axelrod, L. Booker, R. Riolo, K. Winter, and the anonymous
-reviewers for their careful reading of the manuscript and many helpful
-suggestions.  SF gratefully acknowledges the partial support of NSF
-(NeTS 1518878, CNS 1444500), DARPA (FA8750-15-C-0118), 
-Air Force Office of Scientific Research (FA8750-15-2-0075), and the Santa Fe Institute.
+The authors thank R. Axelrod, L. Booker, R. Riolo, K. Winter, and the
+anonymous reviewers for their careful reading of the manuscript and
+many helpful suggestions.  SF gratefully acknowledges the partial
+support of NSF (NeTS 1518878, CNS 1444500), DARPA (FA8750-15-C-0118),
+Air Force Office of Scientific Research (FA8750-15-2-0075), and the
+Santa Fe Institute.  MM gratefully acknowledges the partial support of
+NSF (IIS-1423651).  Any opinions, findings, and conclusions or
+recommendations expressed in this material are those of the authors
+and do not necessarily reflect the views of the National Science
+Foundation.
 

### 2016-02-24 paper.tex
steph's corrections on top of mel's

@@
 actual'' \cite{Eddington1927}.
+This view of modeling is not typical.  For many, models
+are useful only to the extent that they generate accurate predictions, a
+perspective that rules out the very kind of exploratory modeling that
+interested Holland the most.
 %Holland's view of modeling was by no means
@@
 interdependent organizations.  ECHO agents sometimes discovered
-mimicry and used it to deceive competitors.  In many runs, the model
-developed increasingly complex structures and behaviors. % up to a point.
-When these runs stabilized, the resulting population of agents was
+mimicry and used it to deceive competitors.  These complex interacting
+patterns arose throughout the model's execution, which started with a single
+minimal agent.
+%In many runs, the model
+%developed increasingly complex structures and behaviors. % up to a point.
+When the runs stabilized, the resulting population of agents was
 found to reproduce several well-known patterns observed in nature,

### 2016-02-25 paper.tex
addressed axelrod comment about mode consistency with env

@@
 these early homomorphic theories of modeling stand as the most elegant
-approach we know of to characterize when a model is consistent with
+approach we know of to characterize how consistent a model is with
 the environment and how an intelligent agent, human or artificial, can
